---
subtitle: "Volatility I: Stylized Facts & Naive Estimation"
---

```{r}
#| echo: false
#| warning: false
#| eval: false
library(tidyverse)
library(scales)
```

## The Central Question

The Markowitz mean-variance investor solves

$$
w_t^* = \arg\max_{w} \; w'\mu_t - \frac{\gamma}{2} w'\Sigma_t w
$$

- $\mu_t = E(r_{t+1} \mid \mathcal{F}_t)$: expected returns — hard to estimate, low signal-to-noise
- $\Sigma_t = \text{Cov}(r_{t+1} \mid \mathcal{F}_t)$: conditional covariance matrix — **this lecture block**

**Why focus on $\Sigma_t$?**

- Estimation error in $\hat\mu$ swamps the signal; many practitioners set $\mu_t = 0$ and focus on the global minimum variance portfolio (GMVP)
- The GMVP weights $w^* \propto \Sigma_t^{-1} \mathbf{1}$ depend *entirely* on $\Sigma_t$
- Even expected-return-based strategies (factor models, ML predictions) require $\Sigma_t$ for risk adjustment and position sizing

**The block roadmap:**

| Lecture | Method | What it estimates |
|---------|--------|-------------------|
| 1 | Rolling SD, EWMA | Marginal $\sigma_{i,t}$ naively |
| 2 | GARCH family | $\sigma_{i,t}$ from daily returns |
| 3 | Realized variance | $\sigma_{i,t}$ from intraday data |
| 4 | HAR models | Forecasts of $\sigma_{i,t+h}$ |
| 5 | DCC-GARCH, shrinkage | Full $\Sigma_t$, $N = 50$ |
| 6 | Realized covariance | Full $\Sigma_t$ from intraday data |

## Running Example

Throughout all six lectures we work with the **same universe of 50 NASDAQ-listed stocks**

- Chosen to align with available data infrastructure
- Large enough to make dimensionality a real challenge ($N = 50 \Rightarrow 1{,}275$ unique covariance elements)
- Small enough to remain computationally manageable

**Data sources by lecture:**

| Data | Source | Lectures |
|------|--------|----------|
| Daily close prices | CRSP via WRDS | 1–2, 5 |
| Tick-by-tick trades & quotes | LOBSTER (NASDAQ) | 3–4, 6 |
| Pre-computed realized measures | VOLARE database | 4, 6 |

**Why NASDAQ?**

- LOBSTER covers NASDAQ natively (direct NASDAQ feed)
- VOLARE provides ready-made realized measures for NASDAQ stocks
- Sector diversity: tech, healthcare, consumer discretionary — interesting cross-sectional heterogeneity

## The Problem: Volatility is Latent

- At time $t$ we observe the return $r_t$ but **not** the variance $\sigma_t^2$ that generated it
- We observe a single draw from $r_t \sim \mathcal{D}(0, \sigma_t^2)$ — the variance is a parameter, not a random variable we can see

**Classical sample variance:** with $T$ observations

$$\hat\sigma^2 = \frac{1}{T-1}\sum_{t=1}^T (r_t - \bar r)^2$$

Problems:

- Assumes $\sigma_t^2 = \sigma^2$ — constant over time
- Returns are not i.i.d.: tomorrow's $\sigma_t^2$ depends on history
- For portfolio rebalancing we need a **forecast**: $E(\sigma_{t+1}^2 \mid \mathcal{F}_t)$, not a historical average

**The key challenge:** find an estimator that

1. Responds quickly to changing market conditions
2. Does not overfit to noise
3. Scales to $N = 50$ assets and produces a coherent $\hat\Sigma_t$

# Stylized Facts

## Fact 1: Fat Tails

Daily returns exhibit **excess kurtosis** — extreme moves are far more likely than Gaussian

$$\text{kurt}(r_t) = \frac{E[(r_t - \mu)^4]}{\sigma^4} \gg 3$$

- S&P 500 kurtosis historically around 5–10 (Gaussian = 3)
- Tails heavier for individual stocks than for indices (diversification reduces kurtosis)
- **Portfolio implication:** Gaussian VaR systematically underestimates tail risk; using $\Sigma$ from a Gaussian model misprices low-probability, high-impact events

:::: {.columns}
::: {.column width="50%"}
**Implication for estimation:**

- OLS on squared returns is not efficient under fat tails
- QML (quasi-maximum likelihood) estimation of GARCH remains consistent even with non-Gaussian errors (Lecture 2)
- For risk management: use Student-$t$ or skewed-$t$ distributions
:::
::: {.column width="50%"}
**Implication for portfolio choice:**

- Mean-variance is exact only under Gaussian returns or quadratic utility
- With fat tails: higher moments matter; mean-variance remains a *useful approximation*
- Practitioners add tail-risk overlays (options, stop-losses) on top of MV portfolios
:::
::::

## All Four Facts in One Stock

```{r}
#| echo: false
#| warning: false
#| cache: true
library(tidyverse)
library(tidyfinance)

prices_sf <- download_data(
  type = "stock_prices", symbols = "AAPL",
  start_date = "2015-01-01", end_date = "2024-12-31"
) |>
  arrange(date) |>
  mutate(ret = log(adjusted_close / lag(adjusted_close))) |>
  drop_na()

ci_sf <- 1.96 / sqrt(nrow(prices_sf))

acf_sq <- with(acf(prices_sf$ret^2, lag.max = 60, plot = FALSE),
               list(lag = as.numeric(lag[-1]), acf = as.numeric(acf[-1])))

cc_obj <- ccf(prices_sf$ret, prices_sf$ret^2, lag.max = 10, plot = FALSE)
cc_lev <- list(lag = as.numeric(cc_obj$lag[as.numeric(cc_obj$lag) >= 0]),
               ccf = as.numeric(cc_obj$acf[as.numeric(cc_obj$lag) >= 0]))

ojs_define(
  sf_returns = prices_sf |> mutate(date = as.character(date)) |>
                 select(date, ret) |> as.list(),
  sf_acf     = map2(acf_sq$lag, acf_sq$acf, ~ list(lag = .x, acf = .y)),
  sf_ccf     = map2(cc_lev$lag, cc_lev$ccf, ~ list(lag = .x, ccf = .y)),
  sf_ci      = ci_sf
)
```

```{ojs}
//| echo: false
ret_data = sf_returns.date.map((d, i) => ({ date: new Date(d), ret: sf_returns.ret[i] * 100 }))

p_ret = Plot.plot({
  width: 340, height: 200, marginLeft: 42,
  marks: [Plot.lineY(ret_data, {x: "date", y: "ret", stroke: "#3d3d3d", strokeWidth: 0.5})],
  x: {label: null}, y: {label: "Return (%)"},
  title: "(a) Daily log returns — AAPL 2015–2024"
})

p_qq = Plot.plot({
  width: 340, height: 200, marginLeft: 42,
  marks: [
    Plot.qq(ret_data, {x: "ret", stroke: "#901A1E", strokeOpacity: 0.4, strokeWidth: 1.5}),
  ],
  title: "(b) QQ-plot vs Gaussian (tails outside line = fat tails)"
})

p_acf = Plot.plot({
  width: 340, height: 200, marginLeft: 42,
  marks: [
    Plot.barY(sf_acf, {x: "lag", y: "acf", fill: "#901A1E", fillOpacity: 0.75}),
    Plot.ruleY([sf_ci, -sf_ci], {stroke: "#666666", strokeDasharray: "3 2"})
  ],
  x: {label: "Lag (days)"}, y: {label: "ACF"},
  title: "(c) ACF of r²ₜ — slow decay = clustering + long memory"
})

p_lev = Plot.plot({
  width: 340, height: 200, marginLeft: 42,
  marks: [
    Plot.barY(sf_ccf, {x: "lag", y: "ccf", fill: "#901A1E", fillOpacity: 0.75}),
    Plot.ruleY([0, sf_ci, -sf_ci], {stroke: ["#3d3d3d","#666666","#666666"],
               strokeDasharray: ["0","3 2","3 2"]})
  ],
  x: {label: "Lag k"}, y: {label: "CCF"},
  title: "(d) Corr(rₜ, r²ₜ₊ₖ) — negative at k>0 = leverage effect"
})

html`<div style="display:grid;grid-template-columns:1fr 1fr;gap:6px">${p_ret}${p_qq}${p_acf}${p_lev}</div>`
```

## Fact 2: Volatility Clustering

**Mandelbrot (1963):** *"large changes tend to be followed by large changes, of either sign, and small changes tend to be followed by small changes"*

$$\text{Corr}(r_t^2, r_{t-k}^2) > 0 \quad \text{for } k = 1, 2, \ldots$$

The autocorrelation function (ACF) of squared returns is significantly positive at many lags

**What this means for portfolio management:**

- Volatility is **predictable** from its own history
- High-volatility regimes persist → position sizing should respond dynamically
- A model that uses a constant $\sigma^2$ will be consistently wrong in both directions

**Key diagnostic:** ACF of $r_t^2$ (or $|r_t|$) should show slow decay; ACF of $r_t$ itself should be near zero

## Fact 3: Mean Reversion

Volatility is stationary — it clusters but eventually **reverts to its long-run mean** $\bar\sigma^2$

- This is distinct from clustering: clustering = short-run persistence; mean reversion = long-run stationarity
- Empirically: half-life of a volatility shock is roughly 10–30 trading days for most stocks
- VIX: mean-reverts around $\approx 20\%$ annualised with well-documented spikes

**For multi-step forecasting:**

$$E(\sigma_{t+h}^2 \mid \mathcal{F}_t) \to \bar\sigma^2 \quad \text{as } h \to \infty$$

Any reasonable volatility model must respect this — an integrated volatility model (IGARCH) does not mean-revert, which is problematic for long-horizon forecasts

## Fact 4: The Leverage Effect

**Black (1976):** negative returns increase volatility more than positive returns of the same magnitude

$$\text{Corr}(r_t, \sigma_{t+1}^2) < 0$$

Two competing explanations:

:::: {.columns}
::: {.column width="50%"}
**Financial leverage channel:**

A price drop increases the debt-to-equity ratio → the firm becomes riskier → volatility rises
:::
::: {.column width="50%"}
**Volatility feedback channel:**

Higher expected volatility raises the required return → price falls to deliver that return
:::
::::

**Portfolio implication:**

- Drawdowns coincide with rising volatility → losses and rising risk arrive together
- Dynamic strategies must increase defensive positioning *quickly* after negative shocks
- Symmetric volatility models (standard GARCH) miss this: we need GJR-GARCH or EGARCH (Lecture 2)

**Diagnostic:** cross-correlation between $r_t$ and $r_{t+k}^2$ should be negative for small $k > 0$

## Fact 5: Long Memory

The ACF of $r_t^2$ decays **hyperbolically**, not exponentially

$$\text{Corr}(r_t^2, r_{t-k}^2) \approx c \cdot k^{2d-1}, \quad 0 < d < 0.5$$

- Exponential decay: $\rho_k \propto \phi^k$ — ARMA and standard GARCH
- Hyperbolic decay: $\rho_k \propto k^{-\alpha}$ — long memory, still positive at lag 100+

**Consequences:**

- Standard GARCH(1,1) is near-integrated ($\alpha + \beta \approx 1$) to approximate long memory — but this implies no mean reversion
- FIGARCH and EGARCH with long-memory specifications capture this directly
- HAR models (Lecture 4) mimic long memory through a sum of AR components at different horizons — simple and effective

## Fact 6: Co-movement & Common Factors

Volatilities of different assets **move together**, especially during crises

- Empirically: a single common factor explains 40–60% of variance in individual stock volatilities
- The VIX ("fear gauge") proxies for this common factor at market level
- Sector-specific volatility factors exist on top of the market factor

**For portfolio choice:**

$$\Sigma_t = D_t C_t D_t$$

where $D_t = \text{diag}(\sigma_{1,t}, \ldots, \sigma_{N,t})$ and $C_t$ is the correlation matrix

- Co-movement means $C_t$ is also time-varying — not just the marginal variances
- A model that captures only $D_t$ (e.g., diagonal GARCH) misses the time-variation in $C_t$
- DCC-GARCH (Lecture 5) captures both components

# Naive Estimators

## Rolling Window Standard Deviation

The simplest time-varying volatility estimator:

$$\hat\sigma_{t}^2 = \frac{1}{M-1} \sum_{s=t-M+1}^{t} (r_s - \bar r_{t,M})^2$$

where $M$ is the window length (e.g., $M = 21$ for one month, $M = 63$ for one quarter)

**The bias-variance tradeoff:**

:::: {.columns}
::: {.column width="50%"}
**Short window ($M$ small):**

- Responds quickly to volatility changes
- High variance: noisy estimates
- Weights past observation equally within the window, then drops to zero abruptly
:::
::: {.column width="50%"}
**Long window ($M$ large):**

- Smooth, stable estimates
- Slow to react to regime changes
- Incorporates stale data — poor in fast-moving markets
:::
::::

**The cliff-edge problem:** when a large return falls out of the window, $\hat\sigma_t^2$ can jump even though no new information has arrived today

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
library(tidyverse)

# Assume crsp_returns: a tibble with columns date, symbol, ret
# for our 50 NASDAQ stocks from CRSP

rolling_vol <- crsp_returns |>
  group_by(symbol) |>
  arrange(date) |>
  mutate(
    vol_21  = slider::slide_dbl(ret, sd, .before = 20, .complete = TRUE),
    vol_63  = slider::slide_dbl(ret, sd, .before = 62, .complete = TRUE),
    vol_126 = slider::slide_dbl(ret, sd, .before = 125, .complete = TRUE)
  ) |>
  ungroup()
```

## Python
```{python}
#| eval: false
import pandas as pd

# crsp_returns: DataFrame with columns date, symbol, ret
rolling_vol = (crsp_returns
  .sort_values(["symbol", "date"])
  .groupby("symbol")["ret"]
  .apply(lambda x: x.rolling(window=21, min_periods=21).std())
  .reset_index(name="vol_21")
)
```
:::

## Exponentially Weighted Moving Average (EWMA)

The **RiskMetrics** estimator (J.P. Morgan 1994) uses exponentially declining weights:

$$\hat\sigma_t^2 = (1 - \lambda) r_{t-1}^2 + \lambda \hat\sigma_{t-1}^2$$

where $\lambda = 0.94$ for daily data (the RiskMetrics default)

**Recursive form:** expanding the recursion gives

$$\hat\sigma_t^2 = (1-\lambda) \sum_{k=1}^{\infty} \lambda^{k-1} r_{t-k}^2$$

- Exponentially declining weights: recent observations count more
- No cliff-edge: all past observations included (with negligible weight at long lags)
- Implicit mean reversion to zero: $E(\hat\sigma_\infty^2) = 0$ — problematic for long-horizon forecasts

**Parameter $\lambda$:**

- $\lambda$ close to 1: slow adjustment, sluggish response
- $\lambda$ close to 0: fast adjustment, noisy estimates
- RiskMetrics $\lambda = 0.94$: half-life of $\log(0.5)/\log(0.94) \approx 11$ trading days

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
ewma_vol <- function(ret, lambda = 0.94) {
  n <- length(ret)
  sigma2 <- numeric(n)
  sigma2[1] <- ret[1]^2
  for (t in 2:n) {
    sigma2[t] <- (1 - lambda) * ret[t-1]^2 + lambda * sigma2[t-1]
  }
  sqrt(sigma2)
}

ewma_estimates <- crsp_returns |>
  group_by(symbol) |>
  arrange(date) |>
  mutate(ewma_94 = ewma_vol(ret, lambda = 0.94),
         ewma_97 = ewma_vol(ret, lambda = 0.97)) |>
  ungroup()
```

## Python
```{python}
#| eval: false
def ewma_vol(ret, lam=0.94):
    sigma2 = [ret.iloc[0]**2]
    for r in ret.iloc[1:]:
        sigma2.append((1 - lam) * r**2 + lam * sigma2[-1])
    return pd.Series(sigma2, index=ret.index) ** 0.5

ewma_estimates = (crsp_returns
  .sort_values(["symbol", "date"])
  .groupby("symbol")
  .apply(lambda df: df.assign(ewma_94=ewma_vol(df["ret"], 0.94)))
  .reset_index(drop=True)
)
```
:::

## Interactive: Rolling SD vs EWMA

*Drag the sliders — watch how each estimator reacts to the same return path*

```{ojs}
//| echo: false
viewof M_roll = Inputs.range([5, 126], {step: 1, value: 21,
  label: "Rolling window  M (days)"})
viewof lam_ewma = Inputs.range([0.85, 0.99], {step: 0.01, value: 0.94,
  label: "EWMA decay  λ"})
```

```{ojs}
//| echo: false
// Simulate a GJR-GARCH(1,1) return series (fixed seed via deterministic init)
garch_path = {
  const n = 750, omega = 4e-5, a = 0.06, g = 0.10, b = 0.89
  let s2 = omega / (1 - a - g/2 - b), r = 0
  const out = []
  // Use a simple LCG for reproducibility
  let seed = 42
  const lcg = () => { seed = (seed * 1664525 + 1013904223) & 0xffffffff
    return (seed >>> 0) / 4294967296 }
  const randn = () => Math.sqrt(-2*Math.log(lcg()+1e-10)) * Math.cos(2*Math.PI*lcg())
  for (let t = 0; t < n; t++) {
    const z = randn()
    r = Math.sqrt(s2) * z
    out.push({ t, r, true_vol: Math.sqrt(s2 * 252) * 100 })
    s2 = omega + (a + (r < 0 ? g : 0)) * r*r + b * s2
  }
  return out
}
```

```{ojs}
//| echo: false
vol_paths = {
  const rets = garch_path.map(d => d.r)
  const n = rets.length
  const out = garch_path.map(d => ({...d, roll: null, ewma: null}))
  // Rolling SD
  for (let t = M_roll - 1; t < n; t++) {
    const w = rets.slice(t - M_roll + 1, t + 1)
    const mu = w.reduce((a,x) => a+x, 0) / M_roll
    const v  = w.reduce((a,x) => a+(x-mu)**2, 0) / (M_roll-1)
    out[t].roll = Math.sqrt(v * 252) * 100
  }
  // EWMA
  let ev = rets[0]**2
  for (let t = 0; t < n; t++) {
    if (t > 0) ev = (1 - lam_ewma) * rets[t-1]**2 + lam_ewma * ev
    out[t].ewma = Math.sqrt(ev * 252) * 100
  }
  return out
}
```

```{ojs}
//| echo: false
Plot.plot({
  width: 700, height: 270, marginLeft: 50,
  marks: [
    Plot.lineY(vol_paths, {x: "t", y: "true_vol",
      stroke: "#cccccc", strokeWidth: 1.2}),
    Plot.lineY(vol_paths.filter(d => d.roll !== null), {x: "t", y: "roll",
      stroke: "#901A1E", strokeWidth: 2}),
    Plot.lineY(vol_paths, {x: "t", y: "ewma",
      stroke: "#3d3d3d", strokeWidth: 2, strokeDasharray: "5 3"}),
  ],
  y: {label: "Annualised vol (%)", grid: true},
  x: {label: "Trading day"},
  caption: html`<span style='color:#cccccc'>——</span> True (GJR-GARCH) &ensp;
    <span style='color:#901A1E'>——</span> Rolling SD (M days) &ensp;
    <span style='color:#3d3d3d'>– –</span> EWMA (λ)`
})
```

## Naive Estimators in Action — NASDAQ Example

**Stylized facts diagnostics** for selected NASDAQ stocks (pre-computed from CRSP daily returns):

- **QQ-plots:** tails extend far beyond the Gaussian reference line — confirming excess kurtosis
- **ACF of $r_t^2$:** significant positive autocorrelation at lags 1–50, slow decay — confirming long memory
- **Leverage cross-correlations:** $\text{Corr}(r_t, r_{t+k}^2)$ negative for $k = 1, \ldots, 5$ — confirming leverage effect

**Comparing rolling SD and EWMA:**

- Both capture clustering qualitatively; EWMA is smoother and reacts faster
- During COVID-19 crash (Feb–Mar 2020): EWMA rises quickly, rolling SD has a delayed and abrupt jump
- After recovery: rolling SD remains elevated due to the cliff-edge; EWMA declines steadily

**Cross-sectional heterogeneity:**

- Tech stocks (AAPL, MSFT, NVDA): higher unconditional volatility, stronger clustering
- Healthcare (JNJ, MRK): lower and more stable volatility
- Consumer discretionary (AMZN, TSLA): high volatility, strong leverage effect

## From Naive Estimators to Portfolio Weights

To see the portfolio implications, plug $\hat\sigma_{i,t}$ into the GMVP:

$$\hat\Sigma_t = \hat D_t \hat C \hat D_t$$

where $\hat C$ is the **sample correlation matrix** (assumed constant) and $\hat D_t = \text{diag}(\hat\sigma_{1,t}, \ldots, \hat\sigma_{N,t})$ from rolling SD or EWMA

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
library(Matrix)

# Compute GMVP weights from EWMA diagonal + sample correlations
compute_gmvp <- function(sigma_vec, corr_mat) {
  D <- diag(sigma_vec)
  Sigma <- D %*% corr_mat %*% D
  Sigma_inv <- solve(Sigma)
  ones <- rep(1, nrow(Sigma))
  w <- Sigma_inv %*% ones
  w / sum(w)
}

# For each date t, compute weights
sample_corr <- cor(returns_wide)  # N x N sample correlation matrix
gmvp_weights <- rolling_vol |>
  group_by(date) |>
  summarise(
    weights = list(compute_gmvp(vol_21, sample_corr))
  )
```

## Python
```{python}
#| eval: false
import numpy as np

def compute_gmvp(sigma_vec, corr_mat):
    D = np.diag(sigma_vec)
    Sigma = D @ corr_mat @ D
    Sigma_inv = np.linalg.inv(Sigma)
    ones = np.ones(len(sigma_vec))
    w = Sigma_inv @ ones
    return w / w.sum()

sample_corr = returns_wide.corr().values
```
:::

**Limitation of naive approach:** $\hat C$ is assumed constant — we will relax this in Lectures 5–6

## Forecasting Comparison: What Can Naive Estimators Do?

Evaluate the one-step-ahead forecast $\hat\sigma_t^2$ against the proxy $r_t^2$ using the **QLIKE loss function**:

$$\text{QLIKE} = \frac{r_t^2}{\hat\sigma_t^2} - \log\frac{r_t^2}{\hat\sigma_t^2} - 1$$

- QLIKE is robust to the noise in $r_t^2$ as a proxy for the true (latent) $\sigma_t^2$
- Under regularity conditions, QLIKE rankings are invariant to the choice of volatility proxy

:::: {.columns}
::: {.column width="60%"}
**Preview of the full-block horse race:**

| Estimator | Average QLIKE | Notes |
|-----------|--------------|-------|
| Rolling SD (M=21) | baseline | Cliff-edge, slow |
| EWMA (λ=0.94) | ↓ | Faster response |
| GARCH(1,1) | ↓↓ | Lecture 2 |
| HAR-RV | ↓↓↓ | Lecture 4 |
| DCC-GARCH | — | Lecture 5, multivariate |
| Realized Cov | — | Lecture 6, multivariate |
:::
::: {.column width="40%"}
**Takeaway:**

Naive estimators are simple and interpretable but leave significant predictive accuracy on the table

Each subsequent lecture shows how to recover that accuracy

The final judge: **out-of-sample GMVP performance**
:::
::::

# Exercises

## Exercise Set 1

**Data preparation**

1. Load the daily CRSP returns for the 50 NASDAQ stocks. Compute log returns and verify that the data covers at least 10 years of history (≥2,500 trading days per stock).

**Stylized facts**

2. For a subset of 10 stocks (2 tech, 2 healthcare, 2 consumer discretionary, 4 of your choice):
   - Plot QQ-plots against the Gaussian and Student-$t(5)$ distributions. Which fits better?
   - Plot the ACF of $r_t^2$ up to lag 100. Do you observe slow hyperbolic decay?
   - Compute and plot the cross-correlations $\text{Corr}(r_t, r_{t+k}^2)$ for $k = -10, \ldots, 10$. Is the leverage effect visible?

**Naive estimators**

3. Compute rolling standard deviations at $M \in \{21, 63, 126\}$ trading days for all 50 stocks.
   - Visualise the three series for two stocks of your choice. Discuss the bias-variance tradeoff.
   - Identify the cliff-edge problem: find a date where the rolling SD jumps despite no unusual return today.

4. Compute EWMA volatility with $\lambda \in \{0.90, 0.94, 0.97\}$.
   - Compare the half-life implied by each $\lambda$.
   - How does the choice of $\lambda$ affect portfolio turnover? (Hint: compute $\hat w_t^{GMVP}$ for each $\lambda$ and count daily rebalancing trades.)

**Portfolio evaluation**

5. Construct the GMVP using each estimator (rolling 21-day, rolling 63-day, EWMA 0.94) combined with the sample correlation matrix.
   - Compute the annualised out-of-sample portfolio volatility for each strategy.
   - Which estimator gives the lowest realised volatility? Does a shorter or longer window win?

**Key references:** Cont (2001); Pagan & Schwert (1990)
