---
subtitle: "Volatility VI: Realized Covariance Matrices"
---

```{r}
#| echo: false
#| warning: false
#| eval: false
library(tidyverse)
library(highfrequency)
library(nlshrink)
library(Matrix)
```

## Closing the Loop

**The journey so far:**

| Lecture | Estimator | What it produces |
|---------|-----------|-----------------|
| 1 | Rolling SD, EWMA | $\hat\sigma_{i,t}$, constant $\hat C$ |
| 2 | GARCH, GJR-GARCH | $\hat\sigma_{i,t|t-1}$ from daily $r_t$ |
| 3 | Realized variance | $\widehat{RV}_{i,t}$, precise estimate of $\sigma_{i,t}^2$ |
| 4 | HAR models | $\hat\sigma_{i,t+h|t}^2$, forecasts from realized measures |
| 5 | DCC-GARCH, LW shrinkage | Full $\hat\Sigma_t$ at $N = 50$ |
| **6** | **Realized covariance** | **Full $\hat\Sigma_t$ from intraday cross-products** |

**The natural extension:** if realized variance gives us a precise $\hat\sigma_{i,t}^2$ from intraday data, can we similarly estimate the **full covariance matrix** from intraday cross-products?

**Yes** — but with additional challenges:

- Non-synchronous trading: stocks don't trade at the same instants
- Noise propagates into off-diagonal elements too
- Positive definiteness is not guaranteed

**Portfolio payoff:** realized covariance provides the most informationally rich estimate of $\Sigma_t$, potentially dominating DCC in portfolio performance

# Realized Covariance

## Definition

The **realized covariance** matrix between $N$ assets on day $t$:

$$\widehat{RC}_t = \sum_{j=1}^n r_{t,j} r_{t,j}'$$

where $r_{t,j}$ is the $N$-dimensional vector of synchronised intraday returns over interval $j$

**In population:** under the continuous semimartingale model and as $n \to \infty$:

$$\widehat{RC}_t \xrightarrow{p} \int_0^1 \sigma_t(s)\sigma_t(s)'\, ds =: IC_t$$

where $IC_t$ is the **integrated covariance** matrix — the true daily covariance

**Element-wise:**

- Diagonal: $[\widehat{RC}_t]_{ii} = \widehat{RV}_{i,t}$ — same as Lecture 3
- Off-diagonal: $[\widehat{RC}_t]_{ij} = \widehat{RCov}_{ij,t}$ — sum of intraday cross-products

## The Epps Effect

**Epps (1979):** as the sampling frequency increases, estimated correlations between assets **decrease toward zero**

This is *not* a feature — it is a bias caused by **non-synchronous trading**:

- Assets do not trade at the same instants
- When we sample at e.g. 1-second intervals, many slots have a return for only one asset (the other is stale)
- Stale returns impute last observed price → their product is zero → correlations are biased down

:::: {.columns}
::: {.column width="55%"}
**Visualising the Epps effect:**

- Compute $\widehat{RCov}_{ij,t}(\Delta)$ at multiple sampling intervals $\Delta$
- Plot pairwise correlation as a function of $\Delta$
- At $\Delta = 1$s: correlation near zero; at $\Delta = 5$min: approaches "true" value
- Effect is stronger for less liquid stocks (more gaps between trades)
:::
::: {.column width="45%"}
```{r}
#| eval: false
# Epps plot for AAPL-MSFT pair
freqs <- c(1, 5, 30, 60, 300, 600, 1800)
corrs <- sapply(freqs, function(f) {
  r_sync <- synchroniseTrades(
    trades_aapl, trades_msft,
    on = "seconds", k = f
  )
  cor(r_sync)[1, 2]
})
tibble(seconds = freqs, correlation = corrs) |>
  ggplot(aes(x = seconds, y = correlation)) +
  geom_line(color = "#901A1E") + geom_point() +
  scale_x_log10() +
  labs(title = "Epps effect: AAPL–MSFT",
       x = "Sampling interval (seconds)",
       y = "Realized correlation")
```
:::
::::

# Synchronisation

## Refresh Time Sampling

Barndorff-Nielsen, Hansen, Lunde & Shephard (2011): **refresh time** sampling is the standard solution to the Epps effect

**Algorithm:**

1. Start at time $s_0 = \max(\text{first trade time across all } N \text{ assets})$
2. Next refresh time $s_{k+1} = \max(\text{next trade time for each asset after } s_k)$
3. At each $s_k$: all $N$ assets have a new observation — compute cross-product

**Properties:**

- By construction, all assets have traded between consecutive refresh times
- Effectively samples at the rate of the *slowest* asset — avoids stale price contamination
- Loses information: fewer observations than if each asset sampled at its own frequency
- Remains consistent for the integrated covariance even under noise

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
library(highfrequency)

# Synchronise trades for a subset of stocks using refresh time
trades_list <- list(
  AAPL = trades_aapl,
  MSFT = trades_msft,
  NVDA = trades_nvda,
  AMZN = trades_amzn,
  TSLA = trades_tsla
)

# Aggregate to refresh times
sync_data <- refreshTime(trades_list)

# Compute realized covariance
rc <- rCov(sync_data, makeReturns = TRUE)
print(rc)
```

## Python
```{python}
#| eval: false
import pandas as pd
import numpy as np

def refresh_time_sync(price_dict):
    """
    Synchronise multiple price series using refresh time algorithm.
    price_dict: {symbol: pd.Series indexed by timestamp}
    """
    # Merge all series, keeping only refresh times
    df = pd.DataFrame(price_dict)
    # Forward fill to propagate last known price
    df_ffill = df.ffill()
    # Refresh time: rows where all assets have a new observation
    new_obs = df.notna()
    refresh_mask = new_obs.all(axis=1)
    return df_ffill[refresh_mask]

sync_prices = refresh_time_sync({
    sym: trades[sym].set_index("time")["price"]
    for sym in selected_symbols
})
sync_returns = np.log(sync_prices).diff().dropna()
RC = sync_returns.T @ sync_returns  # realized covariance matrix
```
:::

## Comparing Synchronisation Methods

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# Naive synchronisation: sample at fixed 5-minute grid
rc_naive <- rCov(
  returns_5min,   # already sampled at 5-min grid
  makeReturns = FALSE
)

# Refresh time synchronisation
rc_refresh <- rCov(sync_data, makeReturns = TRUE)

# How different are the off-diagonal elements?
offdiag_naive   <- rc_naive[upper.tri(rc_naive)]
offdiag_refresh <- rc_refresh[upper.tri(rc_refresh)]

tibble(naive = offdiag_naive, refresh = offdiag_refresh) |>
  ggplot(aes(x = naive, y = refresh)) +
  geom_point(alpha = 0.4, color = "#901A1E") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(title = "Naive vs refresh-time covariances",
       subtitle = "Off-diagonal elements; dashed = 45° line") +
  theme_minimal()
```

## Python
```{python}
#| eval: false
# Naive: simple 5-min grid
rc_naive = (returns_5min.T @ returns_5min).values

# Refresh time
rc_refresh = (sync_returns.T @ sync_returns).values

# Scatter plot of off-diagonal elements
mask = np.triu(np.ones_like(rc_naive, dtype=bool), k=1)
plt.scatter(rc_naive[mask], rc_refresh[mask], alpha=0.4, c="#901A1E")
plt.axline((0, 0), slope=1, linestyle="--", color="gray")
plt.xlabel("Naive covariance"), plt.ylabel("Refresh-time covariance")
```
:::

# Multivariate Realized Kernels

## Extending Realized Kernels to Multiple Assets

Barndorff-Nielsen, Hansen, Lunde & Shephard (2011): the **multivariate realized kernel**

$$\widehat{RK}_t^{ij} = \sum_{h=-H}^{H} k\!\left(\frac{h}{H+1}\right) \Gamma_h^{ij}$$

where $\Gamma_h^{ij} = \sum_{s} r_{t,s}^i r_{t,s-h}^j$ (lead-lag cross-covariance at lag $h$)

**Properties:**

- Consistent estimator of integrated covariance under microstructure noise
- Positive semi-definite by construction with appropriate kernel (flat-top or Parzen)
- Bandwidth $H$ controls the noise-bias tradeoff for each pair
- Computationally demanding for large $N$: $O(N^2 H n)$ — scales quadratically in $N$

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# Multivariate realized kernel (subset of ~10 stocks)
rk_matrix <- rKernelCov(
  rData      = sync_returns_10,   # T_refresh x 10 matrix of returns
  kernelType = "Parzen",
  bw         = "optimal"
)
# Returns a 10 x 10 realized kernel covariance matrix
```

## Python
```{python}
#| eval: false
# Multivariate realized kernel (flat-top)
def multivar_realized_kernel(returns_matrix, H):
    n, N = returns_matrix.shape
    # Compute cross-covariances at each lag h
    RK = np.zeros((N, N))
    for h in range(-H, H+1):
        # Flat-top kernel weight
        weight = 1.0 if abs(h) <= H/2 else 2 * (1 - abs(h)/H)
        if h >= 0:
            gamma_h = returns_matrix[h:].T @ returns_matrix[:n-h]
        else:
            gamma_h = returns_matrix[:n+h].T @ returns_matrix[-h:]
        RK += weight * gamma_h
    return (RK + RK.T) / 2  # symmetrize
```
:::

# Ensuring Positive Definiteness

## The Problem

Raw realized covariance matrices can fail to be **positive definite** (PD) in practice:

- Non-synchronous trading introduces asymmetries
- Small eigenvalues pushed below zero by noise
- For large $N$: probability of a negative eigenvalue grows

**Why PD matters for portfolio optimization:**

- $\hat\Sigma_t$ not PD → $\hat\Sigma_t^{-1}$ does not exist → GMVP weights undefined
- Even near-PD matrices (tiny negative eigenvalues) produce extreme short positions

## Solutions

### Nearest PD Projection (Higham 2002)

Find the PD matrix $\Sigma^*$ closest (in Frobenius norm) to $\hat\Sigma$:

$$\Sigma^* = \arg\min_{\Sigma \succ 0} \|\hat\Sigma - \Sigma\|_F^2$$

```{r}
#| eval: false
library(Matrix)

# Nearest PD projection
RC_pd <- nearPD(RC_raw, corr = FALSE)$mat
```

### Eigenvalue Clipping

Replace all negative eigenvalues with a small positive floor:

```{r}
#| eval: false
make_pd <- function(M, floor = 1e-8) {
  eig <- eigen(M, symmetric = TRUE)
  eig$values <- pmax(eig$values, floor)
  eig$vectors %*% diag(eig$values) %*% t(eig$vectors)
}
RC_clipped <- make_pd(RC_raw)
```

### Shrinkage (preferred)

Apply Ledoit-Wolf shrinkage to the realized covariance matrix:

```{r}
#| eval: false
# Nonlinear shrinkage of realized covariance (also regularises it)
RC_shrunk <- nlshrink_cov(rc_daily_array, k = 1)
# rc_daily_array: T x N x N array, apply per-day or use average
```

**Recommendation:** nonlinear shrinkage is preferred — it simultaneously regularises (controls estimation error) and guarantees PD

# Forecasting Realized Covariance

## Element-wise HAR

**Simplest approach:** apply HAR independently to each of the $N(N+1)/2$ unique elements:

$$[\hat\Sigma_{t+1}]_{ij} = \hat\beta_0^{ij} + \hat\beta_d^{ij} [\widehat{RC}_t]_{ij} + \hat\beta_w^{ij} [\overline{RC}_t^{(w)}]_{ij} + \hat\beta_m^{ij} [\overline{RC}_t^{(m)}]_{ij}$$

where $\overline{RC}_t^{(w)}$ and $\overline{RC}_t^{(m)}$ are weekly and monthly averages of the realized covariance

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# Element-wise HAR for realized covariance (N=10 subset)
N <- 10
har_cov_fits <- list()

for (i in 1:N) {
  for (j in i:N) {
    cov_series <- rc_array[, i, j]  # T-vector of realized cov_{ij}

    df_ij <- tibble(
      y   = lead(cov_series),
      x_d = cov_series,
      x_w = slider::slide_dbl(cov_series, mean, .before = 4, .complete = TRUE),
      x_m = slider::slide_dbl(cov_series, mean, .before = 21, .complete = TRUE)
    ) |> drop_na()

    har_cov_fits[[paste(i, j, sep = "_")]] <- lm(y ~ x_d + x_w + x_m, data = df_ij)
  }
}
```

## Python
```{python}
#| eval: false
har_fits_cov = {}
for i in range(N):
    for j in range(i, N):
        cov_series = rc_array[:, i, j]
        df_ij = pd.DataFrame({
            "y":   pd.Series(cov_series).shift(-1),
            "x_d": cov_series,
            "x_w": pd.Series(cov_series).rolling(5).mean(),
            "x_m": pd.Series(cov_series).rolling(22).mean()
        }).dropna()
        key = f"{i}_{j}"
        har_fits_cov[key] = smf.ols("y ~ x_d + x_w + x_m",
                                     data=df_ij).fit()
```
:::

## DRD Decomposition

**Chiriac & Voev (2011):** decompose $\hat\Sigma_t = D_t R_t D_t$ and model components separately

- $D_t$: diagonal matrix of realized standard deviations — use univariate HAR on $\sqrt{RV_{i,t}}$
- $R_t$: realized correlation matrix — ensure PD, then apply HAR element-wise

**Why DRD helps:**

- Correlations are bounded in $[-1, 1]$ — natural scale for HAR
- Variance and correlation dynamics may be different — separate models can fit better
- Easier to impose PD constraints on correlations than on covariances directly

```{r}
#| eval: false
# DRD decomposition
extract_drd <- function(RC) {
  D_sqrt <- sqrt(diag(RC))
  D_mat  <- diag(D_sqrt)
  D_inv  <- diag(1 / D_sqrt)
  R      <- D_inv %*% RC %*% D_inv
  list(D = D_sqrt, R = R)
}

# Apply to time series of realized covariance matrices
drd_list <- lapply(seq_len(nrow(rc_dates)), function(t) {
  extract_drd(rc_array[t,,])
})
```

# The Final Horse Race

## All Methods Against Each Other

The **global minimum variance portfolio** as a unified testbed — no return forecasting required, pure covariance estimation contest

**GMVP objective:**

$$w_t^* = \arg\min_w w'\hat\Sigma_t w \quad \text{s.t. } \mathbf{1}'w = 1$$

**Competing estimators:**

| # | Method | Lectures |
|---|--------|---------|
| 1 | Naive 1/N | Benchmark |
| 2 | Rolling 252-day sample covariance | 1 |
| 3 | EWMA ($\lambda = 0.94$) diagonal + sample correlation | 1 |
| 4 | GARCH diagonal + sample correlation | 2 |
| 5 | DCC-GARCH | 5 |
| 6 | Ledoit-Wolf nonlinear shrinkage (rolling) | 5 |
| 7 | DCC + LW shrinkage | 5 |
| 8 | HAR-RV diagonal + sample correlation | 4 |
| 9 | Realized covariance (5-min, refresh-time) | 6 |
| 10 | Realized covariance + LW shrinkage | 6 |
| 11 | Element-wise HAR on realized covariance | 6 |

## Portfolio Evaluation

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# Full out-of-sample horse race
# Assumes all Sigma_t_list: named list of T x N x N arrays

eval_gmvp <- function(Sigma_array, returns, label) {
  T <- nrow(returns)
  w_list <- vector("list", T)
  port_rets <- numeric(T)

  for (t in 1:(T-1)) {
    Sigma_t <- Sigma_array[t,,]
    # Ensure PD
    Sigma_t <- as.matrix(nearPD(Sigma_t)$mat)
    w <- solve(Sigma_t, rep(1, ncol(Sigma_t)))
    w <- w / sum(w)
    w_list[[t]] <- w
    port_rets[t+1] <- sum(w * returns[t+1,])
  }

  tibble(
    method     = label,
    ann_var    = var(port_rets, na.rm = TRUE) * 252,
    ann_ret    = mean(port_rets, na.rm = TRUE) * 252,
    ann_sharpe = mean(port_rets, na.rm = TRUE) / sd(port_rets, na.rm = TRUE) * sqrt(252),
    turnover   = mean(sapply(2:T, function(t)
                   sum(abs(w_list[[t]] - w_list[[t-1]]))))
  )
}

results_all <- bind_rows(lapply(names(sigma_list), function(m) {
  eval_gmvp(sigma_list[[m]], daily_returns, label = m)
}))

# Display results
results_all |>
  arrange(ann_var) |>
  mutate(across(where(is.numeric), round, 4)) |>
  knitr::kable()
```

## Python
```{python}
#| eval: false
def eval_gmvp(sigma_list_entry, returns, label):
    T, N = returns.shape
    port_rets = np.zeros(T)
    weights_prev = np.ones(N) / N
    turnovers = []

    for t in range(T-1):
        Sigma_t = sigma_list_entry[t]
        # Ensure PD via eigenvalue clipping
        eigs, vecs = np.linalg.eigh(Sigma_t)
        eigs = np.maximum(eigs, 1e-8)
        Sigma_t = vecs @ np.diag(eigs) @ vecs.T

        ones = np.ones(N)
        w = np.linalg.solve(Sigma_t, ones)
        w /= w.sum()
        port_rets[t+1] = w @ returns[t+1]
        turnovers.append(np.sum(np.abs(w - weights_prev)))
        weights_prev = w

    return {
        "method":     label,
        "ann_var":    np.var(port_rets[1:]) * 252,
        "ann_sharpe": np.mean(port_rets[1:]) / np.std(port_rets[1:]) * np.sqrt(252),
        "turnover":   np.mean(turnovers)
    }
```
:::

## Interpreting the Results

**Expected ranking pattern:**

1. Methods that use intraday information ($RC$, HAR-RV, realized kernel) should beat daily-only methods (GARCH, EWMA, rolling)
2. Shrinkage matters most when $N$ is large — nonlinear LW should beat linear LW
3. DCC beats CCC (time-varying correlations help), but the margin may be small vs. good shrinkage
4. Raw realized covariance is noisy without regularisation — $RC$ + LW should beat $RC$ alone
5. HAR forecasts of covariance improve on same-day realized covariance for next-day portfolios

**Transaction costs:**

- Higher-frequency estimators (realized cov) tend to generate more portfolio turnover
- After deducting transaction costs (assumed 5–10bps per trade), the ranking can shift
- Robust strategies balance accuracy with stability: shrinkage helps here

**Statistical significance:**

Use the Ledoit-Wolf (2008) variance equality test or the model confidence set (MCS) on realised portfolio variances to assess which differences are statistically meaningful

## Synthesis: What Have We Learned?

| Dimension | Key finding |
|-----------|------------|
| Data richness | Intraday data (realized measures) >> daily returns for variance estimation |
| Forecasting | HAR captures long memory simply; beats GARCH at most horizons |
| Dimensionality | Shrinkage is *essential* for $N = 50$; DCC adds on top |
| Jump risk | Transient; remove via BPV before HAR/DCC estimation |
| Correlations | Time-varying; spike in crises → diversification fails exactly when needed |
| Portfolio | GMVP is the clean testbed: better $\hat\Sigma_t$ → lower out-of-sample variance |
| Transaction costs | Accuracy must be balanced against portfolio turnover |

**The central message of the block:**

> *Optimal portfolio choice requires a good estimate of $\Sigma_t$. Volatility modelling is not an end in itself — it is a means to improve the quality of portfolio decisions. Every methodological choice (GARCH vs HAR, shrinkage intensity, sampling frequency) should be evaluated on its ultimate contribution to out-of-sample portfolio performance.*

# Exercises

## Exercise Set 6

**Realized covariance construction**

1. Using LOBSTER data, construct daily realized covariance matrices for a subset of **10 stocks** over a 3-month period:
   - Naive synchronisation (5-minute grid, simultaneous sampling)
   - Refresh-time synchronisation using `highfrequency::refreshTime()`
   - Compare the resulting matrices: how different are the off-diagonal elements? Which has more negative eigenvalues?

2. Demonstrate the Epps effect for one pair of stocks: plot the realized correlation as a function of sampling frequency from 1 second to 30 minutes.

**Positive definiteness**

3. Check which fraction of your daily realized covariance matrices (naive and refresh-time) are positive definite.
   - Apply nearest-PD projection and eigenvalue clipping. How different are the adjusted matrices from the raw ones?
   - Apply Ledoit-Wolf nonlinear shrinkage to the realized covariance. Does this solve the PD problem automatically?

**HAR forecasting for covariance matrices**

4. Using VOLARE data (pre-computed) for all 50 stocks, estimate element-wise HAR models for the realized covariance matrix.
   - Report the cross-sectional distribution of $R^2$ across all $N(N+1)/2$ element-wise regressions. Are off-diagonal elements easier or harder to forecast than diagonal ones?

**The full horse race**

5. Bring together all methods from Lectures 1–6:
   - Methods 1–8 from the table above (using pre-computed estimates from previous exercises)
   - Add realized covariance + LW shrinkage and element-wise HAR-RCov
   - Evaluate using annualised out-of-sample portfolio variance, Sharpe ratio, and daily turnover
   - Apply the model confidence set (MCS) to the realized portfolio variances — which methods survive?

6. **Transaction costs sensitivity:** assume proportional costs of $c \in \{0, 5, 10, 20\}$ bps per one-way trade. Compute net-of-costs portfolio returns for each method. Does the ranking change significantly?

**Key references:** Barndorff-Nielsen, Hansen, Lunde & Shephard (2011); Chiriac & Voev (2011); De Nard, Ledoit & Wolf (2021)
