---
subtitle: "Volatility V: Multivariate Volatility — DCC-GARCH & Shrinkage"
---

```{r}
#| echo: false
#| warning: false
#| eval: false
library(tidyverse)
library(tsmarch)
library(nlshrink)
```

## Where We Are

**Lectures 1–4** focused on marginal variances $\sigma_{i,t}^2$ — the diagonal of $\Sigma_t$. But the GMVP requires the **full covariance matrix**:

$$w^* = \frac{\Sigma_t^{-1} \mathbf{1}}{\mathbf{1}' \Sigma_t^{-1} \mathbf{1}}$$

For $N = 50$ stocks: $\Sigma_t$ has 1,275 unique elements, all estimated from the same $T \approx 2{,}500$ daily observations. For $N = 100$: 5,050 elements from 2,500 observations — more parameters than data. This is the dimensionality problem.

**Two complementary approaches this lecture:**

| Method | Core idea |
|--------|-----------|
| DCC-GARCH | Model conditional correlations dynamically with just 2 parameters |
| Ledoit-Wolf shrinkage | Regularise the sample covariance — essential when $N$ is large |

Both can be combined. In fact, the best practical strategy uses both.

## Why the Sample Covariance Fails

$$\hat\Sigma = \frac{1}{T-1}\sum_{t=1}^T (r_t - \hat\mu)(r_t - \hat\mu)'$$

The sample covariance is a poor estimator for portfolio optimization. Even when $N < T$, inverting $\hat\Sigma$ *amplifies* estimation errors: small eigenvalues of $\hat\Sigma$ become large eigenvalues of $\hat\Sigma^{-1}$, producing extreme and unstable portfolio weights.

The deeper issue is **eigenvalue bias**: under the Marčenko-Pastur law, even if the true covariance is the identity matrix, sample eigenvalues spread out. The largest are over-estimated, the smallest are pushed toward zero. And $\hat\Sigma^{-1}$ is dominated by the smallest eigenvalues — exactly the noisiest directions.

BEKK (Engle & Kroner 1995) models the full covariance matrix directly but has $O(N^2)$ parameters — infeasible for $N \geq 5$. We need structure.

# The Curse of Dimensionality

## Interactive: The Marčenko-Pastur Distribution

*Guided task: Start at c=0.02 — this represents our case of 50 stocks and 2,500 days. Read off the inflation factor: how much does the smallest eigenvalue shrink, and by how much does inverting it inflate portfolio weights? Now drag to c=0.50 (equivalent to 100 stocks and 200 days). Watch the distribution spread dramatically. This is why every large institutional portfolio uses shrinkage — the problem is not just theoretical, it is severe in practice.*

```{ojs}
//| echo: false
viewof c_rat = Inputs.range([0.02, 0.97], {step: 0.01, value: 0.20,
  label: "c = N / T  (concentration ratio, e.g. 0.02 = 50 stocks / 2500 days)"})
```

```{ojs}
//| echo: false
mp_pts = {
  const lmin = (1 - Math.sqrt(c_rat))**2
  const lmax = (1 + Math.sqrt(c_rat))**2
  return Array.from({length: 400}, (_, i) => {
    const lam = lmin + (lmax - lmin) * (i + 0.5) / 400
    const dens = Math.sqrt(Math.max((lmax - lam) * (lam - lmin), 0)) /
                 (2 * Math.PI * c_rat * lam)
    return { lam, dens }
  })
}
```

```{ojs}
//| echo: false
{
  const lmin = ((1 - Math.sqrt(c_rat))**2).toFixed(3)
  const lmax = ((1 + Math.sqrt(c_rat))**2).toFixed(3)
  const inv_inflate = (1 / (1 - Math.sqrt(c_rat))**2).toFixed(1)
  const note = html`Eigenvalues spread: [${lmin}, ${lmax}] &ensp;|&ensp;
    Smallest eigenvalue ${lmin} → Σ⁻¹ inflates by <b>${inv_inflate}×</b> &ensp;|&ensp;
    True value = 1 (all eigenvalues should be equal)`

  const plt = Plot.plot({
    width: 700, height: 265, marginLeft: 55,
    marks: [
      Plot.areaY(mp_pts, {x: "lam", y: "dens",
        fill: "#901A1E", fillOpacity: 0.25}),
      Plot.lineY(mp_pts, {x: "lam", y: "dens",
        stroke: "#901A1E", strokeWidth: 2.5}),
      Plot.ruleX([1], {stroke: "#3d3d3d", strokeDasharray: "4 2", strokeWidth: 1.5}),
      Plot.text([{x: 1.04, y: mp_pts.reduce((a,d) => Math.max(a, d.dens), 0) * 0.55,
                  label: "True λ = 1"}],
        {x: "x", y: "y", text: "label", fontSize: 11, fill: "#3d3d3d"}),
    ],
    x: {label: "Sample eigenvalue λ̂  (true value = 1)", grid: true},
    y: {label: "Marčenko-Pastur density", grid: true},
  })
  return html`${note}<br>${plt}`
}
```

# DCC-GARCH

## The DCC Model (Engle 2002)

**Dynamic Conditional Correlation:** decompose the covariance matrix as

$$\Sigma_t = D_t R_t D_t$$

where $D_t = \text{diag}(\sigma_{1,t}, \ldots, \sigma_{N,t})$ and $R_t$ is the conditional correlation matrix.

**Two-step estimation — this is the key practical insight:**

**Step 1 (Univariate GARCH):** estimate GJR-GARCH for each of the $N$ stocks separately. This is exactly what we did in Lecture 2 — those estimates are reused here. Fully parallelisable: $N$ independent optimisations.

**Step 2 (Correlation dynamics):** with standardised residuals $\hat z_{i,t} = r_{i,t} / \hat\sigma_{i,t}$, model correlations via:

$$Q_t = (1 - a - b)\bar Q + a\, \hat z_{t-1} \hat z_{t-1}' + b\, Q_{t-1}$$

$$R_t = \text{diag}(Q_t)^{-1/2}\, Q_t\, \text{diag}(Q_t)^{-1/2}$$

where $\bar Q = \text{Cov}(\hat z_t)$. Only **two parameters** $(a, b)$ govern all $N(N-1)/2 = 1{,}225$ pairwise correlations. This is the massive dimension reduction that makes DCC feasible.

## DCC in Practice

Empirically: $a \approx 0.03$–$0.06$, $b \approx 0.92$–$0.96$ — correlations are highly persistent ($a + b \approx 0.97$). Correlations **spike in crises** — COVID March 2020, GFC 2008 — and diversification benefits collapse precisely when they are most needed. Intra-sector correlations (tech-tech) are persistently higher than cross-sector.

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
library(tsmarch)

# Step 1: reuse univariate GJR-GARCH fits from Lecture 2
dcc_spec <- dcc_modelspec(
  uspec = multi_fit,  # list of 50 tsgarch fit objects
  type  = "dcc",
  order = c(1, 1)
)
dcc_fit <- estimate(dcc_spec)

# Extract conditional correlations and covariances
R_t     <- rcor(dcc_fit)   # T x N x N array
Sigma_t <- rcov(dcc_fit)   # T x N x N array
```

## Python
```{python}
#| eval: false
# Step 1: univariate GARCH fits (Lecture 2)
Z = pd.DataFrame({sym: res.std_resid for sym, res in garch_results.items()})

# Step 2: DCC — two-parameter correlation recursion
Q_bar = Z.cov().values

def dcc_filter(Z, a, b, Q_bar):
    T, N = Z.shape
    Q = Q_bar.copy()
    R_list = []
    for t in range(T):
        z = Z.iloc[t].values
        Q = (1 - a - b) * Q_bar + a * np.outer(z, z) + b * Q
        d = np.sqrt(np.diag(Q))
        R_list.append(Q / np.outer(d, d))
    return R_list
```
:::

::: {.callout-note icon=false}
**Portfolio implication:** using DCC instead of a constant correlation matrix ($a = b = 0$) allows the portfolio to reflect the crisis correlation spike in real time. A model with constant correlations will underestimate portfolio variance exactly when markets co-move most.
:::

# Covariance Matrix Regularisation

## Ledoit-Wolf Shrinkage

**The core idea (Ledoit & Wolf 2004):** rather than using $\hat\Sigma$ directly, shrink it toward a well-conditioned **target** $F$:

$$\hat\Sigma^{LW} = (1 - \hat\alpha)\hat\Sigma + \hat\alpha F$$

The optimal shrinkage intensity $\hat\alpha$ is estimated analytically — no cross-validation needed. Common targets include a scaled identity matrix (maximal regularisation) or a constant-correlation matrix.

**Nonlinear shrinkage (Ledoit & Wolf 2017, 2020):** the linear version applies the same pull to all eigenvalues. Nonlinear shrinkage treats each eigenvalue separately, shrinking each one optimally based on the analytical formula for the optimal correction. No cross-validation, optimal for large $N$ and $T$ with $N/T \to c > 0$, and consistently outperforms linear shrinkage for $N > 30$.

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
library(nlshrink)

# Nonlinear analytical shrinkage (Ledoit & Wolf 2020)
Sigma_lw <- nlshrink_cov(returns_matrix, k = 1)
# k = 1: spike removal — appropriate when a market factor exists
```

## Python
```{python}
#| eval: false
from nonlinear_shrinkage import analytical_shrinkage

Sigma_lw = analytical_shrinkage(returns_matrix)
```
:::

::: {.callout-note icon=false}
**Portfolio implication:** shrinkage stabilises portfolio weights by pulling extreme eigenvalues toward sensible values. For our 50-stock universe, nonlinear LW shrinkage reduces GMVP turnover and out-of-sample variance compared to raw sample covariance — even before adding DCC dynamics.
:::

# DCC + Shrinkage: The Practical Combination

## Why You Need Both

DCC and shrinkage address different problems:

| Problem | DCC addresses | Shrinkage addresses |
|---------|--------------|---------------------|
| Time-varying correlations | Yes | No |
| Estimation error in $\hat\Sigma$ | Partially | Yes — this is its purpose |
| Works when $N \approx T$ | Needs care | Yes — designed for this |
| Computational cost | Medium | Very fast |

**The practical combination:** apply Ledoit-Wolf nonlinear shrinkage to the DCC correlation matrix $R_t$ or to the target $\bar Q$:

$$\hat\Sigma_t^{DCC+LW} = \hat D_t \left[(1-\hat\alpha)\hat R_t + \hat\alpha I\right] \hat D_t$$

This captures time-varying correlations (DCC) while regularising the matrix to prevent the inversion from amplifying noise (LW). For $N = 50$, this combination consistently dominates either method alone in out-of-sample portfolio performance.

## Portfolio Horse Race Preview

*The final numbers come in Lecture 6. Here is the current standing (illustrative, representative of typical findings):*

| Method | Annualised GMVP vol |
|--------|---------------------|
| Naive 1/N | 16.8% |
| Rolling SD + sample $\hat C$ | 13.2% |
| EWMA diagonal + sample $\hat C$ | 12.9% |
| GARCH diagonal + sample $\hat C$ | 12.4% |
| DCC-GARCH | 11.8% |
| Ledoit-Wolf nonlinear shrinkage | 11.5% |
| DCC + LW shrinkage | 11.1% |

Realized covariance methods (Lecture 6) will push further. The key finding already visible: **better volatility modelling helps, but shrinkage is equally important** — controlling estimation error in the correlation matrix matters as much as modelling its dynamics.

## Bridge to Lecture 6

We have modelled $\Sigma_t$ using daily returns — both GARCH-based variances and DCC correlations. **Lecture 6** asks: what if we use intraday cross-products instead? The realized covariance matrix provides the most informationally rich estimate of $\Sigma_t$, potentially dominating DCC in portfolio performance. Non-synchronous trading introduces new challenges, but refresh-time sampling and shrinkage solve them.

# Exercises

## Exercise Set 5

**DCC-GARCH**

1. Using `tsmarch`, estimate DCC-GARCH for the 50 NASDAQ stocks, building on your Lecture 2 GJR-GARCH fits. Report the DCC parameters $(a, b)$ and persistence $a + b$. Plot conditional correlations for three pairs: two tech stocks, a tech-healthcare pair, and a tech-consumer pair. Do correlations spike during COVID (Feb–Mar 2020)?

2. Identify the 5 dates with the highest average pairwise conditional correlation. Do they coincide with known market stress events?

**Shrinkage**

3. Apply nonlinear Ledoit-Wolf shrinkage to the rolling 252-day sample covariance matrix. Plot the eigenvalue distribution before and after — how much does the smallest eigenvalue change? Are the portfolio weights more stable (less extreme) under shrinkage?

**Portfolio evaluation**

4. Construct the GMVP using each estimator in a rolling out-of-sample evaluation (5-year training, 2-year test): naive 1/N, rolling 252-day sample covariance, GARCH diagonal + sample correlation, DCC-GARCH, Ledoit-Wolf nonlinear shrinkage, and DCC + LW. Report annualised realised portfolio variance and daily turnover. Is the DCC + LW combination the best?

**Key references:** Engle (2002); Ledoit & Wolf (2004, 2020)
