---
subtitle: "Volatility IV: Forecasting Realized Volatility — HAR Models"
---

```{r}
#| echo: false
#| warning: false
#| eval: false
library(tidyverse)
library(highfrequency)
library(sandwich)
library(lmtest)
```

## Where We Are

- **Lecture 1:** volatility is latent, exhibits clustering, fat tails, leverage, and mean reversion
- **Lecture 2:** GARCH extracts a signal from daily returns alone
- **Lecture 3:** realized variance $RV_t$ from intraday data is a precise, low-noise estimate of $\sigma_t^2$

**This lecture:** we have a clean time series $\{RV_t\}$ for each of our 50 NASDAQ stocks — now we **forecast** it.

At time $t$, the portfolio optimizer needs $E(\sigma_{i,t+1}^2 \mid \mathcal{F}_t)$ — a forecast, not today's realized value. HAR provides that forecast cheaply and accurately: **it is just OLS on rolling averages**. The beauty lies in its simplicity.

$$\hat\Sigma_{t+1|t} = \hat D_{t+1|t} \hat C \hat D_{t+1|t}$$

HAR provides the diagonal $\hat D_{t+1|t}$; DCC (Lecture 5) will provide time-varying $\hat C_{t+1|t}$.

## The Heterogeneous Market Hypothesis

**Müller et al. (1997):** financial markets are populated by agents operating at different time horizons:

- **Daily traders** — short-term speculators and market makers react to yesterday's news
- **Weekly portfolio managers** — institutional investors update based on the past week
- **Monthly investors** — pension funds are almost insensitive to daily fluctuations

Each group's volatility perception operates at their own horizon. The observed **long-memory-like behaviour** in $RV_t$ — slow decay in the ACF — arises naturally from this superposition of shocks at different time scales.

The HAR model operationalises this idea with a regression any econometrics student can estimate in one line.

# The HAR Model

## HAR-RV Specification

**Corsi (2009):** the Heterogeneous AutoRegressive model for Realized Variance

$$RV_{t+1} = \beta_0 + \beta_d\, RV_t^{(d)} + \beta_w\, RV_t^{(w)} + \beta_m\, RV_t^{(m)} + \varepsilon_{t+1}$$

where:

$$RV_t^{(d)} = RV_t, \quad RV_t^{(w)} = \frac{1}{5}\sum_{j=0}^{4} RV_{t-j}, \quad RV_t^{(m)} = \frac{1}{22}\sum_{j=0}^{21} RV_{t-j}$$

Three regressors, one equation. $\beta_d$ captures the daily component, $\beta_w$ the weekly component, $\beta_m$ the monthly component. All three are typically positive and significant — volatility is autocorrelated at every horizon.

HAR is a restricted AR(22) estimated with only **3 parameters**. The averaging structure imposes a specific slow-decay pattern that matches the empirical autocorrelation of $RV_t$ — mimicking long memory without estimating a fractional integration parameter.

## Interactive: HAR vs GARCH — How Shocks Decay

*Guided task: Set β_d=0.36, β_w=0.34, β_m=0.18 (typical NASDAQ estimates). Count the days until the HAR impulse response falls below 0.50 — that is the half-life. Now set GARCH persistence=0.97 and read off its half-life. Which model says shocks decay faster? Now try persistence=0.97 for GARCH and notice that at short horizons (days 1–10) HAR decays more slowly. What does this mean for position sizing after a volatility spike: would a HAR-based manager hold a more defensive position for longer than a GARCH-based manager?*

```{ojs}
//| echo: false
viewof bd = Inputs.range([0.0, 0.65], {step: 0.01, value: 0.36,
  label: "β_d  (daily component)"})
viewof bw = Inputs.range([0.0, 0.65], {step: 0.01, value: 0.34,
  label: "β_w  (weekly component)"})
viewof bm = Inputs.range([0.0, 0.40], {step: 0.01, value: 0.18,
  label: "β_m  (monthly component)"})
viewof gp = Inputs.range([0.85, 0.999], {step: 0.001, value: 0.970,
  label: "GARCH persistence  α + β"})
```

```{ojs}
//| echo: false
har_ir = {
  const H = 65
  const rv = new Float64Array(H + 23).fill(0)
  rv[0] = 1  // unit shock
  const ir = []
  for (let t = 1; t <= H; t++) {
    const n_w = Math.min(t, 5), n_m = Math.min(t, 22)
    let sum_w = 0, sum_m = 0
    for (let k = 1; k <= n_w; k++) sum_w += rv[t - k]
    for (let k = 1; k <= n_m; k++) sum_m += rv[t - k]
    rv[t] = bd * rv[t-1] + bw * (sum_w / n_w) + bm * (sum_m / n_m)
    ir.push({ h: t, har: rv[t], garch: gp**t,
              persist_ok: (bd + bw + bm) < 1 })
  }
  return ir
}
```

```{ojs}
//| echo: false
{
  const total = bd + bw + bm
  const warning = total >= 1
    ? html`<span style='color:#901A1E'>⚠ β_d + β_w + β_m = ${total.toFixed(2)} ≥ 1 — non-stationary</span>`
    : html`β_d + β_w + β_m = ${total.toFixed(2)} &ensp; (< 1 ✓ stationary)`

  const plt = Plot.plot({
    width: 700, height: 265, marginLeft: 55,
    marks: [
      Plot.ruleY([0], {stroke: "#dddddd"}),
      Plot.lineY(har_ir, {x: "h", y: "garch",
        stroke: "#666666", strokeWidth: 2, strokeDasharray: "5 3"}),
      Plot.lineY(har_ir, {x: "h", y: "har",
        stroke: "#901A1E", strokeWidth: 2.5}),
      Plot.areaY(har_ir, {x: "h", y1: "har", y2: "garch",
        fill: "#901A1E", fillOpacity: 0.08}),
    ],
    x: {label: "Horizon h (days after shock)", grid: true},
    y: {label: "Impulse response (fraction of shock remaining)", grid: true},
    caption: html`<span style='color:#901A1E'>——</span> HAR &ensp;
      <span style='color:#666666'>– –</span> GARCH &ensp;
      Shaded area = HAR forecasts more persistent volatility than GARCH`
  })
  return html`${warning}<br>${plt}`
}
```

## Estimation: Just OLS

OLS with Newey-West standard errors (to account for autocorrelation in $\varepsilon_{t+1}$):

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# rv_data: columns date, symbol, rv, bpv (from VOLARE)

har_data <- rv_data |>
  filter(symbol == "AAPL") |>
  arrange(date) |>
  mutate(
    rv_d    = rv,
    rv_w    = slider::slide_dbl(rv, mean, .before = 4,  .complete = TRUE),
    rv_m    = slider::slide_dbl(rv, mean, .before = 21, .complete = TRUE),
    rv_lead = lead(rv)
  ) |>
  drop_na()

har_fit <- lm(rv_lead ~ rv_d + rv_w + rv_m, data = har_data)
coeftest(har_fit, vcov = NeweyWest(har_fit, lag = 5))
```

## Python
```{python}
#| eval: false
import statsmodels.formula.api as smf

df = (rv_data[rv_data["symbol"] == "AAPL"]
      .sort_values("date")
      .assign(rv_d    = lambda x: x["rv"],
              rv_w    = lambda x: x["rv"].rolling(5).mean(),
              rv_m    = lambda x: x["rv"].rolling(22).mean(),
              rv_lead = lambda x: x["rv"].shift(-1))
      .dropna())

har_fit = smf.ols("rv_lead ~ rv_d + rv_w + rv_m", data=df).fit()
print(har_fit.get_robustcov_results(cov_type="HAC", maxlags=5).summary())
```
:::

## Extensions: HAR-CJ

**HAR-J** adds the jump component $J_t = \max(RV_t - BPV_t, 0)$ as an additional predictor. The typical finding: $\hat\beta_J$ is small and often insignificant — jumps are transient and do not forecast future variance well.

**HAR-CJ** goes further, separating continuous and jump components as distinct regressors:

$$RV_{t+1} = \beta_0 + \beta_{Cd} C_t^{(d)} + \beta_{Cw} C_t^{(w)} + \beta_{Cm} C_t^{(m)} + \beta_J J_t + \varepsilon_{t+1}$$

where $C_t = BPV_t$ (the continuous component). This matters for portfolios: the continuous component is persistent and should feed into risk forecasts; jump variation is transient and should not inflate multi-day risk estimates.

```{r}
#| eval: false
har_cj <- HARmodel(
  data    = cbind(rv = rv_aapl, bpv = bpv_aapl),
  periods = c(1, 5, 22),
  type    = "HARCJ",
  h       = 1
)
summary(har_cj)
```

::: {.callout-note icon=false}
**Portfolio implication:** on jump days, substituting $BPV_t$ for $RV_t$ as the volatility input avoids spuriously inflated risk estimates — and avoids the unnecessary defensive positioning that would follow.
:::

# GARCH vs HAR: The Key Comparison

## Why HAR Usually Wins

| Feature | GARCH(1,1) | HAR-RV |
|---------|-----------|--------|
| Input data | Daily returns $r_t$ | Intraday realized variance $RV_t$ |
| Estimation | MLE/QML | OLS — trivial |
| Memory structure | Exponential decay | Slow decay via weekly/monthly avg |
| Data requirement | Daily prices | Intraday data or VOLARE |
| SNR of input | $\approx 0.05$ (very noisy) | High — $RV_t \approx \sigma_t^2$ |

**Empirical finding:** HAR dominates at daily and weekly horizons because $RV_t$ is essentially a direct observation of $\sigma_t^2$ — the regression signal-to-noise ratio is dramatically higher. GARCH is competitive only when intraday data is unavailable.

The forecasting evaluation uses **QLIKE loss** with $RV_t$ as the proxy for the true variance: $\text{QLIKE} = RV_{t+1}/\hat\sigma_{t+1|t}^2 - \log(RV_{t+1}/\hat\sigma_{t+1|t}^2) - 1$. This loss is proper and robust to the fact that $RV_t$ is itself noisy.

## Cross-Section: HAR Across 50 Stocks

Estimate HAR-RV for all 50 stocks from VOLARE. Across-stock patterns to look for: tech stocks often show stronger daily persistence ($\hat\beta_d$); large caps tend to have a more prominent weekly component ($\hat\beta_w$); $R^2$ is higher for more liquid stocks because $RV_t$ is less noisy.

```{r}
#| eval: false
library(future.apply)
plan(multisession, workers = 8)

har_fits <- future_lapply(unique(volare_data$symbol), function(sym) {
  df <- volare_data |>
    filter(symbol == sym) |> arrange(date) |>
    mutate(rv_d = rv,
           rv_w = slider::slide_dbl(rv, mean, .before = 4,  .complete = TRUE),
           rv_m = slider::slide_dbl(rv, mean, .before = 21, .complete = TRUE),
           rv_lead = lead(rv)) |>
    drop_na()
  lm(rv_lead ~ rv_d + rv_w + rv_m, data = df)
})
names(har_fits) <- unique(volare_data$symbol)
```

Across the NASDAQ 50, the sum $\hat\beta_d + \hat\beta_w + \hat\beta_m$ is typically in the range 0.85–0.95, confirming the strong persistence of realized variance. The monthly component $\hat\beta_m$ is usually smaller than the daily and weekly components, but reliably positive — pension fund–horizon investors leave a detectable imprint in the autocorrelation structure.

## Bridge to Lectures 5 & 6

We now have precise univariate forecasts $\hat\sigma_{i,t+1|t}^2$ for all 50 stocks. But the portfolio optimizer needs the **full covariance matrix** $\hat\Sigma_{t+1|t}$.

Two complementary approaches:

1. **HAR diagonal + DCC correlations** (Lecture 5): use HAR for $\hat D_{t+1|t}$, DCC for time-varying $\hat C_{t+1|t}$
2. **Realized covariance** (Lecture 6): estimate the full $\hat\Sigma_{t+1|t}$ directly from intraday cross-products, then forecast element-wise with HAR

# Exercises

## Exercise Set 4

**HAR estimation**

1. Using VOLARE data for 5 stocks, estimate HAR-RV and HAR-CJ. Is the jump coefficient $\hat\beta_J$ statistically significant? Does separating continuous and jump components improve in-sample $R^2$?

2. Apply the HAR impulse response interactive with your estimated coefficients. How does the implied half-life compare to a GARCH model with persistence 0.97? What does this mean for how long a manager should remain defensively positioned after a volatility spike?

**Cross-section**

3. Estimate HAR-RV for all 50 stocks. Summarise coefficients $(\hat\beta_d, \hat\beta_w, \hat\beta_m)$ by sector. Do tech stocks show more daily persistence? What do the patterns suggest about different investor horizons across sectors?

**Portfolio evaluation**

4. Combine HAR-RV forecasts for $\hat D_{t+1|t}$ with the constant sample correlation matrix. Construct the GMVP and compare out-of-sample realised portfolio variance to GARCH-based $\hat D_t$ (Lecture 2) and rolling SD (Lecture 1). Does better volatility forecasting translate into better portfolio performance?

**Key references:** Corsi (2009); Andersen, Bollerslev & Diebold (2007); Patton & Sheppard (2015)
