---
subtitle: "Volatility II: GARCH Models in Practice"
---

```{r}
#| echo: false
#| warning: false
#| eval: false
library(tidyverse)
library(tsgarch)
library(future)
library(future.apply)
```

## Where We Are

**Lecture 1 established:** volatility is latent, time-varying, and exhibits clustering, fat tails, leverage, and mean reversion — four facts any good model must respect

**EWMA shortcomings:**

- $\lambda$ is not estimated from data — it is hard-coded at 0.94
- Symmetric — treats a -3% day and a +3% day identically (ignores the leverage effect)
- No mean reversion — $\alpha + \beta = 1$ means forecasts stay elevated forever after a shock

**This lecture:** GARCH fixes all three — likelihood-based estimation, asymmetric extensions (GJR), and mean reversion by construction

$$\hat\Sigma_t = \hat D_{t|t-1} \hat C_t \hat D_{t|t-1}, \quad \hat D_{t|t-1} = \text{diag}(\hat\sigma_{1,t|t-1}, \ldots, \hat\sigma_{N,t|t-1})$$

The $\hat\sigma_{i,t|t-1}$ will come from GARCH; handling $\hat C_t$ is the job of Lecture 5 (DCC)

# GARCH Models

## GARCH(1,1) — The Workhorse

The **Generalised ARCH** model (Bollerslev 1986):

$$r_t = \sigma_t \varepsilon_t, \quad \varepsilon_t \overset{i.i.d.}{\sim} \mathcal{D}(0, 1)$$

$$\sigma_t^2 = \omega + \alpha r_{t-1}^2 + \beta \sigma_{t-1}^2$$

- $\omega > 0$: intercept — anchors the long-run variance
- $\alpha \geq 0$: **ARCH effect** — how much yesterday's squared return raises today's variance
- $\beta \geq 0$: **GARCH effect** — how much yesterday's variance persists into today
- Stationarity: $\alpha + \beta < 1$; unconditional variance: $\bar\sigma^2 = \omega / (1 - \alpha - \beta)$

**Persistence:** $\alpha + \beta \approx 0.97$–$0.99$ for most equities — shocks decay slowly but they do decay

**Connection to EWMA:** setting $\omega = 0$ and $\alpha + \beta = 1$ gives exactly the EWMA recursion — every time you use EWMA you are implicitly using a non-mean-reverting GARCH with the intercept forced to zero.

## Capturing the Leverage Effect: GJR-GARCH

**Why:** Fact 4 from Lecture 1 — negative shocks raise variance more than positive shocks of the same size

$$\sigma_t^2 = \omega + (\alpha + \gamma\, \mathbf{1}_{r_{t-1}<0})\, r_{t-1}^2 + \beta\, \sigma_{t-1}^2$$

- $\gamma > 0$: the extra impact when the news is bad
- Negative shock: total ARCH coefficient is $\alpha + \gamma$
- Positive shock: total ARCH coefficient is $\alpha$
- Stationarity: $\alpha + \beta + \gamma/2 < 1$

**Hansen & Lunde (2005):** no model systematically beats GARCH(1,1) on symmetric data — but asymmetric models (GJR, EGARCH) consistently beat symmetric GARCH when the leverage effect is present.

**This is our default specification for all 50 stocks.**

## Interactive: Build Your Own News Impact Curve

*Try these experiments:*

**(1)** Set γ = 0 — you have a symmetric GARCH. The curve should be a U-shape centred on zero.

**(2)** Increase γ to 0.20. What happens to the left arm (bad news) vs the right arm (good news)?

**(3)** Find a setting of (α, γ, β) where the stationarity condition α + β + γ/2 < 1 is violated. What does the model predict for long-run volatility?

```{ojs}
//| echo: false
viewof a_nic = Inputs.range([0.01, 0.25], {step: 0.005, value: 0.08,
  label: "α  (ARCH — symmetric reaction to news)"})
viewof g_nic = Inputs.range([0.00, 0.35], {step: 0.005, value: 0.14,
  label: "γ  (GJR leverage — extra reaction to bad news)"})
viewof b_nic = Inputs.range([0.55, 0.98], {step: 0.01,  value: 0.88,
  label: "β  (GARCH persistence)"})
```

```{ojs}
//| echo: false
nic_curves = {
  const omega = 1e-6
  const denom = 1 - a_nic - g_nic/2 - b_nic
  const s2bar = denom > 0.001 ? omega / denom : 0.04 / 252
  return Array.from({length: 201}, (_, i) => {
    const r = (i - 100) * 0.001
    const garch_var = omega + a_nic * r**2 + b_nic * s2bar
    const gjr_var   = omega + (a_nic + (r < 0 ? g_nic : 0)) * r**2 + b_nic * s2bar
    return {
      r_pct: r * 100,
      garch: Math.sqrt(Math.max(garch_var, 0) * 252) * 100,
      gjr:   Math.sqrt(Math.max(gjr_var,   0) * 252) * 100,
      lrun:  Math.sqrt(s2bar * 252) * 100
    }
  })
}
```

```{ojs}
//| echo: false
Plot.plot({
  width: 700, height: 280, marginLeft: 55,
  marks: [
    Plot.ruleY([nic_curves[0].lrun],
      {stroke: "#cccccc", strokeDasharray: "4 2", strokeWidth: 1.5}),
    Plot.ruleX([0], {stroke: "#dddddd"}),
    Plot.lineY(nic_curves, {x: "r_pct", y: "garch",
      stroke: "#666666", strokeWidth: 2}),
    Plot.lineY(nic_curves, {x: "r_pct", y: "gjr",
      stroke: "#901A1E", strokeWidth: 2.5}),
  ],
  x: {label: "Yesterday's return r_{t−1} (%)", grid: true},
  y: {label: "σ_t — annualised vol (%)", grid: true},
  caption: html`<span style='color:#666666'>——</span> GARCH (symmetric) &ensp;
    <span style='color:#901A1E'>——</span> GJR-GARCH (asymmetric) &ensp;
    <span style='color:#cccccc'>– –</span> Long-run vol`
})
```

## Mean Reversion in GARCH

Multi-step forecasts converge to the unconditional variance:

$$E(\sigma_{t+h}^2 \mid \mathcal{F}_t) = \bar\sigma^2 + (\alpha + \beta)^{h-1}\bigl(\sigma_{t+1|t}^2 - \bar\sigma^2\bigr)$$

Forecasts converge at the geometric rate $(\alpha + \beta)^h$:

- At $\alpha + \beta = 0.97$: half-life $\approx$ **23 trading days**
- At $\alpha + \beta = 1.00$ (EWMA): **no convergence** — forecasts remain flat forever

**Portfolio implication:** long-horizon forecasts from GARCH converge to $\bar\sigma^2$; EWMA stays flat at the shocked level. After a crisis, a pension fund using EWMA stays defensively positioned indefinitely — even as actual volatility subsides.

## Interactive: GARCH Mean Reversion vs EWMA

*Try these experiments:*

**(1)** Set persistence = 0.970, shock = 3×. At what horizon does the forecast return within 10% of the long-run level?

**(2)** Drag persistence toward 0.999. When does GARCH start to resemble the flat EWMA line?

**(3)** At shock = 3× (vol currently 60% annualised): what does GARCH predict one year from now? What does EWMA predict? Which should a pension fund trust?

```{ojs}
//| echo: false
viewof pers = Inputs.range([0.86, 0.999], {step: 0.001, value: 0.970,
  label: "Persistence  α + β"})
viewof shock_m = Inputs.range([1.0, 5.0], {step: 0.1, value: 3.0,
  label: "Current vol / long-run vol  (shock size)"})
```

```{ojs}
//| echo: false
conv_data = {
  const sigma_bar = 20  // long-run annualised vol, %
  const sigma_now = sigma_bar * shock_m
  const half_life = Math.log(0.5) / Math.log(pers)
  const hl_day = Math.round(half_life)
  return Array.from({length: 121}, (_, h) => ({
    h,
    garch: Math.sqrt(sigma_bar**2 + pers**h * (sigma_now**2 - sigma_bar**2)),
    ewma:  sigma_now,
    lrun:  sigma_bar,
    is_hl: h === hl_day
  }))
}
```

```{ojs}
//| echo: false
{
  const hl = conv_data.find(d => d.is_hl)
  const half_life_label = hl ? `½-life ≈ ${hl.h} days` : ""
  return Plot.plot({
    width: 700, height: 280, marginLeft: 55,
    marks: [
      Plot.ruleY([conv_data[0].lrun],
        {stroke: "#cccccc", strokeDasharray: "4 2", strokeWidth: 1.5}),
      Plot.lineY(conv_data, {x: "h", y: "ewma",
        stroke: "#aaaaaa", strokeWidth: 1.8, strokeDasharray: "5 3"}),
      Plot.lineY(conv_data, {x: "h", y: "garch",
        stroke: "#901A1E", strokeWidth: 2.5}),
      hl ? Plot.dot([hl], {x: "h", y: "garch", r: 5, fill: "#901A1E"}) : null,
      hl ? Plot.text([{...hl, label: half_life_label}],
        {x: "h", y: "garch", text: "label", dy: -14, fontSize: 11}) : null,
    ].filter(Boolean),
    x: {label: "Forecast horizon h (trading days)", grid: true},
    y: {label: "E[σ_{t+h} | F_t]  (%)", grid: true},
    caption: html`<span style='color:#901A1E'>——</span> GARCH (mean-reverts to long-run level) &ensp;
      <span style='color:#aaaaaa'>– –</span> EWMA / IGARCH (flat — no mean reversion) &ensp;
      <span style='color:#cccccc'>– –</span> Long-run vol (20%)`
  })
}
```

# Practical Estimation

## Fitting GARCH in Practice

`tsgarch` (Galanos) is the modern R package for GARCH estimation, built on **automatic differentiation via TMB** (Template Model Builder) — analytic gradients give numerical stability and accurate standard errors.

Two choices that matter in practice:

- **Distribution:** skewed Student-$t$ (Hansen 1994) captures both fat tails and return asymmetry; QML with Gaussian errors gives consistent $(\hat\omega, \hat\alpha, \hat\beta)$ but requires robust standard errors
- **Variance targeting:** pins $\hat\omega = \hat{\bar\sigma}^2(1 - \alpha - \beta)$ from the sample variance, reducing the optimisation to two free parameters — dramatically improves convergence

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
library(tsgarch)

# Specify GJR-GARCH(1,1) with skewed-t errors and variance targeting
spec <- garch_modelspec(
  y = returns[, "AAPL"],
  model = "gjrgarch",
  order = c(1, 1),
  constant = TRUE,
  distribution = "sstd",
  variance_targeting = TRUE
)

fit <- estimate(spec)
summary(fit)
```

## Python
```{python}
#| eval: false
from arch import arch_model

# GJR-GARCH(1,1) with Student-t innovations
model = arch_model(
    returns["AAPL"],
    vol="Garch",
    p=1, o=1, q=1,   # o=1 enables the GJR leverage term
    dist="t",
    mean="Zero"
)
result = model.fit(disp="off")
print(result.summary())
```
:::

**Diagnostics to check after every fit:**

1. Ljung-Box on $\hat\varepsilon_t^2$ — no remaining ARCH effects (p-value > 0.05)
2. QQ-plot of $\hat\varepsilon_t$ — distribution assumption adequate?
3. Persistence $\hat\alpha + \hat\beta + \hat\gamma/2 < 1$ — long-run forecasts are reliable

## What We See in the NASDAQ 50

Estimate GJR-GARCH(1,1) for all 50 stocks in parallel — `future_lapply` handles the parallelisation:

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
library(tsgarch)
library(future)
library(future.apply)

plan(multisession, workers = 8)

symbols <- colnames(returns)

multi_spec <- lapply(symbols, function(sym) {
  garch_modelspec(
    y = returns[, sym],
    model = "gjrgarch",
    order = c(1, 1),
    distribution = "sstd",
    variance_targeting = TRUE
  )
})

multi_fit <- future_lapply(multi_spec, estimate)
names(multi_fit) <- symbols

# Extract persistence and leverage for each stock
params_df <- bind_rows(lapply(symbols, function(sym) {
  p <- coef(multi_fit[[sym]])
  tibble(
    symbol      = sym,
    alpha       = p["alpha1"],
    beta        = p["beta1"],
    gamma       = p["gamma1"],
    persistence = p["alpha1"] + p["beta1"] + p["gamma1"] / 2
  )
}))
```

## Python
```{python}
#| eval: false
from arch import arch_model
from joblib import Parallel, delayed
import pandas as pd

def fit_gjr(symbol, returns):
    model = arch_model(returns[symbol], vol="Garch", p=1, o=1, q=1,
                       dist="t", mean="Zero")
    return model.fit(disp="off")

results = Parallel(n_jobs=8)(
    delayed(fit_gjr)(sym, returns) for sym in returns.columns
)

params_df = pd.DataFrame([
    {"symbol": sym,
     "alpha": res.params["alpha[1]"],
     "beta":  res.params["beta[1]"],
     "gamma": res.params.get("gamma[1]", 0),
     "persistence": res.params["alpha[1]"] + res.params["beta[1]"]}
    for sym, res in zip(returns.columns, results)
])
```
:::

**Typical findings across the NASDAQ 50:** persistence concentrates in the 0.96–0.99 range; the leverage coefficient $\hat\gamma > 0$ for the majority of stocks — asymmetry is the rule, not the exception; tech stocks tend to show higher persistence and stronger leverage than healthcare names; a handful of stocks exhibit low persistence, consistent with high idiosyncratic, news-driven volatility.

## From GARCH to Portfolio Weights

With $\hat\sigma_{i,t|t-1}^2$ estimated for all 50 stocks, we form the time-varying diagonal covariance matrix and combine with a constant sample correlation:

$$\hat\Sigma_t = \hat D_{t|t-1} \hat C \hat D_{t|t-1}$$

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# Extract conditional variance matrix (T x 50)
sigma2_matrix <- do.call(cbind, lapply(multi_fit, sigma))^2

corr_sample <- cor(returns)

gmvp_garch <- map_dfr(1:nrow(sigma2_matrix), function(t) {
  sig_vec <- sqrt(sigma2_matrix[t, ])
  D <- diag(sig_vec)
  Sigma_t <- D %*% corr_sample %*% D
  w <- solve(Sigma_t, rep(1, 50))
  w <- w / sum(w)
  tibble(date = index(sigma2_matrix)[t],
         port_var = as.numeric(t(w) %*% Sigma_t %*% w))
})
```

## Python
```{python}
#| eval: false
import numpy as np

sigma2_matrix = pd.DataFrame(
    {sym: res.conditional_volatility**2
     for sym, res in zip(returns.columns, results)}
)

corr_sample = returns.corr().values

gmvp_vars = []
for t in range(len(sigma2_matrix)):
    sig_vec = np.sqrt(sigma2_matrix.iloc[t].values)
    D = np.diag(sig_vec)
    Sigma_t = D @ corr_sample @ D
    w = np.linalg.solve(Sigma_t, np.ones(50))
    w /= w.sum()
    gmvp_vars.append(w @ Sigma_t @ w)
```
:::

**Limitation:** $\hat C$ is constant — DCC-GARCH (Lecture 5) allows $\hat C_t$ to evolve. Preview: using GARCH-based $\hat D_t$ instead of rolling SD typically reduces out-of-sample GMVP variance by **10–20%** before any improvement to the correlation structure.

# Exercises

## Exercise Set 2

**Estimation**

1. Using `tsgarch` (R) or `arch` (Python), estimate GJR-GARCH(1,1) for 5 stocks from the NASDAQ universe. For each: report $(\hat\omega, \hat\alpha, \hat\beta, \hat\gamma)$ with robust standard errors, compute persistence, and assess whether the leverage coefficient $\hat\gamma$ is statistically significant.

2. Check the diagnostics: are the standardised residuals $\hat\varepsilon_t$ free of ARCH effects (Ljung-Box on $\hat\varepsilon_t^2$, $p > 0.05$)? Does the distribution assumption look adequate from the QQ-plot?

3. Plot news impact curves for symmetric GARCH vs GJR-GARCH for two stocks — one tech and one healthcare. How large is the asymmetry, and does it differ by sector?

**Cross-section**

4. Estimate GJR-GARCH(1,1) for all 50 stocks in parallel. Plot the cross-sectional distribution of persistence $\hat\alpha + \hat\beta + \hat\gamma/2$ and the leverage coefficient $\hat\gamma$. Do tech stocks show more leverage than healthcare stocks?

**Portfolio evaluation**

5. Construct the GMVP using GARCH conditional variances combined with the sample correlation matrix. Compare out-of-sample portfolio volatility to the naive estimators from Lecture 1 (rolling 21-day SD, EWMA 0.94). Does time-varying GARCH variance alone reduce realised portfolio variance?

**Key references:** Bollerslev (1986); Glosten, Jagannathan & Runkle (1993); Hansen & Lunde (2005); Engle & Patton (2001)
