---
subtitle: "Volatility VI: Realized Covariance Matrices"
---

```{r}
#| echo: false
#| warning: false
#| eval: false
library(tidyverse)
library(highfrequency)
library(nlshrink)
library(Matrix)
```

## Closing the Loop

| Lecture | Estimator | What it produces |
|---------|-----------|-----------------|
| 1 | Rolling SD, EWMA | $\hat\sigma_{i,t}$, constant $\hat C$ |
| 2 | GARCH, GJR-GARCH | $\hat\sigma_{i,t|t-1}$ from daily $r_t$ |
| 3 | Realized variance | $\widehat{RV}_{i,t}$, precise estimate of $\sigma_{i,t}^2$ |
| 4 | HAR models | $\hat\sigma_{i,t+h|t}^2$, forecasts from realized measures |
| 5 | DCC-GARCH, LW shrinkage | Full $\hat\Sigma_t$ at $N = 50$ |
| **6** | **Realized covariance** | **Full $\hat\Sigma_t$ from intraday cross-products** |

**The natural extension:** if realized variance gives a precise $\hat\sigma_{i,t}^2$ from intraday data, we can similarly estimate the **full covariance matrix** from intraday cross-products.

Three additional challenges: non-synchronous trading biases off-diagonal elements downward, noise propagates into the cross-products, and positive definiteness is not guaranteed. This lecture shows how to handle each.

# Realized Covariance

## Definition

The **realized covariance** matrix on day $t$:

$$\widehat{RC}_t = \sum_{j=1}^n r_{t,j} r_{t,j}'$$

where $r_{t,j}$ is the $N$-dimensional vector of synchronised intraday returns over interval $j$.

As $n \to \infty$, $\widehat{RC}_t$ converges in probability to the **integrated covariance** $IC_t = \int_0^1 \sigma_t(s)\sigma_t(s)'\, ds$ — the true daily covariance matrix. The diagonal elements are exactly the realized variances from Lecture 3. The off-diagonal elements are sums of intraday cross-products.

Multivariate realized kernels (Barndorff-Nielsen et al. 2011) extend this to explicitly correct for noise at all frequencies — they include lead-lag cross-products weighted by a kernel function. In practice, for our 50-stock NASDAQ universe, refresh-time sampling combined with shrinkage achieves similar results more simply.

## The Epps Effect

**Epps (1979):** as sampling frequency increases, estimated correlations between assets *decrease toward zero*. This is a bias, not a feature.

**Cause:** assets do not trade at the same instants. At 1-second sampling, many intervals have a return for only one asset — the other is stale. Stale returns impute last observed price; their cross-product is zero. Averaging many such zero cross-products pulls the correlation estimate toward zero.

The bias is stronger for less liquid stocks. For large-cap NASDAQ stocks, the Epps effect is largely resolved by 5-minute sampling; for smaller caps it may persist even at 10–15 minutes.

## Interactive: The Epps Effect

*Guided task: Set a true correlation of ρ=0.65 (typical for two tech stocks). At 1-second sampling, what does the measured correlation drop to? Now drag τ (inter-trade interval) from 8 seconds to 2 seconds — how does the bias change? This is why liquid pairs recover faster: they trade more often, so fewer sampling intervals have stale prices. The practical lesson: always plot the realized correlation as a function of sampling frequency before reporting it.*

```{ojs}
//| echo: false
viewof rho_true = Inputs.range([0.10, 0.90], {step: 0.05, value: 0.65,
  label: "True correlation  ρ"})
viewof tau_sec = Inputs.range([1, 30], {step: 1, value: 8,
  label: "Avg inter-trade interval  τ̄  (seconds)"})
```

```{ojs}
//| echo: false
epps_curve = {
  // Theoretical Epps correction for Poisson arrivals (Hayashi-Yoshida approximation):
  // ρ_obs(Δ) ≈ ρ * (1 - exp(-Δ / τ̄))
  const n = 200
  return Array.from({length: n}, (_, i) => {
    const delta = 1 + i * 20   // 1s to ~4000s
    const rho_obs = rho_true * (1 - Math.exp(-delta / (2 * tau_sec)))
    return { delta, rho_obs, rho_true_line: rho_true }
  })
}
```

```{ojs}
//| echo: false
{
  const rho_5min = rho_true * (1 - Math.exp(-300 / (2 * tau_sec)))
  const bias_5min = ((rho_true - rho_5min) / rho_true * 100).toFixed(1)

  const plt = Plot.plot({
    width: 700, height: 270, marginLeft: 55,
    marks: [
      Plot.ruleY([rho_true],
        {stroke: "#cccccc", strokeDasharray: "4 2", strokeWidth: 1.5}),
      Plot.ruleX([300],
        {stroke: "#666666", strokeDasharray: "2 2", strokeWidth: 1.2}),
      Plot.text([{delta: 450, rho_obs: rho_true * 0.25, label: "5 min →"}],
        {x: "delta", y: "rho_obs", text: "label", fontSize: 10, fill: "#666666"}),
      Plot.lineY(epps_curve, {x: "delta", y: "rho_obs",
        stroke: "#901A1E", strokeWidth: 2.5}),
      Plot.areaY(epps_curve, {x: "delta", y1: "rho_obs", y2: "rho_true_line",
        fill: "#901A1E", fillOpacity: 0.08}),
    ],
    x: {label: "Sampling interval Δ (seconds)", grid: true,
        tickValues: [1, 60, 300, 600, 1800, 3600],
        tickFormat: d => d < 60 ? `${d}s` : `${Math.round(d/60)}m`},
    y: {label: "Measured correlation ρ̂(Δ)", domain: [0, 1], grid: true},
  })
  return html`At 5-min sampling: ρ̂ ≈ <b>${rho_5min.toFixed(2)}</b>
    (bias = ${bias_5min}% of true ρ = ${rho_true}) &ensp; — &ensp;
    refresh-time sampling corrects this<br>${plt}`
}
```

# Synchronisation

## Refresh-Time Sampling

Barndorff-Nielsen, Hansen, Lunde & Shephard (2011): **refresh time** sampling is the standard solution to the Epps effect.

**The algorithm:** starting from the latest first-observation time across all assets, the next refresh time is defined as the latest of each asset's next trade time. At each refresh time, all $N$ assets have traded at least once — so no cross-product involves a stale price.

Refresh-time sampling effectively operates at the rate of the *slowest* asset. It loses some observations but completely eliminates the staleness bias. For our 50-stock universe, the effective sampling frequency is typically 3–8 minutes depending on the composition of the portfolio.

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
library(highfrequency)

trades_list <- list(AAPL = trades_aapl, MSFT = trades_msft,
                    NVDA = trades_nvda, AMZN = trades_amzn)

# Synchronise to refresh times and compute realized covariance
sync_data <- refreshTime(trades_list)
rc        <- rCov(sync_data, makeReturns = TRUE)
```

## Python
```{python}
#| eval: false
def refresh_time_sync(price_dict):
    """Barndorff-Nielsen et al. (2011) refresh-time sampling."""
    series = {sym: s.sort_index().dropna() for sym, s in price_dict.items()}
    t_prev = max(s.index[0] for s in series.values())
    refresh_times, prices = [], {sym: [] for sym in series}
    while True:
        nxt = {sym: s[s.index > t_prev] for sym, s in series.items()}
        if any(v.empty for v in nxt.values()):
            break
        t_k = max(v.index[0] for v in nxt.values())
        for sym, s in series.items():
            prices[sym].append(s[s.index <= t_k].iloc[-1])
        refresh_times.append(t_k)
        t_prev = t_k
    df = pd.DataFrame(prices, index=pd.Index(refresh_times))
    return np.log(df).diff().dropna()

sync_returns = refresh_time_sync({sym: trades[sym] for sym in symbols})
RC = sync_returns.T @ sync_returns
```
:::

## Ensuring Positive Definiteness

Raw realized covariance matrices can fail to be **positive definite** (PD) in practice — non-synchronous trading introduces asymmetries and small sample effects push eigenvalues below zero. An indefinite $\hat\Sigma_t$ cannot be inverted, making GMVP weights undefined.

Three approaches, in order of preference:

1. **Nearest PD projection** (Higham 2002): finds the closest PD matrix in Frobenius norm — `Matrix::nearPD()` in R
2. **Eigenvalue clipping:** replace negative eigenvalues with a small positive floor
3. **Shrinkage (preferred):** apply Ledoit-Wolf nonlinear shrinkage — it simultaneously regularises the matrix and guarantees PD. This is preferred because it solves both problems at once: positive definiteness *and* estimation error

```{r}
#| eval: false
# Nonlinear shrinkage of realized covariance (regularises and guarantees PD)
RC_shrunk <- nlshrink_cov(rc_daily_array, k = 1)
```

# Forecasting Realized Covariance

## Element-wise HAR

Apply HAR independently to each of the $N(N+1)/2 = 1{,}275$ unique elements:

$$[\hat\Sigma_{t+1}]_{ij} = \hat\beta_0^{ij} + \hat\beta_d^{ij} [\widehat{RC}_t]_{ij} + \hat\beta_w^{ij} [\overline{RC}_t^{(w)}]_{ij} + \hat\beta_m^{ij} [\overline{RC}_t^{(m)}]_{ij}$$

This is 1,275 separate OLS regressions — fast despite the number, because each is trivial. The DRD decomposition (Chiriac & Voev 2011) is a related approach that separates variances and correlations before applying HAR element-wise; it can improve fit by exploiting the bounded scale of correlations, but adds complexity.

```{r}
#| eval: false
# Element-wise HAR for a 10-stock subset (N=10 → 55 regressions)
har_cov_fits <- list()
for (i in 1:N) {
  for (j in i:N) {
    cov_series <- rc_array[, i, j]
    df_ij <- tibble(
      y   = lead(cov_series),
      x_d = cov_series,
      x_w = slider::slide_dbl(cov_series, mean, .before = 4,  .complete = TRUE),
      x_m = slider::slide_dbl(cov_series, mean, .before = 21, .complete = TRUE)
    ) |> drop_na()
    har_cov_fits[[paste(i, j, sep = "_")]] <- lm(y ~ x_d + x_w + x_m, data = df_ij)
  }
}
```

# The Final Horse Race

## All Methods Against Each Other

The **GMVP as a unified testbed** — no return forecasting required, pure covariance estimation contest.

These numbers are **illustrative, representative of typical findings in the literature** for a 50-stock NASDAQ universe (annualised GMVP volatility):

| # | Method | Ann. GMVP vol |
|---|--------|--------------|
| 1 | Naive 1/N | 16.8% |
| 2 | Rolling SD + sample $\hat C$ | 13.2% |
| 3 | EWMA diagonal + sample $\hat C$ | 12.9% |
| 4 | GARCH diagonal + sample $\hat C$ | 12.4% |
| 5 | DCC-GARCH | 11.8% |
| 6 | Ledoit-Wolf nonlinear shrinkage | 11.5% |
| 7 | DCC + LW shrinkage | 11.1% |
| 8 | HAR-RV diagonal + sample $\hat C$ | 11.9% |
| 9 | Realized covariance (5-min, refresh-time) + LW | 10.8% |
| 10 | Element-wise HAR on realized covariance | 10.5% |

::: {.callout-note icon=false}
**Reading the table:** each row represents one methodological improvement. The gap from row 1 to row 10 — from 16.8% to 10.5% — is the total value of six lectures of volatility modelling, measured in reduced portfolio risk.
:::

## Portfolio Evaluation Code

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
eval_gmvp <- function(Sigma_array, returns, label) {
  port_rets <- numeric(nrow(returns))
  for (t in seq_len(nrow(returns) - 1)) {
    Sigma_t <- as.matrix(nearPD(Sigma_array[t,,])$mat)
    w <- solve(Sigma_t, rep(1, ncol(Sigma_t)))
    w <- w / sum(w)
    port_rets[t + 1] <- sum(w * returns[t + 1, ])
  }
  tibble(method  = label,
         ann_vol = sd(port_rets, na.rm = TRUE) * sqrt(252) * 100,
         sharpe  = mean(port_rets, na.rm = TRUE) / sd(port_rets, na.rm = TRUE) * sqrt(252))
}

results_all <- bind_rows(lapply(names(sigma_list), function(m) {
  eval_gmvp(sigma_list[[m]], daily_returns, label = m)
})) |> arrange(ann_vol)
```

## Python
```{python}
#| eval: false
def eval_gmvp(sigma_list_entry, returns, label):
    port_rets = np.zeros(len(returns))
    for t in range(len(returns) - 1):
        eigs, vecs = np.linalg.eigh(sigma_list_entry[t])
        Sigma_t = vecs @ np.diag(np.maximum(eigs, 1e-8)) @ vecs.T
        w = np.linalg.solve(Sigma_t, np.ones(returns.shape[1]))
        w /= w.sum()
        port_rets[t + 1] = w @ returns[t + 1]
    return {"method": label,
            "ann_vol": np.std(port_rets[1:]) * np.sqrt(252) * 100}
```
:::

## Synthesis: What Have We Learned?

This six-lecture block has built up a complete toolkit for volatility estimation and portfolio construction. Five key findings:

**1. Data richness matters most.** Intraday realized measures outperform daily returns for variance estimation. The jump from GARCH to realized variance (rows 4→8 in the table) comes purely from measurement precision.

**2. Forecasting is HAR's domain.** HAR beats GARCH at daily and weekly horizons because $RV_t$ is a direct observation of $\sigma_t^2$, not a squared daily return. The weekly and monthly averaging structure captures the slow autocorrelation decay that GARCH's exponential decay misses.

**3. Shrinkage is not optional at $N = 50$.** The Marčenko-Pastur interactive showed that even modest concentration ratios inflate $\hat\Sigma^{-1}$ dramatically. Nonlinear LW shrinkage alone (row 6) beats DCC-GARCH (row 5) — controlling estimation error in the correlation matrix matters as much as modelling its dynamics.

**4. DCC and shrinkage are complementary.** Combining them (row 7) beats both alone. The practical recommendation: use GJR-GARCH for marginals, DCC for dynamics, nonlinear LW shrinkage for regularisation.

**5. Realized covariance is the frontier.** With proper synchronisation (refresh-time sampling) and regularisation (LW shrinkage), realized covariance matrices deliver the best out-of-sample GMVP performance (rows 9–10). The remaining gap from DCC+LW to realized covariance+HAR (11.1% → 10.5%) represents the value of intraday cross-product information.

> *Every methodological choice — GARCH vs HAR, shrinkage intensity, sampling frequency — should be evaluated on its ultimate contribution to out-of-sample portfolio performance. The GMVP is the clean, return-agnostic testbed that makes this evaluation honest.*

# Exercises

## Exercise Set 6

**Realized covariance construction**

1. For a 10-stock subset, construct daily realized covariance matrices using naive 5-minute grid synchronisation and refresh-time synchronisation. Compare the off-diagonal elements. Which method produces fewer negative eigenvalues?

2. Demonstrate the Epps effect for one liquid pair (AAPL-MSFT) and one less liquid pair: plot realized correlation as a function of sampling frequency from 1 second to 30 minutes. At what frequency does each curve stabilise?

**Positive definiteness and shrinkage**

3. Apply Ledoit-Wolf nonlinear shrinkage to your realized covariance matrices. Does this solve the PD problem? Compare the eigenvalue distributions before and after. How stable are the portfolio weights?

**The full horse race**

4. Bring together all methods from Lectures 1–6. Evaluate GMVP performance using annualised out-of-sample portfolio volatility and daily turnover. Do the illustrative numbers in the lecture table match your actual findings? Where does your cross-section differ from the typical pattern, and why?

**Key references:** Barndorff-Nielsen, Hansen, Lunde & Shephard (2011); Chiriac & Voev (2011); De Nard, Ledoit & Wolf (2021)
