---
subtitle: "Volatility I: Stylized Facts & First Estimates"
---

```{r}
#| echo: false
#| warning: false
#| eval: false
library(tidyverse)
library(scales)
```

## The Central Question

The Markowitz investor sets $\mu_t = 0$ and minimises portfolio variance — the **Global Minimum Variance Portfolio**:

$$w^* \propto \Sigma_t^{-1} \mathbf{1}$$

The GMVP is our testbed throughout — every modelling choice evaluated by one question: does it reduce out-of-sample portfolio variance?

**The block roadmap:**

| Lecture | What we estimate |
|---------|-----------------|
| 1 | Individual $\sigma_{i,t}$ — naive rolling SD and EWMA |
| 2 | $\sigma_{i,t}$ from GARCH family (daily returns) |
| 3 | $\sigma_{i,t}$ from intraday realized measures |
| 4 | $\sigma_{i,t+h}$ — HAR forecasts at multiple horizons |
| 5 | Full $\Sigma_t$ — DCC-GARCH and shrinkage |
| 6 | Full $\Sigma_t$ — realized covariance from intraday data |

## Running Example

We work with the **same 50 NASDAQ stocks throughout all six lectures** (2010–2024)

| Data | Source | Used in |
|------|--------|---------|
| Daily close prices | CRSP via WRDS | Lectures 1–2, 5 |
| Tick-by-tick trades & quotes | LOBSTER (NASDAQ) | Lectures 3–4, 6 |
| Pre-computed realized measures | VOLARE database | Lectures 4, 6 |

$N = 50$ means 1,275 unique covariance elements — large enough to make dimensionality a real challenge, small enough to remain computationally manageable

**Sector diversity:** tech, healthcare, consumer discretionary — interesting cross-sectional heterogeneity in all stylized facts

## The Core Problem: Volatility is Latent

At time $t$ we observe the return but not the variance that generated it:

$$r_t \sim \mathcal{D}(0,\, \sigma_t^2) \quad \text{— } \sigma_t^2 \text{ is unknown}$$

**Why sample variance fails:** $\hat\sigma^2 = \frac{1}{T-1}\sum r_t^2$ assumes $\sigma_t^2$ is constant — returns are not i.i.d. For portfolio rebalancing we need a **forecast**: $E(\sigma_{t+1}^2 \mid \mathcal{F}_t)$, not a historical average

**Three things the data make unambiguously clear:**

1. Volatility is **not constant** — periods of calm and crisis are visibly different
2. Volatility is **predictable from its own history** — today's level tells us something about tomorrow's
3. Volatility **mean-reverts** — elevated periods eventually end and return to a long-run level

These are the stylized facts. Models that ignore them leave accuracy on the table.

# Stylized Facts

## Four Facts, One Stock

```{r}
#| echo: false
#| warning: false
#| cache: true
library(tidyverse)
library(tidyfinance)

prices_sf <- download_data(
  type = "stock_prices", symbols = "AAPL",
  start_date = "2015-01-01", end_date = "2024-12-31"
) |>
  arrange(date) |>
  mutate(ret = log(adjusted_close / lag(adjusted_close))) |>
  drop_na()

ci_sf <- 1.96 / sqrt(nrow(prices_sf))

acf_sq <- with(acf(prices_sf$ret^2, lag.max = 60, plot = FALSE),
               list(lag = as.numeric(lag[-1]), acf = as.numeric(acf[-1])))

cc_obj <- ccf(prices_sf$ret, prices_sf$ret^2, lag.max = 10, plot = FALSE)
cc_lev <- list(lag = as.numeric(cc_obj$lag[as.numeric(cc_obj$lag) >= 0]),
               ccf = as.numeric(cc_obj$acf[as.numeric(cc_obj$lag) >= 0]))

ojs_define(
  sf_returns = prices_sf |> mutate(date = as.character(date)) |>
                 select(date, ret) |> as.list(),
  sf_acf     = map2(acf_sq$lag, acf_sq$acf, ~ list(lag = .x, acf = .y)),
  sf_ccf     = map2(cc_lev$lag, cc_lev$ccf, ~ list(lag = .x, ccf = .y)),
  sf_ci      = ci_sf
)
```

```{ojs}
//| echo: false
ret_data = sf_returns.date.map((d, i) => ({ date: new Date(d), ret: sf_returns.ret[i] * 100 }))

p_ret = Plot.plot({
  width: 340, height: 200, marginLeft: 42,
  marks: [Plot.lineY(ret_data, {x: "date", y: "ret", stroke: "#3d3d3d", strokeWidth: 0.5})],
  x: {label: null}, y: {label: "Return (%)"},
  title: "(a) Daily log returns — AAPL 2015–2024"
})

p_qq = Plot.plot({
  width: 340, height: 200, marginLeft: 42,
  marks: [
    Plot.qq(ret_data, {x: "ret", stroke: "#901A1E", strokeOpacity: 0.4, strokeWidth: 1.5}),
  ],
  title: "(b) QQ-plot vs Gaussian (tails outside line = fat tails)"
})

p_acf = Plot.plot({
  width: 340, height: 200, marginLeft: 42,
  marks: [
    Plot.barY(sf_acf, {x: "lag", y: "acf", fill: "#901A1E", fillOpacity: 0.75}),
    Plot.ruleY([sf_ci, -sf_ci], {stroke: "#666666", strokeDasharray: "3 2"})
  ],
  x: {label: "Lag (days)"}, y: {label: "ACF"},
  title: "(c) ACF of r²ₜ — slow decay = clustering + long memory"
})

p_lev = Plot.plot({
  width: 340, height: 200, marginLeft: 42,
  marks: [
    Plot.barY(sf_ccf, {x: "lag", y: "ccf", fill: "#901A1E", fillOpacity: 0.75}),
    Plot.ruleY([0, sf_ci, -sf_ci], {stroke: ["#3d3d3d","#666666","#666666"],
               strokeDasharray: ["0","3 2","3 2"]})
  ],
  x: {label: "Lag k"}, y: {label: "CCF"},
  title: "(d) Corr(rₜ, r²ₜ₊ₖ) — negative at k>0 = leverage effect"
})

html`<div style="display:grid;grid-template-columns:1fr 1fr;gap:6px">${p_ret}${p_qq}${p_acf}${p_lev}</div>`
```

**What to look for:** (a) calm and turbulent stretches — *clustering*. (b) points outside the diagonal at both ends — *fat tails*. (c) positive bars that decay slowly — *volatility autocorrelated for months*. (d) negative bars at positive lags — *bad news raises future variance more than good news*.

## Fact 1: Fat Tails

Daily returns have **excess kurtosis** — extreme moves are far more likely than a Gaussian model predicts:

$$\text{kurt}(r_t) = \frac{E[(r_t - \mu)^4]}{\sigma^4} \gg 3$$

| Threshold | Gaussian predicts | Reality |
|-----------|------------------|---------|
| $\|r_t\| > 3\sigma$ | 0.3% of days | ~1–2% of days |
| $\|r_t\| > 5\sigma$ | One in 3.5 million days | Several per year |

**Portfolio implication:** Gaussian VaR systematically underestimates tail risk; a covariance matrix calibrated under normality misprices exactly the scenarios that matter most — the large crashes, not the quiet days.

## Fact 2: Volatility Clustering

**Mandelbrot (1963):** *"large changes tend to be followed by large changes, of either sign, and small changes tend to be followed by small changes"*

$$\text{Corr}(r_t^2,\, r_{t-k}^2) > 0 \quad \text{for } k = 1, 2, \ldots$$

The ACF of $r_t^2$ is significantly positive at many lags and decays slowly — hyperbolically rather than exponentially.

**Portfolio implication:** volatility is predictable from its own history. A dynamic $\hat\sigma_{i,t}$ that responds to yesterday's squared return will outperform a fixed $\hat\sigma$. Position sizing should respond to the current volatility regime.

## Fact 3: Mean Reversion

Volatility clusters, but it does not stay elevated forever:

$$E(\sigma_{t+h}^2 \mid \mathcal{F}_t) \to \bar\sigma^2 \quad \text{as } h \to \infty$$

- Empirical half-life of a volatility shock: roughly **10–30 trading days** for most equities
- The VIX illustrates this clearly — it spikes in crises but reliably reverts toward ~20% annualised

**Portfolio implication:** risk forecasts should converge to a long-run level at long horizons. A model that does not mean-revert — IGARCH, EWMA — maintains perpetually elevated forecasts after a crisis, generating defensive positioning that never unwinds. This is costly for a long-horizon investor.

## Fact 4: The Leverage Effect

Negative returns raise future volatility more than positive returns of equal magnitude:

$$\text{Corr}(r_t,\, \sigma_{t+1}^2) < 0$$

:::: {.columns}
::: {.column width="50%"}
**Financial leverage channel**

A price drop increases the debt-to-equity ratio — the firm becomes riskier — volatility rises
:::
::: {.column width="50%"}
**Volatility feedback channel**

Higher expected volatility raises the required return — the price must fall to deliver it
:::
::::

**Portfolio implication:** drawdowns and rising risk arrive simultaneously. Symmetric volatility models miss this — they treat a -3% day and a +3% day identically. GJR-GARCH (Lecture 2) captures the asymmetry directly.

## Fact 5: Co-movement

Volatilities of different assets move together, especially in crises. A single common factor explains **40–60% of individual stock volatility variance**.

The decomposition that structures the entire lecture block:

$$\Sigma_t = D_t C_t D_t$$

where $D_t = \text{diag}(\sigma_{1,t}, \ldots, \sigma_{N,t})$ and $C_t$ is the conditional correlation matrix

Lectures 1–4 model the diagonal $D_t$ — the individual volatilities. Lectures 5–6 model the full matrix, including time-varying $C_t$.

**Portfolio implication:** a model that captures only $D_t$ with a constant $\hat C$ will mis-weight assets when correlations spike in a crisis — as they always do.

# Naive Estimators

## Rolling Window Standard Deviation

The simplest time-varying volatility estimator:

$$\hat\sigma_{t}^2 = \frac{1}{M-1} \sum_{s=t-M+1}^{t} (r_s - \bar r_{t,M})^2$$

**Bias-variance tradeoff:**

| Window | Behaviour | Problem |
|--------|-----------|---------|
| Short $M$ | Reacts quickly to regime changes | Noisy — high variance |
| Long $M$ | Smooth, stable estimates | Stale in fast-moving markets |

The **cliff-edge problem:** when a large return exits the window, $\hat\sigma_t^2$ jumps even though no new information arrived today.

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
library(tidyverse)

# crsp_returns: tibble with columns date, symbol, ret
rolling_vol <- crsp_returns |>
  group_by(symbol) |>
  arrange(date) |>
  mutate(
    vol_21  = slider::slide_dbl(ret, sd, .before = 20, .complete = TRUE),
    vol_63  = slider::slide_dbl(ret, sd, .before = 62, .complete = TRUE),
    vol_126 = slider::slide_dbl(ret, sd, .before = 125, .complete = TRUE)
  ) |>
  ungroup()
```

## Python
```{python}
#| eval: false
import pandas as pd

# crsp_returns: DataFrame with columns date, symbol, ret
rolling_vol = (crsp_returns
  .sort_values(["symbol", "date"])
  .groupby("symbol")["ret"]
  .apply(lambda x: x.rolling(window=21, min_periods=21).std())
  .reset_index(name="vol_21")
)
```
:::

## Exponentially Weighted Moving Average (EWMA)

The **RiskMetrics** estimator (J.P. Morgan 1994):

$$\hat\sigma_t^2 = (1 - \lambda)\, r_{t-1}^2 + \lambda\, \hat\sigma_{t-1}^2$$

- $\lambda = 0.94$: RiskMetrics default for daily data
- Half-life: $\log(0.5) / \log(0.94) \approx \mathbf{11}$ **trading days**
- No cliff-edge: all past returns included with exponentially declining weights

**Important:** setting $\alpha + \beta = 1$ (as EWMA does) means no mean reversion — volatility forecasts never converge to the long-run level. Compare to GARCH in Lecture 2, which pins $\omega > 0$ to enforce convergence.

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
ewma_vol <- function(ret, lambda = 0.94) {
  n <- length(ret)
  sigma2 <- numeric(n)
  sigma2[1] <- ret[1]^2
  for (t in 2:n) {
    sigma2[t] <- (1 - lambda) * ret[t-1]^2 + lambda * sigma2[t-1]
  }
  sqrt(sigma2)
}

ewma_estimates <- crsp_returns |>
  group_by(symbol) |>
  arrange(date) |>
  mutate(ewma_94 = ewma_vol(ret, lambda = 0.94),
         ewma_97 = ewma_vol(ret, lambda = 0.97)) |>
  ungroup()
```

## Python
```{python}
#| eval: false
def ewma_vol(ret, lam=0.94):
    sigma2 = [ret.iloc[0]**2]
    for r in ret.iloc[1:]:
        sigma2.append((1 - lam) * r**2 + lam * sigma2[-1])
    return pd.Series(sigma2, index=ret.index) ** 0.5

ewma_estimates = (crsp_returns
  .sort_values(["symbol", "date"])
  .groupby("symbol")
  .apply(lambda df: df.assign(ewma_94=ewma_vol(df["ret"], 0.94)))
  .reset_index(drop=True)
)
```
:::

## Interactive: Rolling SD vs EWMA

*Try these experiments:*

**(1)** Set M = 21 and λ = 0.94. Find the largest volatility spike — which estimator reacts first?

**(2)** Set M = 5 and M = 126 in turn. What is the tradeoff you see?

**(3)** Slowly increase λ from 0.85 to 0.99. At what λ does EWMA behave like a long rolling window?

```{ojs}
//| echo: false
viewof M_roll = Inputs.range([5, 126], {step: 1, value: 21,
  label: "Rolling window  M (days)"})
viewof lam_ewma = Inputs.range([0.85, 0.99], {step: 0.01, value: 0.94,
  label: "EWMA decay  λ"})
```

```{ojs}
//| echo: false
// Simulate a GJR-GARCH(1,1) return series (fixed seed via deterministic init)
garch_path = {
  const n = 750, omega = 4e-5, a = 0.06, g = 0.10, b = 0.89
  let s2 = omega / (1 - a - g/2 - b), r = 0
  const out = []
  // Use a simple LCG for reproducibility
  let seed = 42
  const lcg = () => { seed = (seed * 1664525 + 1013904223) & 0xffffffff
    return (seed >>> 0) / 4294967296 }
  const randn = () => Math.sqrt(-2*Math.log(lcg()+1e-10)) * Math.cos(2*Math.PI*lcg())
  for (let t = 0; t < n; t++) {
    const z = randn()
    r = Math.sqrt(s2) * z
    out.push({ t, r, true_vol: Math.sqrt(s2 * 252) * 100 })
    s2 = omega + (a + (r < 0 ? g : 0)) * r*r + b * s2
  }
  return out
}
```

```{ojs}
//| echo: false
vol_paths = {
  const rets = garch_path.map(d => d.r)
  const n = rets.length
  const out = garch_path.map(d => ({...d, roll: null, ewma: null}))
  // Rolling SD
  for (let t = M_roll - 1; t < n; t++) {
    const w = rets.slice(t - M_roll + 1, t + 1)
    const mu = w.reduce((a,x) => a+x, 0) / M_roll
    const v  = w.reduce((a,x) => a+(x-mu)**2, 0) / (M_roll-1)
    out[t].roll = Math.sqrt(v * 252) * 100
  }
  // EWMA
  let ev = rets[0]**2
  for (let t = 0; t < n; t++) {
    if (t > 0) ev = (1 - lam_ewma) * rets[t-1]**2 + lam_ewma * ev
    out[t].ewma = Math.sqrt(ev * 252) * 100
  }
  return out
}
```

```{ojs}
//| echo: false
Plot.plot({
  width: 700, height: 270, marginLeft: 50,
  marks: [
    Plot.lineY(vol_paths, {x: "t", y: "true_vol",
      stroke: "#cccccc", strokeWidth: 1.2}),
    Plot.lineY(vol_paths.filter(d => d.roll !== null), {x: "t", y: "roll",
      stroke: "#901A1E", strokeWidth: 2}),
    Plot.lineY(vol_paths, {x: "t", y: "ewma",
      stroke: "#3d3d3d", strokeWidth: 2, strokeDasharray: "5 3"}),
  ],
  y: {label: "Annualised vol (%)", grid: true},
  x: {label: "Trading day"},
  caption: html`<span style='color:#cccccc'>——</span> True (GJR-GARCH) &ensp;
    <span style='color:#901A1E'>——</span> Rolling SD (M days) &ensp;
    <span style='color:#3d3d3d'>– –</span> EWMA (λ)`
})
```

## What Naive Estimators Give Us

Plugging $\hat\sigma_{i,t}$ into the GMVP:

$$\hat\Sigma_t = \hat D_t \hat C \hat D_t$$

where $\hat C$ is the **sample correlation matrix** (constant) and $\hat D_t = \text{diag}(\hat\sigma_{1,t}, \ldots, \hat\sigma_{N,t})$

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# Compute GMVP weights from diagonal vol + sample correlation
compute_gmvp <- function(sigma_vec, corr_mat) {
  D <- diag(sigma_vec)
  Sigma <- D %*% corr_mat %*% D
  Sigma_inv <- solve(Sigma)
  ones <- rep(1, nrow(Sigma))
  w <- Sigma_inv %*% ones
  w / sum(w)
}

sample_corr <- cor(returns_wide)  # N x N sample correlation matrix
gmvp_weights <- rolling_vol |>
  group_by(date) |>
  summarise(
    weights = list(compute_gmvp(vol_21, sample_corr))
  )
```

## Python
```{python}
#| eval: false
import numpy as np

def compute_gmvp(sigma_vec, corr_mat):
    D = np.diag(sigma_vec)
    Sigma = D @ corr_mat @ D
    Sigma_inv = np.linalg.inv(Sigma)
    ones = np.ones(len(sigma_vec))
    w = Sigma_inv @ ones
    return w / w.sum()

sample_corr = returns_wide.corr().values
```
:::

**Honest assessment:**

- Both rolling SD and EWMA capture clustering qualitatively
- EWMA is more responsive; rolling SD has the cliff-edge discontinuity
- $\hat C$ is constant — but correlations spike in crises (Fact 5), so this is a real limitation
- Better variance models (Lectures 2–4) improve $\hat D_t$; time-varying correlations (Lectures 5–6) improve $\hat C_t$

# Exercises

## Exercise Set 1

**Stylized facts**

1. For a subset of 10 stocks (2 tech, 2 healthcare, 2 consumer discretionary, 4 of your choice): plot QQ-plots against the Gaussian and Student-$t(5)$ distributions. Which fits better? Plot the ACF of $r_t^2$ up to lag 100 — do you observe slow hyperbolic decay?

2. Compute and plot the cross-correlations $\text{Corr}(r_t, r_{t+k}^2)$ for $k = -10, \ldots, 10$ for the same 10 stocks. Is the leverage effect (negative CCF at $k > 0$) consistently present across sectors?

3. Rank the 10 stocks by excess kurtosis. Does the ranking align with your prior about which sectors have fatter tails?

**Naive estimators**

4. Compute rolling standard deviations at $M \in \{21, 63, 126\}$ for all 50 stocks. Find a specific date where the rolling SD jumps despite no unusual return today (the cliff-edge problem). What happened $M$ days earlier?

5. Compute EWMA with $\lambda \in \{0.90, 0.94, 0.97\}$. Compare the implied half-life for each. How does the choice of $\lambda$ affect weekly portfolio turnover? (Hint: compute $\hat w_t^{\text{GMVP}}$ for each $\lambda$ and count daily rebalancing trades.)

**Portfolio evaluation**

6. Construct the GMVP using rolling 21-day SD, rolling 63-day SD, and EWMA(0.94) combined with the sample correlation matrix. Compute annualised out-of-sample portfolio volatility for each. Which estimator gives the lowest realised variance?

**Key references:** Cont (2001); Pagan & Schwert (1990)
