---
subtitle: "Volatility III: High-Frequency Data & Realized Variance"
---

```{r}
#| echo: false
#| warning: false
#| eval: false
library(tidyverse)
library(highfrequency)
```

## Where We Are

**Lectures 1–2** estimated $\sigma_{i,t}^2$ from daily returns. The problem: a daily return is a single draw from $\mathcal{N}(0, \sigma_t^2)$ — the signal-to-noise ratio of $r_t^2$ as a proxy for $\sigma_t^2$ is about 0.05.

**The key insight:** within a single trading day we observe *hundreds* of returns. As the sampling interval $\Delta \to 0$:

$$RV_t(\Delta) = \sum_{j=1}^{1/\Delta} r_{t,j}^2 \xrightarrow{p} \int_0^1 \sigma_t^2(s)\, ds =: IV_t$$

The **integrated variance** $IV_t$ is the true daily variance — $RV_t$ is a direct, low-noise estimate of it. With $m = 78$ five-minute intervals: the variance of $RV_t$ is roughly **78× lower** than the variance of $r_t^2$.

::: {.callout-note icon=false}
**Portfolio implication:** better estimates of $\sigma_{i,t}^2$ feed directly into better estimates of $\Sigma_t$ and thus better GMVP weights $w^* = \Sigma_t^{-1} \mathbf{1} / \mathbf{1}' \Sigma_t^{-1} \mathbf{1}$.
:::

## Data Infrastructure

We use two complementary data sources:

**LOBSTER** (Limit Order Book System — The Efficient Reconstructor) provides reconstructed limit order books for all NASDAQ-listed stocks, available from 2007 onwards. Each stock-day comes in two files: a message file (all events) and an order book file (depth snapshots). A single active stock generates 1–5 GB of raw data per year.

**VOLARE** (VOLatility Archive for Realized Estimates) provides pre-computed daily realized measures — RV, BPV, realized kernels, semi-variances — for our 50 NASDAQ stocks. This is the practical shortcut: use LOBSTER when you need to understand the raw data structure or compute non-standard measures; use VOLARE for estimation across the full 50-stock cross-section.

## The Limit Order Book

```{r}
#| label: fig-orderbook
#| echo: false
#| warning: false
#| fig-cap: "Simulated LOB snapshot around mid-price $150. The bid-ask spread sits at the centre; alternating executions at bid and ask create the **bid-ask bounce** — a key source of microstructure noise."
#| fig-width: 9
#| fig-height: 4.5

library(tidyverse)

mid <- 150.00; tick <- 0.01
set.seed(42)

lob <- bind_rows(
  tibble(side = "Bid (buy)",  level = 1:10,
         price = mid - tick * (1:10),
         qty   = c(1200, 850, 2300, 540, 1800, 760, 430, 1100, 920, 680)),
  tibble(side = "Ask (sell)", level = 1:10,
         price = mid + tick * (1:10),
         qty   = c(950, 1400, 620, 1750, 490, 2100, 870, 350, 1300, 760))
)

ggplot(lob, aes(y = sprintf("%.2f", price),
                x = ifelse(side == "Bid (buy)", -qty, qty),
                fill = side)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_vline(xintercept = 0, linewidth = 0.9, colour = "#3d3d3d") +
  annotate("text", x = 0, y = 10.6, label = "← Spread →",
           hjust = 0.5, size = 3.2, colour = "#666666") +
  scale_fill_manual(values = c("Ask (sell)" = "#901A1E",
                                "Bid (buy)"  = "#3d3d3d")) +
  scale_x_continuous(
    labels = function(x) scales::comma(abs(x)),
    breaks = c(-2000, -1000, 0, 1000, 2000)
  ) +
  labs(x = "Quantity (shares)", y = "Price ($)", fill = NULL,
       title = "NASDAQ limit order book snapshot (10 levels each side)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top",
        panel.grid.major.y = element_blank())
```

# Microstructure Noise

## Sources of Noise

Raw transaction prices deviate from the "efficient" price $p_t^*$. The key sources are:

- **Bid-ask bounce:** transactions alternate between bid and ask, creating spurious oscillation of order $\frac{1}{2}(ask - bid)$
- **Price discreteness:** US stocks trade in 1-cent ticks — at high frequency, the rounding error is not negligible relative to the true price change
- **Stale quotes:** the best quotes do not update instantaneously; mid-quote lags the efficient price

**Model:** $p_{t,j} = p_{t,j}^* + u_{t,j}$ where $u_{t,j}$ is noise with $E(u) = 0$, $\text{Var}(u) = \xi^2$. At high frequency, the noise term $\Delta u_{t,j}$ dominates the true price signal.

## The Signature Plot: Your First Diagnostic

The **signature plot** shows how $RV_t(\Delta)$ varies with the sampling interval $\Delta$:

$$E[RV_t(\Delta)] = IV_t + \frac{2\xi^2}{\Delta}$$

As $\Delta \to 0$ (sampling more finely), the noise bias $2\xi^2/\Delta \to \infty$. As $\Delta \to \infty$, you converge to the single daily return — just one observation. The practitioner's question: *where does the curve flatten?* That is the sweet spot.

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
frequencies <- c(1, 5, 10, 30, 60, 300, 600, 900, 1800)  # seconds

rv_by_freq <- sapply(frequencies, function(freq) {
  r_sampled <- aggregateReturns(intraday_returns, on = "seconds", k = freq)
  RV(r_sampled)
})

tibble(seconds = frequencies, rv = rv_by_freq) |>
  ggplot(aes(x = seconds, y = sqrt(rv) * sqrt(252))) +
  geom_line(color = "#901A1E") + geom_point() +
  scale_x_log10() +
  labs(x = "Sampling interval (seconds, log scale)",
       y = "Annualised RV (sqrt)", title = "Signature plot — AAPL") +
  theme_minimal()
```

## Python
```{python}
#| eval: false
frequencies = [1, 5, 10, 30, 60, 300, 600, 900, 1800]

rv_by_freq = []
for freq in frequencies:
    r_s = intraday_prices.resample(f"{freq}s").last().pct_change().dropna()
    rv_by_freq.append((r_s**2).sum())

plt.semilogx(frequencies, np.sqrt(rv_by_freq) * np.sqrt(252))
plt.xlabel("Sampling interval (seconds)")
plt.ylabel("Annualised volatility")
```
:::

## Interactive: Microstructure Noise & the Signature Plot

*Guided task: Set ξ=0.004. At 1-second sampling, read off the estimated annualised volatility — it is massively inflated by noise. Now look at the noise percentage shown. Where does the curve start to flatten — that is your practical sweet spot. Drag ξ back to 0. What does the completely flat line mean? It means with zero noise, you should sample as finely as possible. This is why the 5-minute convention exists: it is the approximate sweet spot for large-cap US stocks where ξ is small but nonzero.*

```{ojs}
//| echo: false
viewof xi_noise = Inputs.range([0.000, 0.006], {step: 0.0001, value: 0.002,
  label: "Noise level  ξ  (std dev of bid-ask noise)"})
viewof iv_daily = Inputs.range([0.005, 0.060], {step: 0.001, value: 0.020,
  label: "True integrated variance IV  (daily)"})
```

```{ojs}
//| echo: false
sig_data = {
  // Sampling intervals in seconds; 23400 = 6.5h trading day
  const deltas = [1, 2, 5, 10, 30, 60, 120, 300, 600, 900, 1800, 3600]
  const T_day  = 23400
  return deltas.map(delta => {
    const n     = Math.floor(T_day / delta)
    const rv    = iv_daily + 2 * xi_noise**2 * n   // E[RV(Δ)] = IV + 2ξ²/Δ
    const noise_pct = (2 * xi_noise**2 * n) / rv * 100
    return {
      delta,
      annvol:     Math.sqrt(rv * 252) * 100,
      true_vol:   Math.sqrt(iv_daily * 252) * 100,
      noise_pct,
      label: delta < 60 ? `${delta}s` : delta < 3600 ? `${delta/60}m` : "1h"
    }
  })
}
```

```{ojs}
//| echo: false
Plot.plot({
  width: 700, height: 275, marginLeft: 55,
  marks: [
    Plot.ruleY([sig_data[0].true_vol],
      {stroke: "#cccccc", strokeDasharray: "4 2", strokeWidth: 1.5}),
    Plot.ruleX([300], {stroke: "#666666", strokeDasharray: "2 2", strokeWidth: 1}),
    Plot.text([{delta: 420, annvol: sig_data[0].true_vol * 0.60, label: "5 min"}],
      {x: "delta", y: "annvol", text: "label", fontSize: 10, fill: "#666666"}),
    Plot.lineY(sig_data, {x: "delta", y: "annvol",
      stroke: "#901A1E", strokeWidth: 2.5}),
    Plot.dot(sig_data, {x: "delta", y: "annvol",
      fill: "#901A1E", r: 4}),
    Plot.text(sig_data, {x: "delta", y: "annvol", text: "label",
      dy: -12, fontSize: 9, fill: "#3d3d3d"}),
  ],
  x: {type: "log", label: "Sampling interval Δ (seconds, log scale)",
      grid: true, tickValues: [1, 5, 30, 60, 300, 1800, 3600]},
  y: {label: "RV estimate — annualised vol (%)", grid: true},
  caption: html`Noise bias = 2ξ²/Δ → ∞ as Δ → 0 &ensp;|&ensp;
    <span style='color:#cccccc'>– –</span> True (noise-free) volatility`
})
```

## Five-Minute RV: The Practitioner Standard

The simplest noise-robust solution: sample at **5-minute intervals**:

$$\widehat{RV}_t^{(5\text{min})} = \sum_{j=1}^{78} r_{t,j,5\text{min}}^2$$

78 five-minute intervals per 6.5-hour trading day. This is the standard benchmark used throughout academic and practitioner literature. It largely eliminates bid-ask bounce, is simple to implement, and is what the VOLARE database pre-computes. More sophisticated estimators exist (realized kernels correct for noise explicitly at all frequencies), but 5-minute sparse sampling remains the workhorse.

# Jump Detection

## Bipower Variation

A key question: is today's high realized variance driven by continuous diffusive risk or by a jump (earnings surprise, macro release)?

**Intuition:** a price jump affects *one* return only. The product of two *adjacent* returns contains at most one jump; averaging these products dilutes the jump's contribution. Squared returns amplify a jump by the full jump size squared; adjacent products are diluted.

$$BPV_t = \frac{\pi}{2} \sum_{j=2}^{n} |r_{t,j}| |r_{t,j-1}|$$

$BPV_t$ converges to the integrated variance $IV_t$ even in the presence of jumps, while $RV_t$ picks up both. So $RV_t - BPV_t \geq 0$ estimates the **jump variation** — the sum of squared jump sizes.

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
rv_val  <- RV(intraday_returns)
bpv_val <- BPV(intraday_returns)

jump_var <- max(rv_val - bpv_val, 0)
cont_var <- bpv_val
```

## Python
```{python}
#| eval: false
def bipower_variation(returns):
    return (np.pi / 2) * np.sum(np.abs(returns[1:]) * np.abs(returns[:-1]))

rv  = (returns**2).sum()
bpv = bipower_variation(returns)
jump_var = max(rv - bpv, 0)
```
:::

**For portfolio management:** continuous variance is persistent and predictable — this is the component we forecast with HAR (Lecture 4). Jump variance is transient and event-driven — options are needed to hedge it; dynamic rebalancing alone is insufficient. GARCH models incorporate yesterday's jump into $\hat\sigma_t^2$ via $r_{t-1}^2$, inflating volatility estimates spuriously.

# From Realized Variance to Portfolio Weights

## Connecting to the GMVP

The realized variance $\widehat{RV}_{i,t}$ is an estimate of today's $\sigma_{i,t}^2$. To use it for portfolio construction we need a *forecast* of tomorrow's variance — this is the job of HAR models (Lecture 4). But we can already see the portfolio payoff from better measurement:

$$\hat\Sigma_t = \hat D_t^{RV} \hat C \hat D_t^{RV}, \quad \hat D_t^{RV} = \text{diag}(\sqrt{\widehat{RV}_{1,t}}, \ldots, \sqrt{\widehat{RV}_{N,t}})$$

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# Load pre-computed RV from VOLARE
rv_today <- volare_data |>
  filter(date == rebalance_date) |>
  select(symbol, rv)

sig_vec  <- sqrt(rv_today$rv)
D_rv     <- diag(sig_vec)
Sigma_rv <- D_rv %*% sample_corr %*% D_rv

w_rv <- solve(Sigma_rv, rep(1, 50))
w_rv <- w_rv / sum(w_rv)
```

## Python
```{python}
#| eval: false
rv_today = volare_data[volare_data["date"] == rebalance_date]["rv"].values
sig_vec  = np.sqrt(rv_today)
D_rv     = np.diag(sig_vec)
Sigma_rv = D_rv @ sample_corr @ D_rv
w_rv     = np.linalg.solve(Sigma_rv, np.ones(50))
w_rv    /= w_rv.sum()
```
:::

::: {.callout-note icon=false}
**Portfolio implication:** using $\widehat{RV}_{i,t}$ instead of $r_{i,t}^2$ as the volatility input gives a 78× reduction in measurement noise on any given day. This translates into more stable portfolio weights and lower realised portfolio variance. Lectures 4–6 refine this further by adding forecasting (HAR), time-varying correlations (DCC), and full realized covariance matrices.
:::

## Bridge to Lecture 4

We now have a precise time series $\{\widehat{RV}_{i,t}\}$ for each of our 50 NASDAQ stocks, available from VOLARE. **Lecture 4** asks: how do we *forecast* $RV_{i,t+1}$ from the history $\{RV_{i,s}\}_{s \leq t}$? The answer is the HAR model — a regression on daily, weekly, and monthly averages of past RV, motivated by the heterogeneous market hypothesis.

# Exercises

## Exercise Set 3

**Signature plots**

1. Construct signature plots for a liquid stock (AAPL) and a less liquid stock from the NASDAQ 50. Compute $RV_t(\Delta)$ for $\Delta \in \{1s, 5s, 30s, 1m, 5m, 15m, 30m\}$ and plot on a log scale. At what frequency does each curve flatten? How does the sweet spot differ between liquid and illiquid?

**Jump detection**

2. Using VOLARE or LOBSTER, compute $RV_t$ and $BPV_t$ for 5 stocks over 60 trading days. Plot the time series of $RV_t$, $BPV_t$, and the jump component $\max(RV_t - BPV_t, 0)$. Identify the 3 largest jump days. What firm-specific news occurred on those dates?

**Portfolio evaluation**

3. Using VOLARE daily $RV_t$ for all 50 stocks, construct the GMVP with $\hat\Sigma_t = \hat D_t^{RV} \hat C \hat D_t^{RV}$. Compare out-of-sample portfolio volatility to GARCH-based $\hat D_t$ (Lecture 2) and rolling SD (Lecture 1). Does the more precise measurement translate into lower realised portfolio variance?

**Key references:** Andersen, Bollerslev, Diebold & Labys (2003); Barndorff-Nielsen & Shephard (2004); Boudt, Kleen & Sjoerup (2022)
