---
subtitle: "Volatility II: GARCH Models in Practice"
---

```{r}
#| echo: false
#| warning: false
#| eval: false
library(tidyverse)
library(tsgarch)
library(future)
library(future.apply)
```

## Where We Are

**Lecture 1 established:** volatility is latent, time-varying, and exhibits clustering, fat tails, leverage, and long memory

**The EWMA shortcoming:** it has no mean reversion, the parameter $\lambda$ is not estimated from data, and it is symmetric — it treats positive and negative shocks equally

**This lecture:** GARCH models provide a principled, likelihood-based framework for estimating and forecasting $\sigma_{i,t}^2$ from daily returns

**Portfolio goal:** replace the naive $\hat\sigma_{i,t}$ in the covariance matrix with a well-specified, data-driven conditional variance forecast

$$\hat\Sigma_t = \hat D_t \hat C_t \hat D_t, \quad \hat D_t = \text{diag}(\hat\sigma_{1,t|t-1}, \ldots, \hat\sigma_{N,t|t-1})$$

The $\hat\sigma_{i,t|t-1}$ will come from GARCH; handling $\hat C_t$ is the job of Lecture 5 (DCC)

# GARCH Models

## GARCH(1,1) — The Workhorse

The **Generalised ARCH** model (Bollerslev 1986):

$$r_t = \sigma_t \varepsilon_t, \quad \varepsilon_t \overset{i.i.d.}{\sim} \mathcal{D}(0, 1)$$

$$\sigma_t^2 = \omega + \alpha r_{t-1}^2 + \beta \sigma_{t-1}^2$$

- $\omega > 0$: intercept (anchors the long-run variance)
- $\alpha \geq 0$: **ARCH effect** — how much yesterday's squared return raises today's variance
- $\beta \geq 0$: **GARCH effect** — how much yesterday's variance persists
- Stationarity: $\alpha + \beta < 1$

**Unconditional variance:**

$$\bar\sigma^2 = \frac{\omega}{1 - \alpha - \beta}$$

**Persistence:** $\alpha + \beta$ measures how quickly a volatility shock decays. For most stocks: $\alpha + \beta \approx 0.97$–$0.99$

**Connection to EWMA:** setting $\omega = 0$ and $\alpha + \beta = 1$ gives exactly the EWMA recursion — so EWMA is a restricted IGARCH(1,1) with no mean reversion

## Asymmetric Extensions

**Why we need them:** the leverage effect (Fact 4, Lecture 1) — negative shocks increase variance more than positive shocks of equal size

### GJR-GARCH (Glosten, Jagannathan, Runkle 1993)

$$\sigma_t^2 = \omega + (\alpha + \gamma \mathbf{1}_{r_{t-1}<0}) r_{t-1}^2 + \beta \sigma_{t-1}^2$$

- Additional parameter $\gamma$: the **asymmetry coefficient**
- Negative shock ($r_{t-1} < 0$): impact is $\alpha + \gamma$ (larger if $\gamma > 0$)
- Positive shock ($r_{t-1} > 0$): impact is $\alpha$
- Stationarity: $\alpha + \frac{\gamma}{2} + \beta < 1$

### EGARCH (Nelson 1991)

$$\log \sigma_t^2 = \omega + \beta \log \sigma_{t-1}^2 + \alpha \left|\frac{r_{t-1}}{\sigma_{t-1}}\right| + \gamma \frac{r_{t-1}}{\sigma_{t-1}}$$

- Operates on $\log\sigma_t^2$ — no non-negativity constraints needed
- Asymmetry via $\gamma < 0$: negative standardised residuals increase variance
- More flexible functional form for the news impact curve

## The News Impact Curve

The **news impact curve** (NIC) shows $\sigma_t^2$ as a function of $r_{t-1}$ holding $\sigma_{t-1}^2 = \bar\sigma^2$:

:::: {.columns}
::: {.column width="50%"}
**GARCH(1,1):**

$$\text{NIC}(r) = \omega + \alpha r^2 + \beta\bar\sigma^2$$

- U-shaped, **symmetric** around zero
- Equal response to positive and negative shocks
:::
::: {.column width="50%"}
**GJR-GARCH(1,1):**

$$\text{NIC}(r) = \begin{cases} \omega + (\alpha + \gamma) r^2 + \beta\bar\sigma^2 & r < 0 \\ \omega + \alpha r^2 + \beta\bar\sigma^2 & r \geq 0 \end{cases}$$

- U-shaped, **asymmetric**: steeper left arm
- Negative shocks are more damaging to the portfolio
:::
::::

**Portfolio implication:** using a symmetric GARCH when the true process is asymmetric under-estimates the variance spike following a market crash — exactly when accurate estimates matter most

## Interactive: Build Your Own News Impact Curve

*Increase γ and watch the left arm of the curve steepen — that is the leverage effect in action*

```{ojs}
//| echo: false
viewof a_nic = Inputs.range([0.01, 0.25], {step: 0.005, value: 0.08,
  label: "α  (ARCH — symmetric reaction to news)"})
viewof g_nic = Inputs.range([0.00, 0.35], {step: 0.005, value: 0.14,
  label: "γ  (GJR leverage — extra reaction to bad news)"})
viewof b_nic = Inputs.range([0.55, 0.98], {step: 0.01,  value: 0.88,
  label: "β  (GARCH persistence)"})
```

```{ojs}
//| echo: false
nic_curves = {
  const omega = 1e-6
  const denom = 1 - a_nic - g_nic/2 - b_nic
  const s2bar = denom > 0.001 ? omega / denom : 0.04 / 252
  return Array.from({length: 201}, (_, i) => {
    const r = (i - 100) * 0.001
    const garch_var = omega + a_nic * r**2 + b_nic * s2bar
    const gjr_var   = omega + (a_nic + (r < 0 ? g_nic : 0)) * r**2 + b_nic * s2bar
    return {
      r_pct: r * 100,
      garch: Math.sqrt(Math.max(garch_var, 0) * 252) * 100,
      gjr:   Math.sqrt(Math.max(gjr_var,   0) * 252) * 100,
      lrun:  Math.sqrt(s2bar * 252) * 100
    }
  })
}
```

```{ojs}
//| echo: false
Plot.plot({
  width: 700, height: 280, marginLeft: 55,
  marks: [
    Plot.ruleY([nic_curves[0].lrun],
      {stroke: "#cccccc", strokeDasharray: "4 2", strokeWidth: 1.5}),
    Plot.ruleX([0], {stroke: "#dddddd"}),
    Plot.lineY(nic_curves, {x: "r_pct", y: "garch",
      stroke: "#666666", strokeWidth: 2}),
    Plot.lineY(nic_curves, {x: "r_pct", y: "gjr",
      stroke: "#901A1E", strokeWidth: 2.5}),
  ],
  x: {label: "Yesterday's return r_{t−1} (%)", grid: true},
  y: {label: "σ_t — annualised vol (%)", grid: true},
  caption: html`<span style='color:#666666'>——</span> GARCH (symmetric) &ensp;
    <span style='color:#901A1E'>——</span> GJR-GARCH (asymmetric) &ensp;
    <span style='color:#cccccc'>– –</span> Long-run vol`
})
```

## Why GARCH(1,1) and GJR-GARCH(1,1) Dominate

Hansen & Lunde (2005): *"A Forecast Comparison of Volatility Models"*

- Estimated 330 GARCH-type models on daily IBM returns
- No model significantly outperforms GARCH(1,1) in forecasting variance
- **Exception:** models that capture the leverage effect (GJR, EGARCH) consistently beat symmetric GARCH
- Adding higher lags (GARCH(p,q) with $p,q>1$) does not improve forecasts

**Practical takeaway for our 50-stock universe:**

- Default specification: GJR-GARCH(1,1) — captures leverage, parsimonious
- GARCH(1,1) as benchmark
- Verify with model confidence sets (MCS) — do not over-specify

# Practical Estimation

## Distributional Choice

Three options for $\mathcal{D}(0, 1)$:

| Distribution | Parameters | Tail behavior | Portfolio use case |
|---|---|---|---|
| Gaussian | — | Light tails | Simple; VaR underestimates tail risk |
| Student-$t$ | df $\nu$ | Heavy tails | Better VaR; $\nu \approx 5$–8 for stocks |
| Skewed-$t$ (Hansen 1994) | df $\nu$, skewness $\xi$ | Heavy + asymmetric | Most flexible; recommended for individual stocks |

**QML estimation:** even if the distribution is misspecified (e.g., using Gaussian when returns are Student-$t$), the GARCH parameter estimates $(\omega, \alpha, \beta)$ remain **consistent** — this is quasi-maximum likelihood (QML)

- QML gives consistent $\hat\sigma_t^2$ but incorrect standard errors → use robust sandwich standard errors
- VaR and ES from the conditional distribution require correct distributional specification — QML is not enough here

## Variance Targeting

**The intercept $\omega$ is often poorly identified** — it is a small number and the optimiser can take a long time to converge

**Variance targeting** (Engle & Mezrich 1996): set $\omega$ from the sample variance

$$\hat\omega = \hat{\bar\sigma}^2 (1 - \alpha - \beta)$$

- Reduces the optimisation to two free parameters $(\alpha, \beta)$ for GARCH(1,1)
- Dramatically improves numerical stability and convergence speed
- Slight efficiency loss but negligible in practice
- Default in `tsgarch`: `variance.targeting = TRUE`

**Boundary solutions:** when the optimised $\hat\alpha + \hat\beta \approx 1$

- Indicates near-integrated variance process
- Common for individual stocks over long samples
- Consequence: the implied long-run mean is very large — long-horizon forecasts are unreliable
- Solution: impose stationarity constraint or use FIGARCH

## Estimation with `tsgarch`

`tsgarch` (Galanos) is the modern successor to `rugarch`, built on the `tsmodels` framework with **automatic differentiation via TMB** (Template Model Builder)

- Analytic gradients and Hessians via TMB → numerical stability, accurate standard errors
- S3 class system, consistent with `tsmarch` (Lecture 5)

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
library(tsgarch)

# Specify GJR-GARCH(1,1) with Student-t errors and variance targeting
spec <- garch_modelspec(
  y = returns[, "AAPL"],
  model = "gjrgarch",
  order = c(1, 1),
  constant = TRUE,
  distribution = "sstd",   # skewed Student-t
  variance_targeting = TRUE
)

# Estimate
fit <- estimate(spec)
summary(fit)

# Key output: persistence alpha + beta + gamma/2, unconditional sigma
```

## Python
```{python}
#| eval: false
from arch import arch_model

# GJR-GARCH(1,1) with Student-t innovations
model = arch_model(
    returns["AAPL"],
    vol="Garch",
    p=1, o=1, q=1,      # o=1 enables the leverage term (GJR)
    dist="t",
    mean="Zero"
)
result = model.fit(disp="off")
print(result.summary())
```
:::

## Diagnostics

After estimation, always inspect the standardised residuals $\hat\varepsilon_t = r_t / \hat\sigma_t$:

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# Standardised residuals
e_hat <- residuals(fit, standardize = TRUE)

# 1. ACF of squared standardised residuals — should be near zero
acf(e_hat^2, lag.max = 20)

# 2. Ljung-Box test on e_hat^2 — H0: no autocorrelation
Box.test(e_hat^2, lag = 10, type = "Ljung-Box")

# 3. QQ-plot against specified distribution
plot(fit, which = "residuals_qq")

# 4. News impact curve
plot(fit, which = "news_impact")
```

## Python
```{python}
#| eval: false
from statsmodels.stats.diagnostic import acorr_ljungbox
import matplotlib.pyplot as plt

std_resid = result.std_resid

# Ljung-Box test on squared standardised residuals
lb_test = acorr_ljungbox(std_resid**2, lags=[10], return_df=True)
print(lb_test)

# QQ-plot
import scipy.stats as stats
fig, ax = plt.subplots()
stats.probplot(std_resid, dist="t", sparams=(result.params["nu"],), plot=ax)
plt.show()
```
:::

**Pass criteria:**

- Ljung-Box p-value > 0.05 on $\hat\varepsilon_t^2$ → no remaining ARCH effects
- Near-normal QQ-plot for the standardised residuals under the assumed distribution

# Forecasting

## One-Step and Multi-Step Forecasts

**One-step-ahead:** directly from the recursion

$$\hat\sigma_{t+1|t}^2 = \hat\omega + \hat\alpha r_t^2 + \hat\beta \hat\sigma_t^2$$

**Multi-step-ahead ($h > 1$):** iterate the recursion, using $E(r_{t+k}^2 \mid \mathcal{F}_t) = \hat\sigma_{t+k|t}^2$

$$\hat\sigma_{t+h|t}^2 = \hat{\bar\sigma}^2 + (\hat\alpha + \hat\beta)^{h-1} (\hat\sigma_{t+1|t}^2 - \hat{\bar\sigma}^2)$$

- Forecasts converge to $\hat{\bar\sigma}^2$ at rate $(\alpha + \beta)^h$ — mean reversion!
- For $\alpha + \beta \approx 0.98$: half-life $\approx 34$ days
- EWMA ($\alpha + \beta = 1$): no convergence — forecasts remain at $\hat\sigma_{t+1|t}^2$ forever

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# Rolling one-step-ahead forecasts (re-estimate monthly)
fc <- predict(fit, n.ahead = 10)  # 10-step ahead
plot(fc)

# Extract conditional variance path
sigma2_path <- sigma(fit)^2
```

## Python
```{python}
#| eval: false
forecasts = result.forecast(horizon=10)
# variance forecasts
var_forecasts = forecasts.variance
```
:::

## Interactive: GARCH Mean Reversion vs EWMA

*Move the shock slider up to simulate a market crash. Move persistence toward 1 to see how EWMA behaviour emerges.*

```{ojs}
//| echo: false
viewof pers = Inputs.range([0.86, 0.999], {step: 0.001, value: 0.970,
  label: "Persistence  α + β"})
viewof shock_m = Inputs.range([1.0, 5.0], {step: 0.1, value: 3.0,
  label: "Current vol / long-run vol  (shock size)"})
```

```{ojs}
//| echo: false
conv_data = {
  const sigma_bar = 20  // long-run annualised vol, %
  const sigma_now = sigma_bar * shock_m
  const half_life = Math.log(0.5) / Math.log(pers)
  const hl_day = Math.round(-half_life)
  return Array.from({length: 121}, (_, h) => ({
    h,
    garch: Math.sqrt(sigma_bar**2 + pers**h * (sigma_now**2 - sigma_bar**2)),
    ewma:  sigma_now,
    lrun:  sigma_bar,
    is_hl: h === hl_day
  }))
}
```

```{ojs}
//| echo: false
{
  const hl = conv_data.find(d => d.is_hl)
  const half_life_label = hl ? `½-life ≈ ${hl.h} days` : ""
  return Plot.plot({
    width: 700, height: 280, marginLeft: 55,
    marks: [
      Plot.ruleY([conv_data[0].lrun],
        {stroke: "#cccccc", strokeDasharray: "4 2", strokeWidth: 1.5}),
      Plot.lineY(conv_data, {x: "h", y: "ewma",
        stroke: "#aaaaaa", strokeWidth: 1.8, strokeDasharray: "5 3"}),
      Plot.lineY(conv_data, {x: "h", y: "garch",
        stroke: "#901A1E", strokeWidth: 2.5}),
      hl ? Plot.dot([hl], {x: "h", y: "garch", r: 5, fill: "#901A1E"}) : null,
      hl ? Plot.text([{...hl, label: half_life_label}],
        {x: "h", y: "garch", text: "label", dy: -14, fontSize: 11}) : null,
    ].filter(Boolean),
    x: {label: "Forecast horizon h (trading days)", grid: true},
    y: {label: "E[σ_{t+h} | F_t]  (%)", grid: true},
    caption: html`<span style='color:#901A1E'>——</span> GARCH (mean-reverts to long-run level) &ensp;
      <span style='color:#aaaaaa'>– –</span> EWMA / IGARCH (flat — no mean reversion) &ensp;
      <span style='color:#cccccc'>– –</span> Long-run vol (20%)`
  })
}
```

## Forecast Evaluation

**What do we evaluate against?** $r_t^2$ is a noisy proxy for $\sigma_t^2$ — using it as the target introduces measurement error

**Loss functions robust to proxy noise:**

| Loss | Formula | Notes |
|------|---------|-------|
| MSE | $(r_t^2 - \hat\sigma_t^2)^2$ | Proxy noise biases rankings |
| QLIKE | $r_t^2/\hat\sigma_t^2 - \log(r_t^2/\hat\sigma_t^2) - 1$ | Consistent under proxy noise |

Patton (2011): use a **homogeneous** loss function — QLIKE satisfies this; MSE does not

**Mincer-Zarnowitz regression:**

$$r_t^2 = a + b\hat\sigma_t^2 + u_t$$

- Unbiasedness: $H_0: a = 0, b = 1$
- $R^2$ measures explanatory power (typically 5–20% with daily $r_t^2$ as proxy)

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
library(sandwich)
library(lmtest)

mz_reg <- lm(ret^2 ~ sigma2_hat, data = forecast_df)
coeftest(mz_reg, vcov = NeweyWest(mz_reg, lag = 5))
```

## Python
```{python}
#| eval: false
import statsmodels.formula.api as smf
from statsmodels.stats.sandwich_covariance import cov_hac

mz_model = smf.ols("ret_sq ~ sigma2_hat", data=forecast_df).fit()
print(mz_model.get_robustcov_results(cov_type="HAC", maxlags=5).summary())
```
:::

## Model Confidence Set (MCS)

Hansen, Lunde & Nason (2011): the **Model Confidence Set** provides a set of models that are statistically equivalent given the data

- Starts with all $M$ candidate models
- Iteratively eliminates the model with the worst loss using a bootstrap test
- Returns the set $\mathcal{M}^*_{1-\alpha}$ of models that cannot be distinguished at level $\alpha$

**Why MCS instead of Diebold-Mariano?**

- Diebold-Mariano is pairwise; MCS handles multiple comparisons simultaneously
- Controls FWER: no need to adjust for the number of pairwise tests

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# MCSTest from MCS package (or rugarch/tsgarch wrappers)
library(MCS)

# Collect QLIKE losses for all competing models
qlike_matrix <- cbind(qlike_garch, qlike_gjr, qlike_egarch, qlike_ewma)
colnames(qlike_matrix) <- c("GARCH", "GJR", "EGARCH", "EWMA")

mcs_result <- MCSprocedure(
  Loss = qlike_matrix,
  alpha = 0.10,
  B = 5000,
  statistic = "Tmax"
)
show(mcs_result)
```

## Python
```{python}
#| eval: false
# No direct MCS package in Python — implement manually or use R via rpy2
# Alternatively: use pairwise DM tests with Holm-Bonferroni correction
from arch.tests.utility import diebold_mariano

dm_stat, dm_pval = diebold_mariano(
    e1=losses_garch, e2=losses_gjr, power=1
)
```
:::

# Scaling to 50 Stocks

## Batch Estimation with `tsgarch` + `future`

Estimating GARCH models for 50 stocks individually would be tedious — use `tsgarch`'s **multi-specification** framework with parallel execution:

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
library(tsgarch)
library(future)
library(future.apply)

plan(multisession, workers = 8)  # parallelise across 8 cores

# Build a multi-spec: same model family, one per stock
symbols <- colnames(returns)  # 50 column names

multi_spec <- lapply(symbols, function(sym) {
  garch_modelspec(
    y = returns[, sym],
    model = "gjrgarch",
    order = c(1, 1),
    distribution = "sstd",
    variance_targeting = TRUE
  )
})

# Estimate all 50 in parallel
multi_fit <- future_lapply(multi_spec, estimate)
names(multi_fit) <- symbols

# Extract one-step-ahead conditional variances for each stock
sigma2_matrix <- do.call(cbind, lapply(multi_fit, sigma))^2
# sigma2_matrix: T x 50 matrix of sigma_{i,t}^2
```

## Python
```{python}
#| eval: false
from arch import arch_model
from joblib import Parallel, delayed

def fit_gjr(symbol, returns):
    model = arch_model(returns[symbol], vol="Garch", p=1, o=1, q=1,
                       dist="t", mean="Zero")
    return model.fit(disp="off")

results = Parallel(n_jobs=8)(
    delayed(fit_gjr)(sym, returns) for sym in returns.columns
)
sigma2_matrix = pd.DataFrame(
    {sym: res.conditional_volatility**2
     for sym, res in zip(returns.columns, results)}
)
```
:::

## Cross-Sectional Heterogeneity

Not all 50 stocks need the same specification — inspect the cross-section:

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# Extract key parameters for all 50 stocks
params_df <- bind_rows(lapply(symbols, function(sym) {
  p <- coef(multi_fit[[sym]])
  tibble(
    symbol = sym,
    alpha  = p["alpha1"],
    beta   = p["beta1"],
    gamma  = p["gamma1"],   # leverage (GJR)
    persistence = p["alpha1"] + p["beta1"] + p["gamma1"] / 2
  )
}))

# Persistence distribution across the 50 stocks
ggplot(params_df, aes(x = persistence)) +
  geom_histogram(bins = 20, fill = "#901A1E", alpha = 0.8) +
  labs(title = "GARCH persistence across 50 NASDAQ stocks",
       x = expression(alpha + beta + gamma/2), y = "Count") +
  theme_minimal()
```

## Python
```{python}
#| eval: false
params_df = pd.DataFrame([
    {"symbol": sym,
     "alpha": res.params["alpha[1]"],
     "beta": res.params["beta[1]"],
     "gamma": res.params.get("gamma[1]", 0),
     "persistence": res.params["alpha[1]"] + res.params["beta[1]"]}
    for sym, res in zip(returns.columns, results)
])
```
:::

**Expected findings:**

- Persistence $\alpha + \beta \approx 0.96$–$0.99$ across most stocks
- Leverage $\hat\gamma > 0$ for the majority — asymmetry is the rule, not the exception
- A few stocks with low persistence (high idiosyncratic volatility) — likely news-driven
- Sector patterns: tech stocks tend to have higher persistence and stronger leverage

## From GARCH to Portfolio Weights

With $\hat\sigma_{i,t|t-1}^2$ estimated for all 50 stocks, we can construct a time-varying diagonal covariance matrix:

$$\hat D_{t|t-1} = \text{diag}(\hat\sigma_{1,t|t-1}, \ldots, \hat\sigma_{50,t|t-1})$$

**Two approaches for combining with correlations:**

| Approach | Correlation | Lecture |
|----------|-------------|---------|
| $\hat\Sigma_t = \hat D_t \hat C \hat D_t$ | Constant sample $\hat C$ | This lecture (naive) |
| $\hat\Sigma_t = \hat D_t \hat C_t \hat D_t$ | Dynamic $\hat C_t$ from DCC | Lecture 5 |

**Preview:** DCC-GARCH in Lecture 5 reuses exactly the univariate estimates from this lecture — the first step of DCC *is* fitting individual GARCH models

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# Compute GMVP weights for each date using GARCH conditional variances
# + constant sample correlation
corr_sample <- cor(returns)

gmvp_garch <- map_dfr(1:nrow(sigma2_matrix), function(t) {
  sig_vec <- sqrt(sigma2_matrix[t, ])
  D <- diag(sig_vec)
  Sigma_t <- D %*% corr_sample %*% D
  w <- solve(Sigma_t, rep(1, 50))
  w <- w / sum(w)
  tibble(date = index(sigma2_matrix)[t],
         port_var = as.numeric(t(w) %*% Sigma_t %*% w))
})
```

## Python
```{python}
#| eval: false
import numpy as np

corr_sample = returns.corr().values

gmvp_vars = []
for t in range(len(sigma2_matrix)):
    sig_vec = np.sqrt(sigma2_matrix.iloc[t].values)
    D = np.diag(sig_vec)
    Sigma_t = D @ corr_sample @ D
    w = np.linalg.solve(Sigma_t, np.ones(50))
    w /= w.sum()
    gmvp_vars.append(w @ Sigma_t @ w)
```
:::

# Exercises

## Exercise Set 2

**Estimation**

1. Using `tsgarch` (R) or `arch` (Python), estimate GARCH(1,1), GJR-GARCH(1,1), and EGARCH(1,1) for a subset of 5 stocks from the NASDAQ universe. For each:
   - Report $(\hat\omega, \hat\alpha, \hat\beta, \hat\gamma)$ with robust standard errors
   - Compute persistence and the implied half-life of a volatility shock
   - Assess whether the leverage coefficient $\hat\gamma$ is statistically significant

2. Plot news impact curves for GARCH vs GJR-GARCH for one stock. How large is the asymmetry?

3. Check the diagnostics: are the standardised residuals $\hat\varepsilon_t$ free of ARCH effects (Ljung-Box on $\hat\varepsilon_t^2$, $p > 0.05$)? Does the distribution assumption look adequate from the QQ-plot?

**Batch estimation and cross-section**

4. Estimate GJR-GARCH(1,1) for all 50 NASDAQ stocks in parallel using `tsgarch` + `future` (R) or `joblib` (Python).
   - Plot the cross-sectional distribution of $\hat\alpha + \hat\beta + \hat\gamma/2$ (persistence)
   - Plot the cross-sectional distribution of the leverage coefficient $\hat\gamma$
   - Do tech stocks show more leverage than healthcare stocks?

**Forecasting and evaluation**

5. For a rolling estimation window of 5 years:
   - Produce one-step-ahead out-of-sample variance forecasts for 2 years
   - Compute QLIKE for GARCH(1,1), GJR-GARCH(1,1), EWMA(0.94), and rolling SD(21) for each stock
   - Apply the MCS (or pairwise DM tests with corrections) — which models survive?

**Portfolio evaluation**

6. Construct the GMVP using GARCH conditional variances combined with the sample correlation matrix. Compare out-of-sample portfolio volatility to the Lecture 1 naive estimators. Does time-varying variance alone help?

**Key references:** Hansen & Lunde (2005); Engle & Patton (2001); Hansen, Lunde & Nason (2011)
