---
subtitle: "Volatility V: Multivariate Volatility — DCC-GARCH & Shrinkage"
---

```{r}
#| echo: false
#| warning: false
#| eval: false
library(tidyverse)
library(tsmarch)
library(nlshrink)
```

## The Portfolio Problem Revisited

Lectures 1–4 focused on **marginal variances** $\sigma_{i,t}^2$ for individual stocks

**But the GMVP requires the full covariance matrix:**

$$w^* = \frac{\Sigma_t^{-1} \mathbf{1}}{\mathbf{1}' \Sigma_t^{-1} \mathbf{1}}$$

For $N = 50$ stocks:

- $\Sigma_t$ has $50 \times 51 / 2 = 1{,}275$ unique elements
- Each needs to be estimated from the **same** $T \approx 2{,}500$ daily observations
- For $N = 100$: 5,050 elements from 2,500 observations — more parameters than data

**Two approaches in this lecture:**

| Method | Core idea | Key assumption |
|--------|-----------|----------------|
| DCC-GARCH | Model conditional correlations dynamically | Correlations vary over time |
| Ledoit-Wolf shrinkage | Regularise the sample covariance toward a structured target | Optimal under large $N$, fixed $T$ |

Both are **complementary** — one can use DCC on top of shrinkage, or combine realized covariance with shrinkage

# The Curse of Dimensionality

## Why Not Just Use the Sample Covariance?

$$\hat\Sigma = \frac{1}{T-1}\sum_{t=1}^T (r_t - \hat\mu)(r_t - \hat\mu)'$$

**Problems:**

:::: {.columns}
::: {.column width="50%"}
**Rank deficiency:**

- For $N > T$: $\hat\Sigma$ is singular (not invertible)
- For our $N = 50$, $T = 2{,}500$: just barely fine, but barely
- After adding estimation noise: effectively near-singular

**Estimation error amplification:**

- Each off-diagonal element has standard error $\approx 1/\sqrt{T}$
- Inverting $\hat\Sigma$ *amplifies* estimation errors
- Small eigenvalues of $\hat\Sigma$ → large eigenvalues of $\hat\Sigma^{-1}$ → extreme portfolio weights
:::
::: {.column width="50%"}
**Eigenvalue bias (Marčenko-Pastur):**

For $N/T \to c \in (0, \infty)$:

- Empirical eigenvalues spread out relative to true eigenvalues
- Largest eigenvalue: over-estimated
- Smallest eigenvalue: under-estimated (pushed toward zero)
- Even with true $\Sigma = I$, the sample eigenvalues are not all equal to 1

**Portfolio consequence:**

- Eigenvectors associated with small eigenvalues receive huge inverse-variance weights
- These are the noisiest directions — the optimizer exploits estimation error (Michaud 1989)
:::
::::

## Why Not BEKK?

The BEKK model (Engle & Kroner 1995) models the full covariance matrix:

$$\Sigma_t = C'C + A' r_{t-1} r_{t-1}' A + B' \Sigma_{t-1} B$$

- Guarantees positive definiteness
- With $N = 50$: BEKK has $\frac{N(N+1)}{2} + 2N^2 = 1{,}275 + 5{,}000 = 6{,}275$ parameters
- **Infeasible** for $N \geq 5$; even diagonal BEKK ($A, B$ diagonal) struggles beyond $N \sim 10$

**We need structure.** Two directions:

## Interactive: The Eigenvalue Problem

*All true eigenvalues equal 1 (identity covariance). Drag c = N/T upward to watch estimation noise spread the sample eigenvalues — the smallest approach zero, the largest inflate.*

```{ojs}
//| echo: false
viewof c_rat = Inputs.range([0.02, 0.97], {step: 0.01, value: 0.20,
  label: "c = N / T  (concentration ratio, e.g. 0.02 = 50 stocks / 2500 days)"})
```

```{ojs}
//| echo: false
mp_pts = {
  const lmin = (1 - Math.sqrt(c_rat))**2
  const lmax = (1 + Math.sqrt(c_rat))**2
  return Array.from({length: 400}, (_, i) => {
    const lam = lmin + (lmax - lmin) * (i + 0.5) / 400
    const dens = Math.sqrt(Math.max((lmax - lam) * (lam - lmin), 0)) /
                 (2 * Math.PI * c_rat * lam)
    return { lam, dens }
  })
}
```

```{ojs}
//| echo: false
{
  const lmin = ((1 - Math.sqrt(c_rat))**2).toFixed(3)
  const lmax = ((1 + Math.sqrt(c_rat))**2).toFixed(3)
  const inv_inflate = (1 / (1 - Math.sqrt(c_rat))**2).toFixed(1)
  const note = html`Eigenvalues spread: [${lmin}, ${lmax}] &ensp;|&ensp;
    Smallest eigenvalue ${lmin} → Σ⁻¹ inflates by <b>${inv_inflate}×</b> &ensp;|&ensp;
    True value = 1 (all eigenvalues should be equal)`

  const plt = Plot.plot({
    width: 700, height: 265, marginLeft: 55,
    marks: [
      Plot.areaY(mp_pts, {x: "lam", y: "dens",
        fill: "#901A1E", fillOpacity: 0.25}),
      Plot.lineY(mp_pts, {x: "lam", y: "dens",
        stroke: "#901A1E", strokeWidth: 2.5}),
      Plot.ruleX([1], {stroke: "#3d3d3d", strokeDasharray: "4 2", strokeWidth: 1.5}),
      Plot.text([{x: 1.04, y: mp_pts.reduce((a,d) => Math.max(a, d.dens), 0) * 0.55,
                  label: "True λ = 1"}],
        {x: "x", y: "y", text: "label", fontSize: 11, fill: "#3d3d3d"}),
    ],
    x: {label: "Sample eigenvalue λ̂  (true value = 1)", grid: true},
    y: {label: "Marčenko-Pastur density", grid: true},
  })
  return html`${note}<br>${plt}`
}
```

1. **Factor structure:** decompose $\Sigma_t$ into a small number of common factors (GO-GARCH, Factor-DCC)
2. **Two-step decomposition:** model variances and correlations separately (DCC) — *this lecture*

# DCC-GARCH

## The DCC Model (Engle 2002)

**Dynamic Conditional Correlation:** decompose the covariance matrix as

$$\Sigma_t = D_t R_t D_t$$

where $D_t = \text{diag}(\sigma_{1,t}, \ldots, \sigma_{N,t})$ and $R_t$ is the **conditional correlation matrix**

**Two-step estimation:**

### Step 1: Univariate GARCH

Estimate GARCH (or GJR-GARCH) for each of the $N$ stocks separately → $\hat\sigma_{i,t}^2$ for all $i$

- This is *exactly* what we did in Lecture 2 — reuse those estimates
- Scalable: $N$ independent optimizations, fully parallelizable

### Step 2: Correlation dynamics

With standardised residuals $\hat z_{i,t} = r_{i,t} / \hat\sigma_{i,t}$, model the correlation:

$$Q_t = (1 - a - b)\bar Q + a\, \hat z_{t-1} \hat z_{t-1}' + b\, Q_{t-1}$$

$$R_t = \text{diag}(Q_t)^{-1/2}\, Q_t\, \text{diag}(Q_t)^{-1/2}$$

where $\bar Q = \text{Cov}(\hat z_t)$ (sample covariance of standardised residuals) and $a, b > 0$, $a + b < 1$

**Key property:** only **two** parameters $(a, b)$ govern all $N(N-1)/2$ pairwise correlations — massive dimension reduction

## DCC Interpretation

**CCC as special case:** $a = b = 0$ → $Q_t = \bar Q$ constant — correlations do not move (Bollerslev 1990)

**DCC dynamics:**

- $a$: reaction of correlations to recent co-movements (like $\alpha$ in univariate GARCH)
- $b$: persistence of correlations (like $\beta$ in univariate GARCH)
- Empirically: $a \approx 0.03$–$0.06$, $b \approx 0.92$–$0.96$ — correlations are persistent

**Limitation:** all $N(N-1)/2$ pairs share the same $(a, b)$ — correlations can move at different speeds in reality. Extensions: asymmetric DCC (ADCC), varying-parameter DCC

## Estimation with `tsmarch`

`tsmarch` (Galanos) is the modern R package for multivariate GARCH models:

- DCC, GO-GARCH, Copula-GARCH
- Parallelised via `future` + RcppParallel
- Consistent interface with `tsgarch` — uses the univariate fits from Lecture 2

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
library(tsmarch)

# Step 1: reuse univariate GJR-GARCH fits from Lecture 2
# multi_fit: named list of 50 tsgarch fit objects

# Step 2: DCC specification
dcc_spec <- dcc_modelspec(
  uspec  = multi_fit,       # list of univariate specs
  type   = "dcc",           # standard DCC
  order  = c(1, 1)
)

# Estimate DCC
dcc_fit <- estimate(dcc_spec)
summary(dcc_fit)

# Extract conditional correlations (T x N x N array)
R_t <- rcor(dcc_fit)       # correlation matrices over time

# Extract conditional covariances
Sigma_t <- rcov(dcc_fit)   # covariance matrices over time
```

## Python
```{python}
#| eval: false
# DCC in Python: use arch for Step 1, implement Step 2 manually
# (no production-ready DCC-GARCH library in Python as of 2026)

from arch import arch_model
import numpy as np

# Step 1: fit GARCH for each stock
garch_results = {sym: arch_model(returns[sym], vol="Garch", p=1, o=1, q=1,
                                  dist="t").fit(disp="off")
                 for sym in returns.columns}

# Standardised residuals: T x N matrix
Z = pd.DataFrame({sym: res.std_resid
                  for sym, res in garch_results.items()})

# Step 2: DCC parameter estimation via MLE (simplified)
Q_bar = Z.cov().values

def dcc_loglik(params, Z, Q_bar):
    a, b = params
    T, N = Z.shape
    Q = Q_bar.copy()
    loglik = 0
    for t in range(1, T):
        z = Z.iloc[t-1].values
        Q = (1 - a - b) * Q_bar + a * np.outer(z, z) + b * Q
        Q_diag_inv = np.diag(1 / np.sqrt(np.diag(Q)))
        R = Q_diag_inv @ Q @ Q_diag_inv
        loglik += -0.5 * (np.log(np.linalg.det(R)) +
                          Z.iloc[t].values @ np.linalg.inv(R) @ Z.iloc[t].values)
    return -loglik
```
:::

## Time-Varying Correlations in Practice

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# Plot conditional correlations for a tech-healthcare pair
cor_ts <- rcor(dcc_fit)  # T x N x N
aapl_jnj_cor <- cor_ts[, "AAPL", "JNJ"]

tibble(
  date   = index(returns),
  cor_dcc = aapl_jnj_cor,
  cor_ccc = R_bar["AAPL", "JNJ"]    # constant CCC correlation
) |>
  pivot_longer(-date) |>
  ggplot(aes(x = date, y = value, color = name)) +
  geom_line() +
  scale_color_manual(
    values = c(cor_dcc = "#901A1E", cor_ccc = "#666666"),
    labels = c("DCC (time-varying)", "CCC (constant)")
  ) +
  labs(title = "AAPL–JNJ conditional correlation",
       x = NULL, y = "Correlation", color = NULL) +
  theme_minimal()
```

## Python
```{python}
#| eval: false
import matplotlib.pyplot as plt

fig, ax = plt.subplots(figsize=(10, 4))
ax.plot(dates, aapl_jnj_dcc, label="DCC (time-varying)", color="#901A1E")
ax.axhline(R_bar["AAPL"]["JNJ"], color="#666666",
           linestyle="--", label="CCC (constant)")
ax.set_title("AAPL–JNJ conditional correlation")
ax.legend()
plt.tight_layout()
```
:::

**Typical empirical findings:**

- Correlations **spike during crises** (COVID March 2020, GFC 2008): diversification benefits collapse precisely when they are most needed
- Intra-sector correlations (tech-tech) persistently higher than cross-sector (tech-healthcare)
- DCC $(a, b)$ estimates: correlations are highly persistent ($a + b \approx 0.97$)

# Covariance Matrix Regularisation

## The Sample Covariance is Inefficient

Even for $N < T$, the sample covariance matrix is a **poor estimator** for portfolio optimization:

- Stein (1956): the sample covariance is **inadmissible** under Frobenius loss — a biased estimator with lower MSE always exists
- Intuition: pull extreme eigenvalues toward their population values — shrink

**Ledoit & Wolf (2004):** the optimal linear combination of $\hat\Sigma$ and a structured **target** $F$:

$$\hat\Sigma^{LW} = (1 - \hat\alpha)\hat\Sigma + \hat\alpha F$$

where $F$ is chosen to be well-conditioned and $\hat\alpha$ minimises the expected Frobenius loss

**Common targets $F$:**

| Target | Structure | Use case |
|--------|-----------|---------|
| Scaled identity $\mu I$ | All assets equal variance, zero correlation | Maximally regularised |
| Constant correlation | Equal off-diagonal correlations | Returns with similar variance |
| Single-factor | Market model covariance | When a market factor dominates |

## Linear Shrinkage (Ledoit & Wolf 2004)

The **oracle shrinkage intensity** $\alpha^*$ minimises $E[\|\hat\Sigma^{LW} - \Sigma\|_F^2]$

Ledoit & Wolf (2004) derive an **analytical formula** for the optimal shrinkage intensity — no cross-validation needed:

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
library(nlshrink)  # or covShrinkage from RiskPortfolios

# Linear shrinkage toward constant-correlation target
Sigma_lw_linear <- linshrink_cov(returns_matrix)

# Compare eigenvalue distributions
ev_sample <- eigen(cov(returns_matrix))$values
ev_shrink  <- eigen(Sigma_lw_linear)$values

tibble(
  type  = rep(c("Sample", "Shrinkage"), each = ncol(returns_matrix)),
  value = c(ev_sample, ev_shrink)
) |>
  ggplot(aes(x = value, fill = type)) +
  geom_histogram(position = "dodge", bins = 30, alpha = 0.7) +
  scale_fill_manual(values = c("Sample" = "#666666", "Shrinkage" = "#901A1E")) +
  labs(title = "Eigenvalue distribution: sample vs shrinkage",
       x = "Eigenvalue", y = "Count", fill = NULL) +
  theme_minimal()
```

## Python
```{python}
#| eval: false
from sklearn.covariance import LedoitWolf

lw = LedoitWolf().fit(returns_matrix)
Sigma_lw = lw.covariance_
alpha_lw = lw.shrinkage_

print(f"Shrinkage intensity: {alpha_lw:.3f}")
```
:::

## Nonlinear Shrinkage (Ledoit & Wolf 2017, 2020)

**Motivation:** linear shrinkage applies the same proportional pull to all eigenvalues — not optimal

**Oracle nonlinear shrinkage:** optimal to shrink each eigenvalue $\hat\lambda_k$ separately toward its oracle value

$$\tilde\lambda_k = \frac{\lambda_k}{|1 + c\, \int \frac{\lambda_k}{\lambda_k - x} dF^{\hat\Sigma}(x)|^2}$$

where $c = N/T$ and the integral is over the empirical spectral distribution — this is the **Ledoit-Wolf (2020) analytical formula**

- No cross-validation
- Optimal for large $N$ and $T$ with $N/T \to c > 0$
- Consistently outperforms linear shrinkage for $N > 30$

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# Nonlinear analytical shrinkage (Ledoit & Wolf 2020)
Sigma_lw_nl <- nlshrink_cov(returns_matrix, k = 1)
# k = 1: leave out 1 eigenvalue (spike) — appropriate when a market factor exists
```

## Python
```{python}
#| eval: false
# nonlinear shrinkage via the nlshrink Python package
# pip install nonlinear_shrinkage
from nonlinear_shrinkage import analytical_shrinkage

Sigma_lw_nl = analytical_shrinkage(returns_matrix)
```
:::

## Shrinkage vs DCC: When to Use Which

| Feature | DCC-GARCH | Ledoit-Wolf Shrinkage |
|---------|----------|----------------------|
| Time variation | Yes — $R_t$ changes daily | No — static estimator |
| Data required | Daily returns (large $T$) | Cross-sectional (large $N$) |
| Works when $N > T$ | No (Step 1 is fine, Step 2 can fail) | Yes — designed for this |
| Positive definiteness | Guaranteed | Guaranteed |
| Computational cost | Medium (two-step) | Very fast ($O(N^2)$) |
| Captures correlation dynamics | Yes | No |

**Best of both worlds:** apply Ledoit-Wolf shrinkage to the DCC correlation matrix $R_t$:

$$\hat\Sigma_t^{DCC+LW} = D_t \left[(1-\alpha)\hat R_t + \alpha I\right] D_t$$

Or use shrinkage as a regularizer for the DCC target $\bar Q$

# Portfolio Horse Race

## GMVP Construction

With four covariance estimators, construct and compare GMVP portfolios:

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# Compute GMVP weights for each estimator at each rebalancing date
compute_gmvp_weights <- function(Sigma) {
  N <- nrow(Sigma)
  ones <- rep(1, N)
  Sigma_inv <- solve(Sigma)
  w <- Sigma_inv %*% ones
  w / sum(w)
}

# For each date t in the out-of-sample period:
results <- map_dfr(oos_dates, function(t) {
  # Covariance estimates at t
  Sigma_roll   <- rolling_cov_matrix[t,,]        # rolling 252-day
  Sigma_garch  <- garch_cov_matrix[t,,]           # GARCH + sample corr
  Sigma_dcc    <- rcov(dcc_fit)[t,,]              # DCC-GARCH
  Sigma_lw     <- nlshrink_cov(returns[1:t, ])   # rolling nonlinear shrinkage

  # Compute weights and realised variance next period
  r_next <- as.numeric(returns[t+1, ])

  bind_rows(
    tibble(method = "Rolling", t = t,
           port_var = as.numeric(t(compute_gmvp_weights(Sigma_roll)) %*%
                                   Sigma_roll %*% compute_gmvp_weights(Sigma_roll))),
    tibble(method = "GARCH",   t = t, ...),
    tibble(method = "DCC",     t = t, ...),
    tibble(method = "LW",      t = t, ...)
  )
})
```

## Python
```{python}
#| eval: false
import numpy as np

def gmvp_weights(Sigma):
    N = Sigma.shape[0]
    ones = np.ones(N)
    Sigma_inv = np.linalg.inv(Sigma)
    w = Sigma_inv @ ones
    return w / w.sum()

portfolio_results = []
for t in oos_dates:
    for method, Sigma in [("Rolling", Sigma_roll[t]),
                           ("GARCH",   Sigma_garch[t]),
                           ("DCC",     Sigma_dcc[t]),
                           ("LW",      Sigma_lw[t])]:
        w = gmvp_weights(Sigma)
        r_real = returns.iloc[t+1].values
        portfolio_results.append({
            "method": method, "t": t,
            "real_ret": w @ r_real,
            "real_var": w @ Sigma @ w
        })
```
:::

## Evaluation Metrics

**Primary metric:** annualised out-of-sample portfolio variance

$$\widehat{\text{PortVar}} = \text{Var}\left(\sum_i w_{i,t} r_{i,t+1}\right) \times 252$$

**Secondary metrics:**

| Metric | Formula | Interpretation |
|--------|---------|----------------|
| Realised variance | $252 \times \text{Var}(w_t' r_{t+1})$ | Lower is better for GMVP |
| Sharpe ratio | $\bar r_p / \sigma_p \times \sqrt{252}$ | Higher is better (risk-adjusted) |
| Turnover | $\sum_t \|w_t - w_{t-1}\|_1$ | Lower = less transaction cost |
| Max drawdown | $\max_t (V_0 - V_t) / V_0$ | Lower is better for risk mgmt |

**Testing for significance:** Ledoit & Wolf (2008) test for equality of two portfolio variances (bootstrap-based, robust to non-i.i.d. returns)

## Preview: Horse Race Results

The final evaluation (Lecture 6 will add realized covariance):

| Method | Est. GMVP variance | Turnover | Shrinkage needed? |
|--------|-------------------|----------|-------------------|
| Naive 1/N | High | Zero | — |
| Rolling SD + sample $\hat C$ | Moderate | Low | No |
| GARCH + sample $\hat C$ | Lower | Medium | No |
| DCC-GARCH | Lower | Medium | Marginal |
| Ledoit-Wolf (nonlinear) | Low | Low | Built-in |
| DCC + LW | Lowest? | Medium | Built-in |
| Realized Cov (Lect. 6) | TBD | Medium | Essential |

**Key takeaway so far:** better volatility modelling reduces GMVP variance, but only if estimation error is controlled — shrinkage is as important as dynamic modelling

# Exercises

## Exercise Set 5

**DCC-GARCH**

1. Using `tsmarch`, estimate DCC-GARCH for the 50 NASDAQ stocks, building on your Lecture 2 GJR-GARCH univariate fits.
   - Report the DCC parameters $(a, b)$ and their standard errors
   - What is the persistence $a + b$ of the correlation process?
   - Plot the time series of conditional correlations for 3 pairs:
     (a) two tech stocks (e.g., AAPL–MSFT)
     (b) tech–healthcare (e.g., AAPL–JNJ)
     (c) tech–consumer (e.g., AAPL–AMZN)

2. Identify the 5 dates with the highest average pairwise conditional correlation. Do they coincide with known market events?

3. Compare DCC correlations to CCC (constant) correlations. Do correlations spike during the COVID-19 market crash (Feb–Mar 2020)?

**Shrinkage**

4. Apply linear (Ledoit-Wolf 2004) and nonlinear analytical shrinkage (Ledoit-Wolf 2020) to the rolling 252-day sample covariance matrix.
   - Plot the eigenvalue distribution before and after shrinkage. How does the distribution change?
   - How does the shrinkage intensity $\hat\alpha$ vary over time?

5. Compare portfolio weights under shrinkage vs sample covariance: are the weights more stable (less extreme) under shrinkage?

**Portfolio horse race**

6. Estimate the GMVP using each of the following covariance estimators in a rolling out-of-sample evaluation (5-year training, 2-year test):
   - Naive 1/N
   - Rolling 252-day sample covariance
   - GARCH diagonal + sample correlation
   - CCC-GARCH
   - DCC-GARCH
   - Ledoit-Wolf nonlinear shrinkage

   Report annualised realised portfolio variance, Sharpe ratio, and daily turnover. Apply the Ledoit-Wolf (2008) test to identify which performance differences are statistically significant.

**Key references:** Engle (2002); Ledoit & Wolf (2004); Ledoit & Wolf (2020)
