---
subtitle: "Volatility I: Stylized Facts & First Estimates"
---

```{r}
#| echo: false
#| warning: false
#| eval: false
library(tidyverse)
library(tidyfinance)
library(scales)
```

## Where We Are

The **Global Minimum Variance Portfolio** is our running testbed throughout all six lectures:

$$w^* = \frac{\Sigma_t^{-1} \mathbf{1}}{\mathbf{1}' \Sigma_t^{-1} \mathbf{1}}$$

Every modelling choice we make will be evaluated by one question: does it reduce out-of-sample portfolio variance?

**The six-lecture roadmap:**

| Lecture | What we estimate | Building block |
|---------|-----------------|----------------|
| 1 | Individual $\hat\sigma_{i,t}$ — rolling SD and EWMA | Diagonal $\hat D_t$ |
| 2 | $\hat\sigma_{i,t}$ from GARCH family | Better $\hat D_t$ |
| 3 | $\hat\sigma_{i,t}$ from intraday realized variance | Precise $\hat D_t$ |
| 4 | $\hat\sigma_{i,t+h}$ — HAR forecasts | Forecast $\hat D_{t+h}$ |
| 5 | Full $\hat\Sigma_t$ — DCC-GARCH and shrinkage | $\hat D_t \hat C_t \hat D_t$ |
| 6 | Full $\hat\Sigma_t$ — realized covariance | Intraday $\hat\Sigma_t$ |

Loss function we use throughout: **QLIKE** — a proper loss for variance forecasts.

## The Portfolio Lens

We work with **50 NASDAQ stocks** (2010–2024) across tech, healthcare, and consumer discretionary. This gives us 1,275 unique covariance elements — large enough to make dimensionality a real challenge.

The decomposition that structures the whole block:

$$\hat\Sigma_t = \hat D_t \hat C_t \hat D_t$$

where $\hat D_t = \text{diag}(\hat\sigma_{1,t}, \ldots, \hat\sigma_{N,t})$ collects individual volatilities and $\hat C_t$ is the correlation matrix.

**Lectures 1–4** build better models for the diagonal $\hat D_t$. **Lectures 5–6** tackle the full $\hat C_t$, which spikes in crises — precisely when diversification is most needed.

## The Core Problem: Volatility is Latent

We observe the return, not the variance that generated it:

$$r_t \sim \mathcal{D}(0,\, \sigma_t^2) \quad \text{— } \sigma_t^2 \text{ is unobservable}$$

A static sample variance assumes $\sigma_t^2$ is constant. But for portfolio rebalancing we need a **forecast**: $E(\sigma_{t+1}^2 \mid \mathcal{F}_t)$.

**Three things the data make unambiguously clear:**

1. Volatility is **not constant** — calm and turbulent periods are visibly different
2. Volatility is **predictable from its own history** — today tells us something about tomorrow
3. Volatility **mean-reverts** — elevated periods eventually return to a long-run level

These are the stylized facts. Models that ignore them leave forecasting accuracy on the table.

# Stylized Facts

## Four Facts, One Stock

```{r}
#| echo: false
#| warning: false
#| cache: true

prices_sf <- download_data(
  type = "stock_prices", symbols = "AAPL",
  start_date = "2015-01-01", end_date = "2024-12-31"
) |>
  arrange(date) |>
  mutate(ret = log(adjusted_close / lag(adjusted_close))) |>
  drop_na()

ci_sf <- 1.96 / sqrt(nrow(prices_sf))

acf_sq <- with(acf(prices_sf$ret^2, lag.max = 60, plot = FALSE),
               list(lag = as.numeric(lag[-1]), acf = as.numeric(acf[-1])))

cc_obj <- ccf(prices_sf$ret, prices_sf$ret^2, lag.max = 10, plot = FALSE)
cc_lev <- list(lag = as.numeric(cc_obj$lag[as.numeric(cc_obj$lag) >= 0]),
               ccf = as.numeric(cc_obj$acf[as.numeric(cc_obj$lag) >= 0]))

ojs_define(
  sf_returns = prices_sf |> mutate(date = as.character(date)) |>
                 select(date, ret) |> as.list(),
  sf_acf     = map2(acf_sq$lag, acf_sq$acf, ~ list(lag = .x, acf = .y)),
  sf_ccf     = map2(cc_lev$lag, cc_lev$ccf, ~ list(lag = .x, ccf = .y)),
  sf_ci      = ci_sf
)
```

```{ojs}
//| echo: false
ret_data = sf_returns.date.map((d, i) => ({ date: new Date(d), ret: sf_returns.ret[i] * 100 }))

p_ret = Plot.plot({
  width: 340, height: 200, marginLeft: 42,
  marks: [Plot.lineY(ret_data, {x: "date", y: "ret", stroke: "#3d3d3d", strokeWidth: 0.5})],
  x: {label: null}, y: {label: "Return (%)"},
  title: "(a) Daily log returns — AAPL 2015–2024"
})

p_qq = Plot.plot({
  width: 340, height: 200, marginLeft: 42,
  marks: [
    Plot.qq(ret_data, {x: "ret", stroke: "#901A1E", strokeOpacity: 0.4, strokeWidth: 1.5}),
  ],
  title: "(b) QQ-plot vs Gaussian (tails outside line = fat tails)"
})

p_acf = Plot.plot({
  width: 340, height: 200, marginLeft: 42,
  marks: [
    Plot.barY(sf_acf, {x: "lag", y: "acf", fill: "#901A1E", fillOpacity: 0.75}),
    Plot.ruleY([sf_ci, -sf_ci], {stroke: "#666666", strokeDasharray: "3 2"})
  ],
  x: {label: "Lag (days)"}, y: {label: "ACF"},
  title: "(c) ACF of r²ₜ — slow decay = clustering"
})

p_lev = Plot.plot({
  width: 340, height: 200, marginLeft: 42,
  marks: [
    Plot.barY(sf_ccf, {x: "lag", y: "ccf", fill: "#901A1E", fillOpacity: 0.75}),
    Plot.ruleY([0, sf_ci, -sf_ci], {stroke: ["#3d3d3d","#666666","#666666"],
               strokeDasharray: ["0","3 2","3 2"]})
  ],
  x: {label: "Lag k"}, y: {label: "CCF"},
  title: "(d) Corr(rₜ, r²ₜ₊ₖ) — negative at k>0 = leverage effect"
})

html`<div style="display:grid;grid-template-columns:1fr 1fr;gap:6px">${p_ret}${p_qq}${p_acf}${p_lev}</div>`
```

**What to look for:** (a) calm and turbulent stretches — *clustering*. (b) points outside the diagonal at both ends — *fat tails*. (c) positive bars decaying slowly — *volatility autocorrelated for months*. (d) negative bars at positive lags — *bad news raises future variance more than good news*.

## Fact 1: Fat Tails

Daily returns have **excess kurtosis** — extreme moves are far more likely than a Gaussian predicts:

$$\text{kurt}(r_t) = \frac{E[(r_t - \mu)^4]}{\sigma^4} \gg 3$$

| Threshold | Gaussian predicts | Reality (US equities) |
|-----------|------------------|-----------------------|
| $\|r_t\| > 3\sigma$ | 0.3% of days | ~1–2% of days |
| $\|r_t\| > 5\sigma$ | Once per 3.5 million days | Several per year |

::: {.callout-note icon=false}
**Portfolio implication:** Gaussian VaR systematically underestimates tail risk. A covariance matrix calibrated under normality misprices exactly the scenarios that matter most — the large crashes, not the quiet days.
:::

## Fact 2: Volatility Clustering

**Mandelbrot (1963):** *"large changes tend to be followed by large changes, of either sign, and small changes tend to be followed by small changes."*

$$\text{Corr}(r_t^2,\, r_{t-k}^2) > 0 \quad \text{for } k = 1, 2, \ldots$$

The ACF of $r_t^2$ is significantly positive at many lags and decays slowly — consistent with long memory. This means volatility is **predictable from its own history**: a dynamic $\hat\sigma_{i,t}$ that responds to recent squared returns will outperform a fixed $\hat\sigma$.

::: {.callout-note icon=false}
**Portfolio implication:** position sizing should respond to the current volatility regime. A fund that ignores clustering holds the same position into a crisis as it does in calm markets.
:::

## Fact 3: Mean Reversion & the Leverage Effect

**Mean reversion:** elevated volatility eventually subsides. The empirical half-life of a volatility shock is roughly **10–30 trading days** for most equities. A model without mean reversion — like EWMA — keeps forecasts elevated indefinitely after a crisis, generating defensive positioning that never unwinds.

**The leverage effect:** negative returns raise future volatility more than positive returns of the same magnitude:

$$\text{Corr}(r_t,\, \sigma_{t+1}^2) < 0$$

Two channels explain this: (i) a price drop increases debt-to-equity, making the firm riskier; (ii) higher expected volatility raises the required return, so prices must fall further. Either way, drawdowns and rising risk arrive simultaneously.

::: {.callout-note icon=false}
**Portfolio implication:** symmetric models (EWMA, basic GARCH) treat a −3% day and a +3% day identically. GJR-GARCH (Lecture 2) captures the asymmetry directly — it matters for hedging costs.
:::

# Naive Estimators

## Rolling Window Standard Deviation

$$\hat\sigma_{t}^2 = \frac{1}{M-1} \sum_{s=t-M+1}^{t} (r_s - \bar r_{t,M})^2$$

The **cliff-edge problem:** when a large return exits the window, $\hat\sigma_t^2$ jumps even though no new information arrived today. The bias-variance tradeoff: short $M$ reacts fast but is noisy; long $M$ is smooth but stale.

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
rolling_vol <- crsp_returns |>
  group_by(symbol) |>
  arrange(date) |>
  mutate(
    vol_21  = slider::slide_dbl(ret, sd, .before = 20, .complete = TRUE),
    vol_63  = slider::slide_dbl(ret, sd, .before = 62, .complete = TRUE),
    vol_126 = slider::slide_dbl(ret, sd, .before = 125, .complete = TRUE)
  ) |>
  ungroup()
```

## Python
```{python}
#| eval: false
rolling_vol = (crsp_returns
  .sort_values(["symbol", "date"])
  .groupby("symbol")["ret"]
  .apply(lambda x: x.rolling(window=21, min_periods=21).std())
  .reset_index(name="vol_21")
)
```
:::

## Exponentially Weighted Moving Average (EWMA)

The **RiskMetrics** estimator (J.P. Morgan 1994):

$$\hat\sigma_t^2 = (1 - \lambda)\, r_{t-1}^2 + \lambda\, \hat\sigma_{t-1}^2$$

$\lambda = 0.94$ is the RiskMetrics default, implying a half-life of about **11 trading days**. No cliff-edge: all past returns included with exponentially declining weights.

**Important:** setting $\alpha + \beta = 1$ (as EWMA does implicitly) means no mean reversion — volatility forecasts never converge to a long-run level. This is the EWMA's biggest practical flaw.

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
ewma_vol <- function(ret, lambda = 0.94) {
  n <- length(ret)
  sigma2 <- numeric(n)
  sigma2[1] <- ret[1]^2
  for (t in 2:n) {
    sigma2[t] <- (1 - lambda) * ret[t-1]^2 + lambda * sigma2[t-1]
  }
  sqrt(sigma2)
}

ewma_estimates <- crsp_returns |>
  group_by(symbol) |>
  arrange(date) |>
  mutate(ewma_94 = ewma_vol(ret, lambda = 0.94),
         ewma_97 = ewma_vol(ret, lambda = 0.97)) |>
  ungroup()
```

## Python
```{python}
#| eval: false
def ewma_vol(ret, lam=0.94):
    sigma2 = [ret.iloc[0]**2]
    for r in ret.iloc[1:]:
        sigma2.append((1 - lam) * r**2 + lam * sigma2[-1])
    return pd.Series(sigma2, index=ret.index) ** 0.5
```
:::

## Interactive: Rolling SD vs EWMA

*Guided task: Set M=5 and λ=0.97. Find the largest spike in the series — which estimator reacts first? Now set M=126 and observe what happens after the spike exits the window. That sudden jump with no new information is the cliff-edge problem. Finally, ask yourself: why does EWMA with λ close to 1 behave like a long rolling window — but without the cliff-edge?*

```{ojs}
//| echo: false
viewof M_roll = Inputs.range([5, 126], {step: 1, value: 21,
  label: "Rolling window  M (days)"})
viewof lam_ewma = Inputs.range([0.85, 0.99], {step: 0.01, value: 0.94,
  label: "EWMA decay  λ"})
```

```{ojs}
//| echo: false
// Simulate a GJR-GARCH(1,1) return series (fixed seed via deterministic init)
garch_path = {
  const n = 750, omega = 4e-5, a = 0.06, g = 0.10, b = 0.89
  let s2 = omega / (1 - a - g/2 - b), r = 0
  const out = []
  // Use a simple LCG for reproducibility
  let seed = 42
  const lcg = () => { seed = (seed * 1664525 + 1013904223) & 0xffffffff
    return (seed >>> 0) / 4294967296 }
  const randn = () => Math.sqrt(-2*Math.log(lcg()+1e-10)) * Math.cos(2*Math.PI*lcg())
  for (let t = 0; t < n; t++) {
    const z = randn()
    r = Math.sqrt(s2) * z
    out.push({ t, r, true_vol: Math.sqrt(s2 * 252) * 100 })
    s2 = omega + (a + (r < 0 ? g : 0)) * r*r + b * s2
  }
  return out
}
```

```{ojs}
//| echo: false
vol_paths = {
  const rets = garch_path.map(d => d.r)
  const n = rets.length
  const out = garch_path.map(d => ({...d, roll: null, ewma: null}))
  // Rolling SD
  for (let t = M_roll - 1; t < n; t++) {
    const w = rets.slice(t - M_roll + 1, t + 1)
    const mu = w.reduce((a,x) => a+x, 0) / M_roll
    const v  = w.reduce((a,x) => a+(x-mu)**2, 0) / (M_roll-1)
    out[t].roll = Math.sqrt(v * 252) * 100
  }
  // EWMA
  let ev = rets[0]**2
  for (let t = 0; t < n; t++) {
    if (t > 0) ev = (1 - lam_ewma) * rets[t-1]**2 + lam_ewma * ev
    out[t].ewma = Math.sqrt(ev * 252) * 100
  }
  return out
}
```

```{ojs}
//| echo: false
Plot.plot({
  width: 700, height: 270, marginLeft: 50,
  marks: [
    Plot.lineY(vol_paths, {x: "t", y: "true_vol",
      stroke: "#cccccc", strokeWidth: 1.2}),
    Plot.lineY(vol_paths.filter(d => d.roll !== null), {x: "t", y: "roll",
      stroke: "#901A1E", strokeWidth: 2}),
    Plot.lineY(vol_paths, {x: "t", y: "ewma",
      stroke: "#3d3d3d", strokeWidth: 2, strokeDasharray: "5 3"}),
  ],
  y: {label: "Annualised vol (%)", grid: true},
  x: {label: "Trading day"},
  caption: html`<span style='color:#cccccc'>——</span> True (GJR-GARCH) &ensp;
    <span style='color:#901A1E'>——</span> Rolling SD (M days) &ensp;
    <span style='color:#3d3d3d'>– –</span> EWMA (λ)`
})
```

## From Volatility Estimates to Portfolio Weights

With $\hat\sigma_{i,t}$ from either rolling SD or EWMA, we form:

$$\hat\Sigma_t = \hat D_t \hat C \hat D_t$$

where $\hat C$ is the constant sample correlation matrix and $\hat D_t = \text{diag}(\hat\sigma_{1,t}, \ldots, \hat\sigma_{N,t})$.

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
compute_gmvp <- function(sigma_vec, corr_mat) {
  D <- diag(sigma_vec)
  Sigma <- D %*% corr_mat %*% D
  w <- solve(Sigma, rep(1, nrow(Sigma)))
  w / sum(w)
}

sample_corr <- cor(returns_wide)
gmvp_weights <- rolling_vol |>
  group_by(date) |>
  summarise(weights = list(compute_gmvp(vol_21, sample_corr)))
```

## Python
```{python}
#| eval: false
def compute_gmvp(sigma_vec, corr_mat):
    D = np.diag(sigma_vec)
    Sigma = D @ corr_mat @ D
    w = np.linalg.solve(Sigma, np.ones(len(sigma_vec)))
    return w / w.sum()

sample_corr = returns_wide.corr().values
```
:::

::: {.callout-note icon=false}
**Portfolio implication:** both rolling SD and EWMA capture clustering qualitatively, but keeping $\hat C$ constant misses the fact that correlations spike in crises (Fact 3). Better variance models (Lectures 2–4) improve $\hat D_t$; time-varying correlations (Lectures 5–6) improve $\hat C_t$.
:::

## Bridge to Lecture 2

**What naive estimators cannot do:**

- $\lambda$ in EWMA is hard-coded, not estimated from data
- Both rolling SD and EWMA are symmetric — they ignore the leverage effect
- Neither enforces mean reversion

**Lecture 2** introduces GARCH: likelihood-based estimation, built-in asymmetry via GJR-GARCH, and mean reversion by construction. The same portfolio testbed applies — we will compare GARCH-based $\hat D_t$ directly against the rolling SD and EWMA estimates from today.

# Exercises

## Exercise Set 1

**Stylized facts**

1. For 5 stocks (tech and healthcare): plot QQ-plots against the Gaussian distribution. Compute excess kurtosis for each. Does the ranking align with your prior about which sectors have fatter tails?

2. Compute and plot the cross-correlations $\text{Corr}(r_t, r_{t+k}^2)$ for $k = -10, \ldots, 10$ for the same stocks. Is the leverage effect (negative CCF at $k > 0$) consistent across sectors?

**Naive estimators**

3. Compute rolling SD at $M \in \{21, 63, 126\}$ for all 50 stocks. Find a date where rolling SD jumps despite no unusual return *today*. What happened $M$ days earlier? This is the cliff-edge problem in the data.

**Portfolio evaluation**

4. Construct the GMVP using rolling 21-day SD, rolling 63-day SD, and EWMA(0.94) combined with the sample correlation matrix. Report annualised out-of-sample portfolio volatility. Which estimator wins, and why does the result match your intuition about the bias-variance tradeoff?

**Key references:** Cont (2001); Mandelbrot (1963)
