---
subtitle: "Volatility IV: Forecasting Realized Volatility — HAR Models"
---

```{r}
#| echo: false
#| warning: false
#| eval: false
library(tidyverse)
library(highfrequency)
library(sandwich)
library(lmtest)
```

## Where We Are

**Progress so far:**

- Lecture 1: volatility is latent, exhibits rich dynamics (clustering, leverage, long memory)
- Lecture 2: GARCH extracts a signal from daily returns alone
- Lecture 3: realized variance $RV_t$ from intraday data is a precise, low-noise estimator of $\sigma_t^2$

**This lecture:** we have a clean time series $\{RV_t\}$ for each of our 50 NASDAQ stocks — now **forecast** it

**Why forecasting matters for portfolios:**

At time $t$, the portfolio optimizer requires $E(\sigma_{i,t+1}^2 \mid \mathcal{F}_t)$ — a **forecast**, not the realized value

$$\hat\Sigma_{t+1|t} = \hat D_{t+1|t} \hat C_{t+1|t} \hat D_{t+1|t}$$

The HAR model provides $\hat D_{t+1|t}$; DCC (Lecture 5) and realized covariance (Lecture 6) will provide $\hat C_{t+1|t}$

## The Heterogeneous Market Hypothesis

**Müller et al. (1997):** financial markets are populated by agents operating at **different time horizons**

- **Daily traders:** short-term speculators, market makers, high-frequency traders
- **Weekly portfolio managers:** institutional investors rebalancing weekly
- **Monthly investors:** pension funds, long-only asset managers

Each group perceives and reacts to volatility at their own horizon:

- Daily traders react to yesterday's news
- Weekly managers update positions based on the past week's activity
- Monthly investors are almost insensitive to daily fluctuations

**Implication for volatility dynamics:**

The observed long-memory-like behavior in $RV_t$ arises from the *superposition* of volatility shocks generated at different time scales

The HAR model operationalises this idea with a simple, estimable regression

# The HAR Model

## HAR-RV Specification

**Corsi (2009):** the Heterogeneous AutoRegressive model for Realized Variance

$$RV_{t+1} = \beta_0 + \beta_d\, RV_t^{(d)} + \beta_w\, RV_t^{(w)} + \beta_m\, RV_t^{(m)} + \varepsilon_{t+1}$$

where:

$$RV_t^{(d)} = RV_t, \quad RV_t^{(w)} = \frac{1}{5}\sum_{j=0}^{4} RV_{t-j}, \quad RV_t^{(m)} = \frac{1}{{22}}\sum_{j=0}^{21} RV_{t-j}$$

**Three regressors, one equation:**

- $\beta_d$: reaction of tomorrow's variance to yesterday's variance (daily component)
- $\beta_w$: reaction to the average of the past week (weekly component)
- $\beta_m$: reaction to the average of the past month (monthly component)
- All three $\beta > 0$ typically: volatility is positively autocorrelated at all horizons

**Why it works:**

- HAR is a restricted AR(22) — but estimated with only 3 parameters
- The averaging structure imposes a specific pattern of AR coefficients that matches the empirical slow decay
- Mimics long memory without estimating a fractional integration parameter

## Estimation

OLS with Newey-West standard errors (to account for autocorrelation in $\varepsilon_{t+1}$):

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
library(highfrequency)
library(sandwich)
library(lmtest)

# Pull RV from VOLARE (pre-computed daily measures for all 50 stocks)
# rv_data: data frame with columns date, symbol, rv, bpv, rv_kernel

# Construct HAR regressors for AAPL
har_data <- rv_data |>
  filter(symbol == "AAPL") |>
  arrange(date) |>
  mutate(
    rv_d = rv,
    rv_w = slider::slide_dbl(rv, mean, .before = 4, .complete = TRUE),
    rv_m = slider::slide_dbl(rv, mean, .before = 21, .complete = TRUE),
    rv_lead = lead(rv)  # next-day RV as dependent variable
  ) |>
  drop_na()

# Estimate HAR-RV
har_fit <- lm(rv_lead ~ rv_d + rv_w + rv_m, data = har_data)
coeftest(har_fit, vcov = NeweyWest(har_fit, lag = 5))
```

## Python
```{python}
#| eval: false
import pandas as pd
import numpy as np
import statsmodels.formula.api as smf
from statsmodels.stats.sandwich_covariance import cov_hac

# Construct HAR regressors
df = rv_data[rv_data["symbol"] == "AAPL"].sort_values("date").copy()
df["rv_d"] = df["rv"]
df["rv_w"] = df["rv"].rolling(5).mean()
df["rv_m"] = df["rv"].rolling(22).mean()
df["rv_lead"] = df["rv"].shift(-1)
df = df.dropna()

# Estimate HAR-RV with HAC standard errors
har_model = smf.ols("rv_lead ~ rv_d + rv_w + rv_m", data=df).fit()
hac_results = har_model.get_robustcov_results(cov_type="HAC", maxlags=5)
print(hac_results.summary())
```
:::

**Using `highfrequency::HARmodel()`:**

```{r}
#| eval: false
# Alternatively, use the dedicated HAR function
har_fit2 <- HARmodel(
  data     = rv_aapl,
  periods  = c(1, 5, 22),
  type     = "HAR",
  h        = 1
)
summary(har_fit2)
```

## Extensions: HAR-J and HAR-CJ

### HAR-J (Andersen, Bollerslev & Diebold 2007)

Add the jump component as an additional predictor:

$$RV_{t+1} = \beta_0 + \beta_d RV_t^{(d)} + \beta_w RV_t^{(w)} + \beta_m RV_t^{(m)} + \beta_J J_t + \varepsilon_{t+1}$$

where $J_t = \max(RV_t - BPV_t, 0)$ is the estimated jump variation

- If $\hat\beta_J < \hat\beta_d$: jumps are less persistent than continuous variance — they don't forecast future volatility well
- Typical finding: $\hat\beta_J$ is small and often insignificant — jumps are transient

### HAR-CJ (Andersen, Bollerslev & Diebold 2007)

Separate continuous and jump components as regressors:

$$RV_{t+1} = \beta_0 + \beta_{Cd} C_t^{(d)} + \beta_{Cw} C_t^{(w)} + \beta_{Cm} C_t^{(m)} + \beta_{Jd} J_t^{(d)} + \varepsilon_{t+1}$$

where $C_t = BPV_t$ (continuous) and $J_t = \max(RV_t - BPV_t, 0)$ (jump)

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# HAR-CJ using highfrequency
har_cj <- HARmodel(
  data    = cbind(rv = rv_aapl, bpv = bpv_aapl),
  periods = c(1, 5, 22),
  type    = "HARCJ",
  h       = 1
)
summary(har_cj)
```

## Python
```{python}
#| eval: false
# Construct HAR-CJ regressors manually
df["cont"]   = df["bpv"]
df["jump"]   = np.maximum(df["rv"] - df["bpv"], 0)
df["cont_w"] = df["cont"].rolling(5).mean()
df["cont_m"] = df["cont"].rolling(22).mean()
df["jump_d"] = df["jump"]

har_cj = smf.ols(
    "rv_lead ~ cont + cont_w + cont_m + jump_d", data=df
).fit()
print(har_cj.get_robustcov_results(cov_type="HAC", maxlags=5).summary())
```
:::

## The Leverage HAR (Patton & Sheppard 2015)

**Motivation:** the leverage effect (Lecture 1) applies to realized variance too — negative return days are followed by higher realized variance

Define **signed semivariances**:

$$RS_t^+ = \sum_{j: r_{t,j} > 0} r_{t,j}^2, \quad RS_t^- = \sum_{j: r_{t,j} < 0} r_{t,j}^2, \quad RV_t = RS_t^+ + RS_t^-$$

**Leverage HAR-RS:**

$$RV_{t+1} = \beta_0 + \beta^+ RS_t^{+(d)} + \beta^- RS_t^{-(d)} + \beta_w RV_t^{(w)} + \beta_m RV_t^{(m)} + \varepsilon_{t+1}$$

- Test: $H_0: \beta^+ = \beta^-$ (symmetry) vs $H_a: \beta^- > \beta^+$ (leverage)
- Patton & Sheppard (2015): "Good volatility, bad volatility" — downside variance predicts future variance more strongly than upside variance

```{r}
#| eval: false
# Semivariances from LOBSTER or VOLARE
har_rs <- HARmodel(
  data    = cbind(rv = rv_aapl, rs_plus = rs_plus, rs_minus = rs_minus),
  periods = c(1, 5, 22),
  type    = "HARSEMV",
  h       = 1
)
```

# VOLARE and the Full Cross-Section

## Using VOLARE for All 50 Stocks

**VOLARE** removes the need to process raw LOBSTER data for every stock-day combination:

- Pre-computed daily $RV_t$, $BPV_t$, realized kernels, $RS_t^+$, $RS_t^-$, quarticity
- Updated regularly, covers our 50 NASDAQ stocks
- Enables HAR estimation across the **full 50-stock cross-section**

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# Load VOLARE data (CSV or API)
volare_data <- read_csv("volare_nasdaq50.csv") |>
  select(date, symbol, rv, bpv, rk, rs_plus, rs_minus)

# Estimate HAR-RV for all 50 stocks in one go
library(future.apply)
plan(multisession, workers = 8)

har_fits <- future_lapply(unique(volare_data$symbol), function(sym) {
  df <- volare_data |>
    filter(symbol == sym) |>
    arrange(date) |>
    mutate(
      rv_d    = rv,
      rv_w    = slider::slide_dbl(rv, mean, .before = 4, .complete = TRUE),
      rv_m    = slider::slide_dbl(rv, mean, .before = 21, .complete = TRUE),
      rv_lead = lead(rv)
    ) |>
    drop_na()

  lm(rv_lead ~ rv_d + rv_w + rv_m, data = df)
})
names(har_fits) <- unique(volare_data$symbol)
```

## Python
```{python}
#| eval: false
from joblib import Parallel, delayed

def fit_har(symbol, data):
    df = (data[data["symbol"] == symbol]
          .sort_values("date")
          .assign(
              rv_d = lambda x: x["rv"],
              rv_w = lambda x: x["rv"].rolling(5).mean(),
              rv_m = lambda x: x["rv"].rolling(22).mean(),
              rv_lead = lambda x: x["rv"].shift(-1)
          )
          .dropna())
    return smf.ols("rv_lead ~ rv_d + rv_w + rv_m", data=df).fit()

har_fits = Parallel(n_jobs=8)(
    delayed(fit_har)(sym, volare_data)
    for sym in volare_data["symbol"].unique()
)
```
:::

## Cross-Sectional Variation in HAR Coefficients

Do HAR coefficients vary systematically across stocks? Patterns to look for:

- **By sector:** tech stocks may show stronger daily persistence ($\hat\beta_d$); large caps may have stronger weekly component ($\hat\beta_w$)
- **By liquidity:** less liquid stocks may show weaker HAR fit (more noise in $RV_t$)
- **By size:** $R^2$ tends to be higher for larger stocks with more liquid intraday data

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# Extract HAR coefficients across all 50 stocks
har_coefs <- bind_rows(lapply(names(har_fits), function(sym) {
  tibble(
    symbol = sym,
    beta_d = coef(har_fits[[sym]])["rv_d"],
    beta_w = coef(har_fits[[sym]])["rv_w"],
    beta_m = coef(har_fits[[sym]])["rv_m"],
    r2     = summary(har_fits[[sym]])$r.squared
  )
}))

# Join with sector metadata
har_coefs |>
  left_join(sector_lookup, by = "symbol") |>
  group_by(sector) |>
  summarise(across(starts_with("beta"), mean), r2 = mean(r2))
```

## Python
```{python}
#| eval: false
coef_records = []
for sym, fit in zip(volare_data["symbol"].unique(), har_fits):
    coef_records.append({
        "symbol": sym,
        "beta_d": fit.params["rv_d"],
        "beta_w": fit.params["rv_w"],
        "beta_m": fit.params["rv_m"],
        "r2": fit.rsquared
    })
har_coefs = pd.DataFrame(coef_records)
```
:::

# GARCH vs HAR: A Forecasting Comparison

## The Information Advantage of HAR

| Feature | GARCH(1,1) | HAR-RV |
|---------|-----------|--------|
| Input data | Daily returns $r_t$ | Intraday realized variance $RV_t$ |
| Information set | $r_t^2$ (very noisy) | $RV_t$ (precise) |
| Memory structure | Exponential decay | Slow hyperb. via weekly/monthly avg |
| Leverage | GJR-GARCH only | HAR-RS (Patton & Sheppard) |
| Estimation | MLE/QML | OLS |
| Scalability | Medium (parallel) | Very fast (OLS) |
| Data requirement | Daily prices | Intraday tick data or VOLARE |

**Empirical finding:** HAR typically dominates at **daily and weekly horizons**; GARCH is competitive only when intraday data is unavailable

**Why HAR wins:**

- $RV_t$ is essentially a direct observation of $\sigma_t^2$ — much less noise than $r_t^2$
- The regression SNR is dramatically higher for HAR than for GARCH
- Variance of $RV_t$ is $O(1/n)$ where $n$ = number of intraday returns; variance of $r_t^2$ is $O(1)$

## Rolling Out-of-Sample Evaluation

::: {.panel-tabset group="language"}
## R
```{r}
#| eval: false
# Rolling forecast evaluation: HAR vs GARCH vs EWMA
# Use 5-year estimation window, 2-year out-of-sample period

oos_results <- lapply(unique(volare_data$symbol), function(sym) {
  rv_sym  <- volare_data |> filter(symbol == sym) |> pull(rv)
  ret_sym <- daily_returns[, sym]

  # Rolling HAR forecasts
  har_fc <- roll_har_forecast(rv_sym, train = 1260, h = 1)

  # Rolling GARCH forecasts (from Lecture 2 fits)
  garch_fc <- roll_garch_forecast(ret_sym, train = 1260, h = 1)

  # EWMA forecasts
  ewma_fc <- ewma_vol(ret_sym, lambda = 0.94)^2

  # Evaluate via QLIKE
  tibble(
    symbol = sym,
    qlike_har   = mean(qlike_loss(rv_sym[-(1:1260)], har_fc)),
    qlike_garch = mean(qlike_loss(rv_sym[-(1:1260)], garch_fc)),
    qlike_ewma  = mean(qlike_loss(rv_sym[-(1:1260)], ewma_fc[-(1:1260)]))
  )
}) |> bind_rows()

oos_results |>
  summarise(across(starts_with("qlike"), mean))
```

## Python
```{python}
#| eval: false
def qlike_loss(actual, forecast):
    return actual / forecast - np.log(actual / forecast) - 1

results = []
for sym in volare_data["symbol"].unique():
    rv = volare_data[volare_data["symbol"] == sym]["rv"].values
    # ... compute rolling forecasts for HAR, GARCH, EWMA ...
    results.append({
        "symbol": sym,
        "qlike_har": np.mean(qlike_loss(rv_oos, har_fc)),
        "qlike_garch": np.mean(qlike_loss(rv_oos, garch_fc))
    })
```
:::

## Realized GARCH: Bridging Both Worlds

Hansen, Huang & Shek (2012): the **Realized GARCH** model uses realized measures *within* the GARCH framework

$$r_t = \sqrt{h_t}\, z_t, \quad z_t \overset{i.i.d.}{\sim} \mathcal{D}(0,1)$$

$$\log h_{t+1} = \omega + \beta \log h_t + \gamma \log x_t$$

$$\log x_t = \xi + \varphi \log h_t + \tau(z_t) + u_t$$

where $x_t$ is a realized measure (e.g., $RV_t$ or realized kernel) and $\tau(z_t) = \tau_1 z_t + \tau_2(z_t^2 - 1)$ captures the leverage effect

**Best of both worlds:**

- Uses daily returns for consistency with the conditional return distribution
- Incorporates realized measures for efficiency gains
- Produces proper conditional variance forecasts with uncertainty quantification

## From Univariate to the Full Covariance Matrix

We now have precise univariate forecasts $\hat\sigma_{i,t+1|t}^2$ for all 50 stocks — but the portfolio optimizer needs the **full covariance matrix**

**Preview of remaining lectures:**

| Lecture | What it adds |
|---------|-------------|
| 5 (DCC) | Time-varying conditional correlations $\hat C_{t+1|t}$ from daily returns |
| 6 (Realized Cov) | Full $\hat\Sigma_{t+1|t}$ estimated from intraday cross-products |

**Two complementary approaches:**

1. **HAR diagonal + DCC correlations:** use HAR forecasts for $D_{t+1|t}$, DCC for $C_{t+1|t}$
2. **Realized GARCH + scalar DCC:** combine realized GARCH for margins with DCC for correlations

# Exercises

## Exercise Set 4

**HAR estimation**

1. Pull realized measures from VOLARE for all 50 NASDAQ stocks. For each stock, estimate:
   - HAR-RV (baseline)
   - HAR-J (add jump component)
   - HAR-CJ (separate continuous and jump components)
   - HAR-RS (leverage: separate $RS_t^+$ and $RS_t^-$)

   Report the estimated coefficients and $R^2$ for each specification.

2. Is the jump component $\hat\beta_J$ statistically significant? Does the HAR-CJ outperform HAR-RV in-sample (adjusted $R^2$)?

3. Test for the leverage effect: is $\hat\beta^- > \hat\beta^+$ in HAR-RS? What fraction of the 50 stocks show a significant leverage effect?

**Cross-sectional analysis**

4. Summarise HAR-RV coefficients $(\hat\beta_d, \hat\beta_w, \hat\beta_m)$ by sector (tech, healthcare, consumer discretionary). Are there systematic differences? What do the patterns tell you about the heterogeneity of market participants across sectors?

**Out-of-sample forecasting horse race**

5. Using a rolling 5-year estimation window and 2-year out-of-sample period, compare:
   - HAR-RV
   - HAR-CJ
   - GARCH(1,1) (from Lecture 2)
   - GJR-GARCH(1,1) (from Lecture 2)
   - EWMA ($\lambda = 0.94$)

   Evaluation metric: QLIKE loss with $RV_t$ as the proxy. Which model wins most often across the 50 stocks?

**Portfolio evaluation**

6. Combine HAR-RV forecasts for $D_{t+1|t}$ with the constant sample correlation matrix $\hat C$. Construct the GMVP and compute out-of-sample realised portfolio variance. Compare to:
   - GARCH-based diagonal $\hat D_{t+1|t}$ + sample correlation (Lecture 2)
   - Rolling SD (Lecture 1)

   Does better volatility forecasting translate into better portfolio performance?

**Key references:** Corsi (2009); Andersen, Bollerslev & Diebold (2007); Patton & Sheppard (2015)
